# <span class="en">Data and study design</span><span class="zh">数据与研究设计</span> {#sec-data}

## <span class="en">Data sources</span><span class="zh">数据来源</span>

```{r}
#| label: tbl-sources
#| tbl-cap: "<span class='en'>Datasets used, with provenance and the role each plays in the analysis.</span><span class='zh'>所用数据集、来源出处,以及各自在分析中承担的角色。</span>"

tibble::tribble(
  ~Dataset, ~Source, ~Coverage, ~Accessed, ~Role,
  "VIIRS S-NPP 375 m active fire (VNP14IMGTDL, NRT)",
  "NASA FIRMS / LANCE",
  "South-East Asia regional extract; 4–11 Sep 2026",
  "12 Sep 2026",
  "**Primary point pattern**",

  "VIIRS NOAA-20 and NOAA-21 375 m active fire (NRT)",
  "NASA FIRMS / LANCE",
  "same window",
  "12 Sep 2026",
  "Independent replication of the observation process",

  "MODIS Terra/Aqua 1 km active fire (C6.1, NRT)",
  "NASA FIRMS / LANCE",
  "same window",
  "12 Sep 2026",
  "Coarse-resolution cross-check",

  "Indonesia subnational boundaries, admin levels 1–3 (COD-AB v01)",
  "Badan Pusat Statistik / UN OCHA, via HDX",
  "National; boundaries valid from 1 Apr 2020, file published 23 Jun 2026",
  "12 Sep 2026",
  "**Study window** (ADM2) and reporting units (ADM3)",

  "geoBoundaries gbOpen IDN ADM2",
  "William & Mary geoLab",
  "National",
  "12 Sep 2026",
  "Independent cross-check of the study-window geometry",

  "OpenStreetMap: waterway=canal/drain/ditch, waterway=river/stream, highway=*, place=*",
  "OpenStreetMap contributors (Overpass API), ODbL 1.0",
  "Study bounding box",
  "12 Sep 2026",
  "Spatial covariates and cartographic reference"
) |>
  gt::gt() |>
  gt::fmt_markdown(columns = c(Dataset, Role)) |>
  gt::cols_width(Dataset ~ gt::pct(26), Source ~ gt::pct(18),
                 Coverage ~ gt::pct(22), Accessed ~ gt::pct(10),
                 Role ~ gt::pct(24)) |>
  gt::tab_options(table.font.size = gt::pct(82),
                  data_row.padding = gt::px(5))
```

::: {.en}
Two acquisition routes are implemented, and both are in the repository.
`scripts/01_download_firms.R` uses the authenticated FIRMS *Area API*, which
supports arbitrary date ranges and returns the science-quality Standard
Processing products; `scripts/01b_assemble_public7d.R` uses the open 7-day
regional CSVs, which need no credentials. Everything downstream is identical.
**The results in this report were produced from the open 7-day route**, so the
temporal window is the eight local days from `r format(DATE_MIN, "%d %B")` to
`r format(DATE_MAX, "%d %B %Y")`, and the products are Near Real-Time. The
consequences of that choice are stated wherever they matter, most
importantly in @sec-typeflag, because the NRT product omits the `type` flag
that identifies static non-vegetation heat sources.
:::

::: {.zh}
仓库中实现了两条数据获取路径。`scripts/01_download_firms.R` 使用需认证的 FIRMS
*Area API*,支持任意日期范围并返回科学质量的标准处理(SP)产品;
`scripts/01b_assemble_public7d.R` 使用无需凭据的开放 7 日区域 CSV。两条路径的
下游处理完全一致。**本报告结果来自开放 7 日路径**,因此时间窗口为八个当地日
(`r format(DATE_MIN, "%Y-%m-%d")` 至 `r format(DATE_MAX, "%Y-%m-%d")`),
产品为近实时(NRT)。这一选择的后果会在相关处逐一说明,最重要的是
@sec-typeflag,因为 NRT 产品缺少用于识别静态非植被热源的 `type` 标记。
:::

## <span class="en">Study area</span><span class="zh">研究区</span>

```{r}
#| label: fig-studyarea
#| fig-cap: "<span class='en'>Study area. Left: Pulang Pisau Regency within Kalimantan, with all VIIRS S-NPP detections recorded across the island in the same eight days. Right: the regency, its eight kecamatan, and the drainage geography that motivates the analysis.</span><span class='zh'>研究区。左:Pulang Pisau 县在加里曼丹岛中的位置,叠加同一八天内全岛的 VIIRS S-NPP 探测点。右:该县、其八个区(kecamatan),以及引出本研究问题的排水地理格局。</span>"
#| fig-alt: "Study area. Left: Pulang Pisau Regency within Kalimantan, with all VIIRS S-NPP detections recorded across the island in the same eight days. Right: the regency, its eight kecamatan, and the drainage geography that motivates the analysis."
#| fig-width: 11
#| fig-height: 8.2

kal_prov <- readRDS(file.path(DER, "kal_prov.rds"))
kal_fire <- readRDS(file.path(DER, "kal_fires_context.rds"))

# --- Left panel: island context ---------------------------------------------
# Showing island-wide detections makes the case for the site selection visible
# rather than asserted: Pulang Pisau is one of the densest parts of a very
# active island in this particular week.
m_ctx <- ggplot() +
  geom_sf(data = kal_prov, fill = "#f5f5f4", colour = "white", linewidth = 0.3) +
  geom_sf(data = kal_fire, colour = PAL$fire_l, size = 0.10, alpha = 0.35) +
  geom_sf(data = bnd, fill = NA, colour = PAL$fire, linewidth = 0.8) +
  annotate("text", x = sf::st_coordinates(sf::st_centroid(bnd))[1],
           y = sf::st_bbox(bnd)["ymin"] - 55000, label = "Pulang Pisau",
           size = 3.1, fontface = "bold", colour = "#1c1917") +
  ggspatial::annotation_scale(location = "br", height = unit(0.12, "cm"),
                              text_cex = 0.55, line_width = 0.4) +
  labs(title = "Kalimantan, 4–11 September 2026",
       subtitle = paste0(n_fmt(nrow(kal_fire)),
                         " VIIRS S-NPP detections (low-confidence excluded)")) +
  theme_map(11) + theme(legend.position = "none")

# --- Right panel: the regency itself ----------------------------------------
m_area <- ggplot() +
  geom_sf(data = bnd, fill = "#fafaf9", colour = NA) +
  geom_sf(data = osm$canal, colour = PAL$water, linewidth = 0.12, alpha = 0.5) +
  geom_sf(data = osm$river, colour = "#155e75", linewidth = 0.42) +
  geom_sf(data = kec, fill = NA, colour = "#a8a29e", linewidth = 0.3,
          linetype = "22") +
  geom_sf(data = bnd, fill = NA, colour = "#1c1917", linewidth = 0.6) +
  geom_sf(data = osm$settle, shape = 21, fill = "white", colour = "#1c1917",
          size = 0.75, stroke = 0.3) +
  ggrepel::geom_text_repel(
    data = sf::st_point_on_surface(kec),
    aes(geometry = geometry, label = adm3_name), stat = "sf_coordinates",
    size = 2.5, fontface = "bold", colour = "#1c1917",
    bg.color = "white", bg.r = 0.14, min.segment.length = 0.4,
    segment.size = 0.25, seed = 7) +
  ggspatial::annotation_scale(location = "br", height = unit(0.12, "cm"),
                              text_cex = 0.55, line_width = 0.4) +
  ggspatial::annotation_north_arrow(
    location = "tr", height = unit(0.7, "cm"), width = unit(0.7, "cm"),
    style = ggspatial::north_arrow_minimal(text_size = 6)) +
  labs(title = "Pulang Pisau Regency",
       subtitle = paste0("Thin blue: ", n_fmt(nrow(osm$canal)),
                         " canal/drain/ditch segments · thick blue: natural rivers\n",
                         "Dashed: kecamatan boundaries · circles: settlements")) +
  theme_map(11) + theme(legend.position = "none")

m_ctx + m_area + patchwork::plot_layout(widths = c(0.47, 0.53))
```

::: {.interp .en}
**Across the island.** Kalimantan burned widely in this week (`r n_fmt(nrow(kal_fire))` S-NPP detections island-wide), but activity is
patchy, and Pulang Pisau sits inside one of the densest southern clusters.
Within the regency, the canal network (thin blue) is dense in the centre and
south and sparse in the forested north, while natural rivers (thick blue) run
broadly north–south. Settlements are few and align with the rivers and the
canal grid. The visual expectation this sets up (that fire should follow the
canal grid rather than the rivers) is tested formally in @sec-fo-covariates
and should not be read off this map, which shows infrastructure, not fire.
*One limit:* OSM completeness is uneven and is likely better near settlements,
which could by itself produce an apparent canal–fire association.
:::

::: {.interp .zh}
**放到全岛看。** 本周加里曼丹全岛普遍过火(全岛
`r n_fmt(nrow(kal_fire))` 个 S-NPP 探测点)但活动呈斑块状,而 Pulang Pisau
正位于南部最密集的簇群之一。县内运河网络(细蓝线)在中部与南部密集、在北部林区
稀疏;天然河流(粗蓝线)大致南北走向。居民点数量不多,沿河流与运河格网分布。
本图建立的视觉预期(火应跟随运河格网而非河流)将在 @sec-fo-covariates 中
正式检验,**不应**直接从本图读出结论:本图显示的是基础设施,不是火。
*一点限度:* OSM 完备性并不均匀,在居民点附近很可能更好,仅此一点就足以造成
「运河, 火」关联的假象。
:::

## <span class="en">What counts as a point event</span><span class="zh">什么算一个点事件</span> {#sec-event-def}

::: {.en}
This is the single most consequential definitional decision in the study, and
it has no unique right answer.

A FIRMS record is one **pixel-overpass**: the algorithm flagged a 375 m pixel as
containing active fire during one satellite pass [@schroeder2014]. Treating each
record as a point event therefore builds three kinds of duplication into the
pattern:

1. **Spatial duplication**, a single burning area spans several adjacent
   pixels, all of which are flagged.
2. **Temporal duplication within a day**, S-NPP passes roughly twice daily, so
   a fire burning for a full day can be flagged twice at the same location.
3. **Multi-day persistence**, peat fires smoulder for days or weeks, and the
   same pixel recurs.

Only the third of these is a property of fire. The first two are properties of
the sensor. Because they operate at scales (375 m; 0 m) *below* the scales at
which any substantive question is posed, they would inflate exactly the
small-distance clustering that a second-order analysis is designed to detect.

Three analysis units are therefore constructed, and every second-order result
is reported for all three.
:::

::: {.zh}
这是全研究中最具决定性影响的定义性抉择,而且它没有唯一正确的答案。

一条 FIRMS 记录是一次**像元-过境**:算法在某次卫星过境时把一个 375 m 像元标记为
含有活跃火 [@schroeder2014]。因此,若把每条记录都当作一个点事件,就会把三种重复
性内建进模式:

1. **空间重复**,一片燃烧区域横跨若干相邻像元,这些像元全部被标记。
2. **日内时间重复**,S-NPP 每天约过境两次,因此烧了一整天的火可能在同一位置被
   标记两次。
3. **跨日持续**,泥炭火会阴燃数天甚至数周,同一像元反复出现。

其中只有第三种是**火**的性质,前两种是**传感器**的性质。由于它们作用的尺度
(375 m;0 m)**低于**任何实质性问题所设定的尺度,它们恰恰会放大二阶分析本就要
检测的小距离聚集。

因此构建三种分析单元,且每一项二阶结果都对三者分别报告。
:::

```{r}
#| label: tbl-units
#| tbl-cap: "<span class='en'>The three analysis units. The pixel-day unit is used for all headline results.</span><span class='zh'>三种分析单元。所有主要结果均采用「像元-日」单元。</span>"

tibble::tibble(
  Unit = c("**Detection**", "**Pixel-day** *(primary)*", "**Event**"),
  Definition = c(
    "Every retained record: one 375 m pixel in one overpass.",
    "One point per unique 375 m grid cell per **local** calendar day; among duplicates the highest-FRP record is kept.",
    "Single-linkage cluster of pixel-days within 750 m *and* 1 day of one another, reduced to its centroid."),
  n = c(N_DET, N_PX, N_EV),
  Removes = c(
    "nothing",
    "same-day repeat overpasses of one pixel",
    "spatial extent and multi-day persistence of one burning area"),
  Interprets_as = c(
    "detection intensity, closest to the raw observation",
    "daily fire-affected pixel, the compromise used throughout",
    "approximate distinct fire event, i.e. nearest to *ignitions*")
) |>
  gt::gt() |>
  gt::cols_label(Interprets_as = "Interpreted as") |>
  gt::fmt_markdown(columns = c(Unit)) |>
  gt::fmt_number(columns = n, decimals = 0) |>
  gt::tab_options(table.font.size = gt::pct(84),
                  data_row.padding = gt::px(6))
```

::: {.en}
The pixel-day unit is primary because it removes the duplication that is purely
instrumental (repeat overpasses) without discarding the duplication that is
substantively meaningful (a pixel that burns on several successive days is
genuinely burning on several successive days). The event unit is the most
appealing conceptually, but @sec-so-robustness shows that at this fire density
single-linkage chaining merges much of the landscape into one component, so it
is used for robustness rather than for headline results.
:::

::: {.zh}
之所以以「像元-日」为主单元,是因为它剔除了纯属仪器性的重复(重复过境),同时
保留了实质上有意义的重复(某像元连续几天燃烧,就是真的连续几天在烧)。「事件」
单元在概念上最有吸引力,但 @sec-so-robustness 表明:在本例的火密度下,单连接
链式合并会把大片景观并成一个连通分量,因此该单元仅用于稳健性检验,而非主要结果。
:::

## <span class="en">Spatial study window</span><span class="zh">空间研究窗口</span> {#sec-window}

::: {.en}
The window is **Pulang Pisau Regency** as delineated in the official COD-AB
ADM2 layer [@codab2026], projected to EPSG:32749.
:::

::: {.zh}
研究窗口为官方 COD-AB ADM2 图层 [@codab2026] 界定的 **Pulang Pisau 县**,
投影至 EPSG:32749。
:::

```{r}
#| label: window-numbers
#| echo: false
gb <- readRDS(file.path(DER, "kal_adm2.rds")) |>
  dplyr::filter(shapeName == "Pulang Pisau") |> sf::st_transform(P$crs_proj)
area_gb <- as.numeric(sf::st_area(gb)) / 1e6
area_attr <- readRDS(file.path(DER, "pp_adm2_cod.rds"))$area_sqkm[1]
```

::: {.en}
Three choices need defending.

**Why an administrative boundary at all?** A point process must be analysed
inside a window that is defined independently of the points; otherwise edge
effects and intensity estimates are conditioned on the data. Candidate windows
were a regency boundary, a peat-hydrological unit, and a convex or concave hull
of the detections. The hull is disqualified outright. It is defined by the
data. A hydrological unit would be scientifically preferable, since the process
is hydrological, but it does not correspond to any administrative mandate. The
regency is chosen because **it is the unit that owns the response**: fire
suppression, land-use permitting and village-level prevention budgets are
administered at regency and kecamatan level, so findings expressed on this
window are directly actionable. The cost is that the boundary is arbitrary with
respect to the fire process, and @sec-so-robustness tests the consequence.
:::

::: {.zh}
有三项选择需要论证。

**为什么要用行政边界?** 点过程必须在一个**独立于点本身**定义的窗口内分析,否则
边界效应与强度估计都会被数据所条件化。候选窗口有三种:县界、泥炭水文单元,以及
探测点的凸包/凹包。凸包直接被排除。它是由数据定义的。水文单元在科学上更可取
(该过程本身就是水文性的),但它不对应任何行政职责。最终选择县界,因为**它是
承担应对责任的单元**:灭火、土地利用许可与村级预防预算都在县与区(kecamatan)
层级管理,因此在这一窗口上表述的发现可以直接落地。代价是该边界相对于火灾过程是
任意的,@sec-so-robustness 检验了这一代价的后果。
:::

::: {.en}
**Why the official COD-AB layer?** It is sourced from Badan Pusat Statistik and
carries a stated validity date (boundaries valid from 1 April 2020), a version,
and kecamatan-level nesting. The independently produced geoBoundaries ADM2
polygon for the same regency gives an area of
`r sprintf("%.1f", area_gb)` km² against
`r sprintf("%.1f", AREA_KM2)` km² here (a difference of
`r sprintf("%.2f%%", 100*abs(area_gb-AREA_KM2)/AREA_KM2)`), which is
reassuring, and unsurprising, since both ultimately derive from the same BPS
geometry.

**What is the window's actual extent?** The polygon is a single part,
topologically valid, with `r n_fmt(nrow(sf::st_coordinates(bnd)))` vertices.
Its area is `r n_fmt(AREA_KM2)` km², closely matching the
`area_sqkm` attribute carried in the source file
(`r sprintf("%.0f", area_attr)` km²). It is strongly **elongated**:
`r sprintf("%.0f", diff(win$xrange)/1000)` km east–west by
`r sprintf("%.0f", diff(win$yrange)/1000)` km north–south. This matters twice
over: it caps the distance at which second-order functions can be estimated
(@sec-so-params), and it makes a single isotropic kernel bandwidth a stronger
assumption than it would be in a compact window (@sec-fo-bandwidth). The
southern boundary is a coast, so part of the perimeter borders water rather
than land, a further source of edge effect that no edge correction models,
because those corrections assume the process continues beyond the boundary.
:::

::: {.zh}
**为什么用官方 COD-AB 图层?** 它源自印尼中央统计局(Badan Pusat Statistik),
带有明确生效日期(边界自 2020 年 4 月 1 日起有效)、版本号,以及区级嵌套结构。
独立制作的 geoBoundaries ADM2 多边形对同一个县给出面积
`r sprintf("%.1f", area_gb)` km²,本研究为 `r sprintf("%.1f", AREA_KM2)` km²,
差异 `r sprintf("%.2f%%", 100*abs(area_gb-AREA_KM2)/AREA_KM2)`。这既令人放心
也不意外,因为两者最终都源自同一套 BPS 几何。

**这个窗口的实际范围是什么?** 该多边形为单一部件、拓扑有效,共
`r n_fmt(nrow(sf::st_coordinates(bnd)))` 个顶点。面积
`r n_fmt(AREA_KM2)` km²,与源文件所带 `area_sqkm` 属性
(`r sprintf("%.0f", area_attr)` km²)高度一致。其形状**高度狭长**:东西向
`r sprintf("%.0f", diff(win$xrange)/1000)` km,南北向
`r sprintf("%.0f", diff(win$yrange)/1000)` km。这有双重影响:它限制了二阶函数
可被估计的最大距离(@sec-so-params),也使「单一各向同性核带宽」这一假设比在
紧凑窗口中更强(@sec-fo-bandwidth)。南界为海岸线,因此部分周长临水而非临陆
。这是又一处边界效应来源,而任何边界校正都无法建模它,因为那些校正都假定过程在
边界之外继续存在。
:::

## <span class="en">Temporal study window</span><span class="zh">时间研究窗口</span> {#sec-temporal-window}

::: {.en}
The analysis covers **`r format(DATE_MIN, "%d %B")` to
`r format(DATE_MAX, "%d %B %Y")`** (`r N_DAYS` local days), which is the full
extent of the open FIRMS archive at the time of access.

This is a short window, and honesty about what it can support matters more than
apparent sophistication:

- It falls in the **peak of the Central Kalimantan dry season**, so it captures
  the hazard at its most intense, not as an annual average. Conclusions
  describe a burning episode, not a fire year.
- **No seasonality can be estimated.** Questions about monthly or inter-annual
  variation are outside the reach of these data, and are not asked.
- Temporal aggregation is to the **local calendar day**. Finer aggregation is
  impossible: as @sec-fo-temporal shows, S-NPP contributes detections at only
  about two local times per day, so anything finer than a day would describe
  the satellite's orbit rather than the fire's behaviour.
- Eight days is enough for a space-time interaction test at lags of 0–5 days
  (@sec-so-knox), which is the relevant scale for fire spread and for
  day-to-day ignition decisions, but not for anything longer.

The authenticated acquisition script is included precisely so that this
limitation is removable: supplying a FIRMS `MAP_KEY` and rerunning
`scripts/01_download_firms.R` with a wider date range reproduces every result
below over a full season, with no change to any other file.
:::

::: {.zh}
分析覆盖 `r format(DATE_MIN, "%Y-%m-%d")` 至
`r format(DATE_MAX, "%Y-%m-%d")`,共 **`r N_DAYS` 个当地日**,即访问时开放 FIRMS
档案的全部范围。

这是一个很短的窗口,而诚实说明它能支持什么,比表面上的复杂度更重要:

- 它正落在**中加里曼丹旱季峰值**,因此捕捉到的是灾害最强烈的状态,而非年度平均。
  结论描述的是一次燃烧事件,不是一个火年。
- **无法估计季节性。** 月际或年际变化的问题超出本数据能力范围,故本报告不提出。
- 时间聚合到**当地日历日**。更细的聚合不可能:如 @sec-fo-temporal 所示,S-NPP
  每天只在约两个当地时刻贡献探测,因此任何比「日」更细的粒度描述的都是卫星轨道,
  而不是火的行为。
- 八天足以支持滞后 0–5 天的时空交互检验(@sec-so-knox),这正是火蔓延与逐日点火
  决策的相关尺度;但不足以支持更长尺度的分析。

之所以把需认证的获取脚本一并纳入,正是为了让这一局限**可以被移除**:提供 FIRMS
`MAP_KEY` 并以更宽日期范围重跑 `scripts/01_download_firms.R`,即可在整个火季上
复现下文全部结果,其余文件一行都不需要改。
:::
