# <span class="en">Data preparation and quality assessment</span><span class="zh">数据准备与质量评估</span> {#sec-prep}

::: {.en}
All preparation is performed by `scripts/03_prepare.R`, which is reproduced in
full in @sec-repro-scripts. This section explains each decision and reports what
it cost in records. The guiding principle is that **no record is discarded
silently**: every filter writes a row into an audit table, so the path from
`r n_fmt(audit$n_in[1])` raw records to the analysed pattern is fully traceable.
:::

::: {.zh}
全部数据准备由 `scripts/03_prepare.R` 完成,该脚本在 @sec-repro-scripts 中完整
附录。以下解释每一项决策,并报告它在记录数上的代价。指导原则是**任何记录都不会被
悄悄丢弃**:每一道筛选都会向审计表写入一行,因此从 `r n_fmt(audit$n_in[1])` 条
原始记录到最终分析模式的整条路径都是可追溯的。
:::

```{r}
#| label: audit-helper
#| echo: false
aud <- audit |> mutate(pct_removed = 100 * n_removed / n_in)
get_aud <- function(pat) aud |> dplyr::filter(grepl(pat, step)) |> as.list()
```

## <span class="en">Import and schema harmonisation</span><span class="zh">数据导入与模式统一</span>

::: {.en}
Every column is read as `character` first. This is deliberate: if `readr`
guessed a numeric type it would coerce a malformed coordinate to `NA` during
parsing, and the record would be lost before it could be counted as a data
fault. Numeric coercion happens afterwards, under our own control.

The four products do not share a schema. VIIRS reports channel brightness
temperatures as `bright_ti4` (4 µm) and `bright_ti5` (11 µm); MODIS reports
`brightness` and `bright_t31`. Confidence is a three-level ordinal class for
VIIRS (`low`/`nominal`/`high`), but an integer percentage 0–100 for MODIS. These
are mapped onto common fields (`bt_fire`, `bt_bg`, `confidence_cls`), so that one
filtering rule can be stated, while `firms_source` preserves the provenance of
every row.
:::

::: {.zh}
所有列先一律按 `character` 读入。这是刻意的:如果让 `readr` 猜测数值类型,它会在
解析阶段就把格式错误的坐标强制转成 `NA`,该记录在能被计为一个数据缺陷之前就已经
丢失了。数值转换在此之后、由我们自己控制地进行。

四种产品的数据模式并不一致。VIIRS 用 `bright_ti4`(4 µm)和 `bright_ti5`(11 µm)
报告通道亮温;MODIS 用 `brightness` 和 `bright_t31`。置信度在 VIIRS 中是三级有序
分类(`low`/`nominal`/`high`),在 MODIS 中却是 0–100 的整数百分比。这些字段被
映射到统一字段(`bt_fire`、`bt_bg`、`confidence_cls`),以便用一条筛选规则表述;
同时 `firms_source` 保留每一行的来源出处。
:::

```{r}
#| label: schema-demo
#| code-fold: true
#| code-summary: "Confidence harmonisation rule"
#| eval: false

fires |> mutate(confidence_cls = case_when(
  sensor_family == "VIIRS" & str_starts(str_to_lower(confidence_raw), "l") ~ "low",
  sensor_family == "VIIRS" & str_starts(str_to_lower(confidence_raw), "n") ~ "nominal",
  sensor_family == "VIIRS" & str_starts(str_to_lower(confidence_raw), "h") ~ "high",
  sensor_family == "MODIS" & conf_num <  30 ~ "low",      # MODIS 0-100 scale
  sensor_family == "MODIS" & conf_num <  80 ~ "nominal",
  sensor_family == "MODIS" & conf_num >= 80 ~ "high"))
```

## <span class="en">Coordinates: validity, and a precision trap</span><span class="zh">坐标:有效性,以及一个精度陷阱</span>

::: {.en}
No record had a missing or non-numeric coordinate, and none fell outside valid
geographic ranges or on the (0, 0) null island, checks that cost nothing and
would have been serious if they had fired.

There is, however, a subtler issue. Latitudes are written with between one and
five decimal places:
:::

::: {.zh}
没有任何记录出现坐标缺失或非数值的情况,也没有任何记录落在有效地理范围之外或
(0, 0) 这个「空岛」上。这些检查成本为零,但一旦触发就会很严重。

不过存在一个更微妙的问题。纬度的小数位数在一位到五位之间不等:
:::

```{r}
#| label: tbl-precision
#| tbl-cap: "<span class='en'>Decimal places in the latitude field as delivered. The variation is an artefact of dropped trailing zeros, not of varying measurement precision.</span><span class='zh'>交付数据中纬度字段的小数位数。这种差异是末尾零被省略造成的假象,而不是测量精度本身在变化。</span>"
raw_chk <- readr::read_csv(
  list.files("data/raw", "^firms_pulangpisau_.*[.]csv$", full.names = TRUE)[1],
  show_col_types = FALSE, col_types = readr::cols(.default = readr::col_character()))
tibble(dp = nchar(stringr::str_replace(raw_chk$latitude, "^[^.]*[.]?", ""))) |>
  count(dp, name = "records") |>
  mutate(`implied precision` = scales::label_comma(accuracy = 0.01)(111320 / 10^dp),
         `implied precision` = paste0(`implied precision`, " m")) |>
  gt::gt() |> gt::cols_label(dp = "Decimal places") |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(4))
```

::: {.en}
Five decimal places implies roughly metre precision, which is spurious. The
reported coordinate is the **centre of a 375 m nominal pixel**, and the pixel
grows towards the swath edge, the `scan` and `track` fields in these data reach
`r sprintf("%.2f", max(fire_sf$scan))` and
`r sprintf("%.2f", max(fire_sf$track))` km respectively. The effective location
uncertainty of a point in this pattern is therefore of order a few hundred
metres, not metres.
:::

::: {.zh}
五位小数意味着约米级精度,而这是虚假的。所报告的坐标是**一个 375 m 标称像元的
中心**,而像元尺寸向扫描边缘增大,本数据中 `scan` 与 `track` 字段分别达到
`r sprintf("%.2f", max(fire_sf$scan))` 与 `r sprintf("%.2f", max(fire_sf$track))`
km。因此本模式中一个点的有效定位不确定度是**数百米量级**,而不是米级。
:::

::: {.callout-important .en}
## The consequence for second-order analysis
Because location uncertainty is of order 375 m, **no second-order result below
about 400–500 m can be interpreted as a property of fire**. Structure at those
distances is a property of the sensor's pixel grid. This bound is used in
@sec-fo-bandwidth to reject two of the four automatic bandwidth selectors, and
it is the reason all pair-correlation interpretation in @sec-so-pcf starts above
500 m.
:::

::: {.callout-important .zh}
## 这对二阶分析的后果

由于定位不确定度为 375 m 量级,**任何低于约 400–500 m 的二阶结果都不能被解释为火的
性质**。这些距离上的结构是传感器像元格网的性质。这一下界被用于 @sec-fo-bandwidth
中否决四个自动带宽选择器中的两个,也是 @sec-so-pcf 中所有对相关函数的解读都从
500 m 以上开始的原因。
:::

## <span class="en">Timestamps: the local-time correction</span><span class="zh">时间戳:本地时校正</span> {#sec-timezone}

::: {.en}
FIRMS reports `acq_date` and `acq_time` in **UTC**. Central Kalimantan observes
*Waktu Indonesia Barat* (WIB), UTC+07. The S-NPP night overpass occurs at
roughly 01:30 local time, which is **18:30 UTC on the previous calendar day**.

Deriving a fire's "day" from the UTC date would therefore assign every night
detection to the wrong day. The effect is not marginal:
:::

::: {.zh}
FIRMS 以 **UTC** 报告 `acq_date` 与 `acq_time`。中加里曼丹使用西印尼时间
(*Waktu Indonesia Barat*, WIB),即 UTC+07。S-NPP 的夜间过境约在当地时间 01:30,
也就是**前一个日历日的 UTC 18:30**。

因此,若用 UTC 日期推导一场火的「日期」,就会把每一个夜间探测都归到错误的日子。
这个影响并不轻微:
:::

```{r}
#| label: tbl-shift
#| tbl-cap: "<span class='en'>Records whose local calendar date differs from the UTC calendar date, by overpass.</span><span class='zh'>本地日历日期与 UTC 日历日期不一致的记录数,按过境类型划分。</span>"
fire_sf |> sf::st_drop_geometry() |>
  count(Overpass = overpass, `Local date differs from UTC date` = day_shifted) |>
  tidyr::pivot_wider(names_from = `Local date differs from UTC date`,
                     values_from = n, values_fill = 0L,
                     names_prefix = "shifted_") |>
  gt::gt() |>
  gt::cols_label(shifted_FALSE = "Same date", shifted_TRUE = "Shifted by one day") |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(4))
```

::: {.en}
`r n_fmt(sum(fire_sf$day_shifted))` of `r n_fmt(N_DET)` detections
(`r sprintf("%.0f%%", 100*mean(fire_sf$day_shifted))`) (precisely the night
overpasses) change date under the correction. Every temporal variable in this
report is built from the local timestamp.

A second fact emerges from the same conversion, and it constrains the whole
temporal analysis:
:::

::: {.zh}
`r n_fmt(N_DET)` 条探测中有 `r n_fmt(sum(fire_sf$day_shifted))` 条
(`r sprintf("%.0f%%", 100*mean(fire_sf$day_shifted))`)(恰好就是夜间过境)在
校正后改变了日期。此后每一个时间变量都基于本地时间戳构建。

同一次时区转换还揭示出第二个事实,它约束了整个时间分析:
:::

```{r}
#| label: fig-overpass
#| fig-cap: "<span class='en'>Local acquisition time of every S-NPP detection. Detections arrive in two narrow windows because that is when the satellite passes, so nothing finer than a daily aggregation can describe fire rather than orbit.</span><span class='zh'>每个 S-NPP 探测点的本地获取时刻。探测集中在两个狭窄的时间窗内,因为那就是卫星过境的时刻;因此任何比「日」更细的聚合描述的都是轨道而不是火。</span>"
#| fig-alt: "Local acquisition time of every S-NPP detection. Detections arrive in two narrow windows because that is when the satellite passes, so nothing finer than a daily aggregation can describe fire rather than orbit."
#| fig-width: 8
#| fig-height: 2.9

fire_sf |> sf::st_drop_geometry() |>
  ggplot(aes(hour_local, fill = overpass)) +
  geom_histogram(binwidth = 1/6, colour = NA) +
  scale_fill_manual(NULL, values = c(Day = PAL$fire_l, Night = PAL$water)) +
  scale_x_continuous(breaks = seq(0, 24, 3), limits = c(0, 24),
                     labels = function(x) sprintf("%02d:00", x)) +
  labs(title = "S-NPP samples this landscape twice a day, not continuously",
       subtitle = paste0("All ", n_fmt(N_DET),
                         " retained detections, local time (WIB, UTC+07)"),
       x = "Local time of acquisition", y = "Detections",
       caption = "Day overpass 12:08-13:54; night overpass 00:11-01:51 local.") +
  theme_th()
```

::: {.interp .en}
**Two overpasses, not a day.** The distribution is bimodal with empty gaps covering
about 21 of 24 hours. This is the satellite's orbit, not the fire's diurnal
cycle: S-NPP crosses the equator at about 13:30 and 01:30 local. **Two
consequences follow.** First, any diurnal question (when in the day do fires
start, peak, or die down) is unanswerable with these data, and is not asked.
Second, a fire that ignites and burns out between overpasses is invisible, so
the pattern systematically under-represents short-lived fires. This is a
first-order bias in the observation process, not random noise. *A caveat:* the
within-window spread (about 1.8 h) reflects orbital drift across the eight
days, not variation in when fires burn.
:::

::: {.interp .zh}
**两次过境,不是一整天。** 分布呈双峰,空白区间覆盖了 24 小时中的约 21 小时。这是卫星的
轨道,而不是火的日变化周期:S-NPP 约在当地 13:30 与 01:30 过赤道。**由此有两个
后果。** 第一,任何日变化问题(火在一天中何时开始、何时最盛、何时熄灭)用这些
数据都无法回答,因此不予讨论。第二,在两次过境之间点燃并烧完的火是不可见的,
因此本模式系统性地低估了短时火。这是观测过程中的**一阶偏差**,不是随机噪声。
*一点保留:* 窗口内约 1.8 小时的时刻散布反映的是八天间的轨道漂移,而不是火燃烧时刻的
变化。
:::

## <span class="en">Confidence, and what "high confidence" actually means</span><span class="zh">置信度,以及「高置信度」的真实含义</span> {#sec-confidence}

::: {.en}
Low-confidence VIIRS detections carry an elevated false-alarm rate
[@schroeder2014] and are removed:
`r n_fmt(get_aud("low confidence")$n_removed)` records
(`r sprintf("%.1f%%", get_aud("low confidence")$pct_removed)`) across all four
products.

Inspecting the retained classes produced an unexpected and important result.
:::

::: {.zh}
低置信度的 VIIRS 探测具有偏高的误报率 [@schroeder2014],已被剔除:四种产品合计
`r n_fmt(get_aud("low confidence")$n_removed)` 条
(`r sprintf("%.1f%%", get_aud("low confidence")$pct_removed)`)。

而检视被保留的各等级时,得到了一个出乎意料且重要的结果。
:::

```{r}
#| label: tbl-saturation
#| tbl-cap: "<span class='en'>Confidence class against I-04 channel saturation for S-NPP detections inside the study window. The correspondence is exact.</span><span class='zh'>研究窗口内 S-NPP 探测的置信度等级与 I-04 通道饱和情况的交叉表。两者完全一一对应。</span>"
fire_sf |> sf::st_drop_geometry() |>
  mutate(`I-04 saturated (>= 367 K)` = bt_fire >= 367) |>
  count(`Confidence class` = confidence_cls, `I-04 saturated (>= 367 K)`) |>
  tidyr::pivot_wider(names_from = `I-04 saturated (>= 367 K)`, values_from = n,
                     values_fill = 0L, names_prefix = "sat_") |>
  gt::gt() |>
  gt::cols_label(sat_FALSE = "Below saturation", sat_TRUE = "At 367 K ceiling") |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(4))
```

::: {.en}
Every one of the `r n_fmt(sum(fire_sf$confidence_cls == "high", na.rm = TRUE))`
"high" confidence detections sits exactly at
`r sprintf("%.1f", max(fire_sf$bt_fire))` K, the I-04 channel saturation limit,
and no "nominal" detection does. So in this product **"high confidence" is a
statement about fire intensity, not about locational reliability**. It flags
pixels so hot that the 4 µm channel saturated. Two things follow. First,
confidence must not be used as a weight for location accuracy. Second, the fire
radiative power reported for those `r sum(fire_sf$confidence_cls == "high", na.rm = TRUE)`
records is derived from a saturated radiance and is a **lower bound**; FRP is
consequently used only descriptively in this report, never as an analytical
weight.
:::

::: {.zh}
全部 `r n_fmt(sum(fire_sf$confidence_cls == "high", na.rm = TRUE))` 条「高」置信度
探测的亮温都恰好落在 `r sprintf("%.1f", max(fire_sf$bt_fire))` K,即 I-04 通道的
饱和上限,而没有任何一条「nominal」探测落在该值。因此在本产品中,**「高置信度」
陈述的是火的强度,而不是定位可靠性**。它标记的是热到让 4 µm 通道饱和的像元。
由此有两点推论。第一,置信度**不可**用作定位精度的权重。第二,这
`r sum(fire_sf$confidence_cls == "high", na.rm = TRUE)` 条记录所报告的火辐射功率
是由已饱和的辐亮度推导的,只是一个**下界**;因此 FRP 在全篇仅作描述性使用,
从不作为分析权重。
:::

## <span class="en">Static heat sources and the missing `type` flag</span><span class="zh">静态热源与缺失的 `type` 标记</span> {#sec-typeflag}

::: {.en}
The FIRMS Standard Processing products carry a `type` flag distinguishing
presumed vegetation fire (0) from active volcano (1), other static land source
(2) and offshore (3). **The open 7-day NRT files omit it**, so the filter cannot
be applied. Rather than ignore the gap, an empirical substitute is used: a
persistent industrial source such as a gas flare should be detected on
essentially every cloud-free overpass, so it would appear as a pixel recurring
on nearly all days.
:::

::: {.zh}
FIRMS 标准处理产品带有一个 `type` 标记,用于区分推定植被火(0)、活火山(1)、
其他静态陆地热源(2)与离岸(3)。**开放的 7 日 NRT 文件不含该标记**,因此这道
筛选无法应用。与其忽略这一缺口,这里采用一个经验替代:像气体火炬这类持续性工业
热源,在几乎每一次无云过境时都应被探测到,因此它会表现为一个几乎每天都复现的
像元。
:::

```{r}
#| label: fig-persist
#| fig-cap: "<span class='en'>How many distinct local days each 375 m pixel was detected on. No pixel recurs on more than six of the eight days, and the decay is smooth, the signature expected of spreading and smouldering fire, not of a static industrial source.</span><span class='zh'>每个 375 m 像元被探测到的不同本地日数。没有像元在八天中出现超过六天,且衰减平滑。这是蔓延与阴燃火所应有的特征,而非静态工业热源的特征。</span>"
#| fig-alt: "How many distinct local days each 375 m pixel was detected on. No pixel recurs on more than six of the eight days, and the decay is smooth, the signature expected of spreading and smouldering fire, not of a static industrial source."
#| fig-width: 7.4
#| fig-height: 2.9

persist |> count(n_days) |>
  ggplot(aes(factor(n_days), n)) +
  geom_col(fill = PAL$fire_l, width = 0.72) +
  geom_text(aes(label = n_fmt(n)), vjust = -0.4, size = 3.1, colour = PAL$grey) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.13))) +
  labs(title = "No candidate static heat source",
       subtitle = paste0(n_fmt(nrow(persist)),
                         " distinct 375 m pixels detected at least once in ",
                         N_DAYS, " days"),
       x = "Number of distinct local days on which the pixel was detected",
       y = "Pixels",
       caption = paste0("A gas flare or similar static source would be expected near ",
                        N_DAYS, " days. The maximum observed is ",
                        max(persist$n_days), ".")) +
  theme_th()
```

::: {.en}
`r n_fmt(sum(persist$n_days == 1))` of `r n_fmt(nrow(persist))` pixels
(`r sprintf("%.0f%%", 100*mean(persist$n_days == 1))`) were detected on a single
day, and the maximum is `r max(persist$n_days)` of `r N_DAYS`. There is no
evidence of a static source in this window, so the inability to apply the `type`
filter is judged not to bias the results. This is an empirical argument, not a
guarantee: a flare obscured by cloud on several days would be missed, and the
authenticated acquisition route (which returns the flag) would settle the
question.
:::

::: {.zh}
`r n_fmt(nrow(persist))` 个像元中有 `r n_fmt(sum(persist$n_days == 1))` 个
(`r sprintf("%.0f%%", 100*mean(persist$n_days == 1))`)只在**一天**被探测到,
最大值为 `r N_DAYS` 天中的 `r max(persist$n_days)` 天。本窗口内没有静态热源的证据,
因此判定无法应用 `type` 筛选不会使结果产生偏差。这是一个**经验性论证,而非保证**:
若某个火炬在若干天被云遮挡,就会被漏掉;而需认证的获取路径(会返回该标记)可以
彻底解决这个问题。
:::

## <span class="en">Duplicates and repeat observation</span><span class="zh">重复记录与重复观测</span>

```{r}
#| label: tbl-dups
#| tbl-cap: "<span class='en'>Coincident-point diagnostics for the three analysis units. Several spatstat summary functions assume a simple point pattern; the result is reported rather than left inside a warning.</span><span class='zh'>三种分析单元的重合点诊断。若干 spatstat 汇总函数假定点模式为简单点模式;此处明确报告结果,而不是把它留在一条警告信息里。</span>"
dup_rep |>
  gt::gt() |>
  gt::cols_label(unit = "Analysis unit", n_points = "Points",
                 n_duplicated = "Coincident points",
                 n_outside_window = "Rejected as outside window") |>
  gt::fmt_number(columns = where(is.numeric), decimals = 0) |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(5))
```

::: {.en}
Exact duplicate records (identical product, coordinate and timestamp) numbered
`r n_fmt(get_aud("exact duplicates")$n_removed)`. More interesting is that
**no two detections share an exact coordinate**, because the pixel centre
shifts slightly between overpasses as the scan geometry changes. The repeats are
therefore invisible to a naive duplicate check and must be found by snapping to
a grid:
:::

::: {.zh}
完全重复的记录(产品、坐标与时间戳三者全同)共
`r n_fmt(get_aud("exact duplicates")$n_removed)` 条。更值得注意的是,
**没有任何两条探测共享完全相同的坐标**,因为随着扫描几何变化,像元中心在不同过境
之间会有微小偏移。因此这些重复对朴素的去重检查是**不可见的**,必须通过对齐到格网
才能找出:
:::

```{r}
#| label: grid-snap
#| eval: false
xy <- sf::st_coordinates(fire_all_sf)
fire_all_sf <- fire_all_sf |>
  mutate(gx = floor(xy[, 1] / P$pixel_grid_m),     # P$pixel_grid_m = 375
         gy = floor(xy[, 2] / P$pixel_grid_m),
         pixel_id = paste(gx, gy, sep = "_"))
```

::: {.en}
Collapsing to one point per pixel per local day removes
`r n_fmt(N_DET - N_PX)` records (`r sprintf("%.1f%%", 100*(N_DET-N_PX)/N_DET)`),  same-day repeat overpasses of the same ground pixel.
:::

::: {.zh}
折叠为「每像元每本地日一个点」后移除了 `r n_fmt(N_DET - N_PX)` 条记录
(`r sprintf("%.1f%%", 100*(N_DET-N_PX)/N_DET)`),即同一地面像元在同一天的重复
过境。
:::

## <span class="en">How much of this pattern is the satellite?</span><span class="zh">这个模式里有多少是卫星造成的?</span> {#sec-crosssensor}

::: {.en}
Three near-identical VIIRS instruments flew over this window in the same eight
days. If detection were a faithful census of burning, they would largely agree.
:::

::: {.zh}
在同样的八天里,有三台近乎相同的 VIIRS 仪器飞越本窗口。如果探测是对燃烧的忠实
普查,它们本应大体一致。
:::

```{r}
#| label: fig-crosssensor
#| fig-cap: "<span class='en'>Agreement between the three VIIRS products. Left: how many of the three detected each fire-affected pixel-day. Right: daily detection counts by product, showing that day-to-day swings are not shared.</span><span class='zh'>三种 VIIRS 产品之间的一致性。左:每个过火像元-日被其中几种产品探测到。右:各产品的逐日探测数,显示逐日波动并不共享。</span>"
#| fig-alt: "Agreement between the three VIIRS products. Left: how many of the three detected each fire-affected pixel-day. Right: daily detection counts by product, showing that day-to-day swings are not shared."
#| fig-width: 10.5
#| fig-height: 3.6

agree <- fire_all |> sf::st_drop_geometry() |>
  dplyr::filter(grepl("VIIRS", firms_source)) |>
  distinct(pixel_id, date_local, firms_source) |>
  count(pixel_id, date_local, name = "n_products")

p1 <- agree |> count(n_products) |>
  mutate(pct = 100 * n / sum(n)) |>
  ggplot(aes(factor(n_products), n)) +
  geom_col(fill = PAL$water, width = 0.68) +
  geom_text(aes(label = sprintf("%s\n(%.0f%%)", n_fmt(n), pct)),
            vjust = -0.25, size = 3, lineheight = 0.9, colour = PAL$grey) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(title = "Most burning pixel-days are seen by only one of three satellites",
       x = "Number of VIIRS products detecting the pixel-day",
       y = "Pixel-days") +
  theme_th(11)

p2 <- FO_T$daily_sensor |>
  tidyr::pivot_longer(-date_local, names_to = "product", values_to = "n") |>
  mutate(product = stringr::str_replace(product, "_NRT", "")) |>
  ggplot(aes(date_local, n, colour = product)) +
  geom_line(linewidth = 0.75) + geom_point(size = 1.5) +
  scale_colour_manual(NULL, values = c(VIIRS_SNPP = PAL$fire,
                                       VIIRS_NOAA20 = PAL$fire_l,
                                       VIIRS_NOAA21 = PAL$water,
                                       MODIS = PAL$grey)) +
  scale_x_date(date_labels = "%d %b") +
  labs(title = "Daily counts disagree between near-identical sensors",
       x = NULL, y = "Detections in window") +
  theme_th(11)

p1 + p2 + patchwork::plot_layout(widths = c(1, 1.15))
```

::: {.interp .en}
**Where the sensors disagree.** Only
`r sprintf("%.1f%%", 100*mean(agree$n_products == 3))` of fire-affected
pixel-days were detected by all three VIIRS satellites, while
`r sprintf("%.0f%%", 100*mean(agree$n_products == 1))` were seen by just one.
The right panel shows the consequence: on 8 September, NOAA-20 recorded
`r FO_T$daily_sensor$VIIRS_NOAA20_NRT[FO_T$daily_sensor$date_local == as.Date("2026-09-08")]`
detections while NOAA-21 recorded
`r FO_T$daily_sensor$VIIRS_NOAA21_NRT[FO_T$daily_sensor$date_local == as.Date("2026-09-08")]`
over the same ground on the same day. **Evidence:** the instruments are
functionally identical, so the disagreement must come from overpass timing,
view geometry and, above all, cloud. **Implication:** detection is a heavily
sub-sampled view of the fire field, so day-to-day count variation cannot be read
as fire-activity variation, and absolute intensities are lower bounds
throughout. *Worth noting:* the three satellites' orbits differ, so some genuine
disagreement is expected even with perfect instruments.
:::

::: {.interp .zh}
**传感器分歧出现在哪里。** 过火像元-日中只有
`r sprintf("%.1f%%", 100*mean(agree$n_products == 3))` 被三颗 VIIRS 卫星全部探测到,
而 `r sprintf("%.0f%%", 100*mean(agree$n_products == 1))` 只被其中一颗看到。右图
显示了后果:9 月 8 日,NOAA-20 记录了
`r FO_T$daily_sensor$VIIRS_NOAA20_NRT[FO_T$daily_sensor$date_local == as.Date("2026-09-08")]`
条探测,而 NOAA-21 在同一天、同一片地面上记录了
`r FO_T$daily_sensor$VIIRS_NOAA21_NRT[FO_T$daily_sensor$date_local == as.Date("2026-09-08")]`
条。**证据:** 这些仪器在功能上是相同的,因此分歧必然来自过境时刻、观测几何,
以及最主要的云层遮挡。**含义:** 探测是对火场的**严重欠采样**视图,因此逐日计数
变化不能被读作火灾活动的变化,而全篇的绝对强度都只是下界。*需注意:* 三颗卫星轨道
不同,因此即使仪器完美也会有一部分真实分歧。
:::

## <span class="en">Projection and spatial filtering</span><span class="zh">投影与空间筛选</span> {#sec-crs}

::: {.en}
Coordinates are transformed from EPSG:4326 to **EPSG:32749 (WGS 84 / UTM zone
49S)**. The reasons, and the alternatives rejected:
:::

::: {.zh}
坐标由 EPSG:4326 转换为 **EPSG:32749(WGS 84 / UTM 49S 带)**。理由,以及被
否决的替代方案如下(下表为英文原表):
:::

| Requirement | Why | Outcome |
|---|---|---|
| Metre units | `spatstat` requires planar coordinates; bandwidths, distances and areas must be in a physical unit | Rules out any geographic CRS |
| Low area distortion | Intensity is events per unit area, so area error propagates directly into every density estimate | UTM 49S gives `r n_fmt(AREA_KM2)` km² against `r n_fmt(AREA_GEOD)` km² measured geodesically on the WGS84 ellipsoid, an error of **`r sprintf("%.2f%%", CRS_ERR_PCT)`** |
| Conformality | Isotropic kernels and isotropic distance functions assume distance means the same thing in all directions | Transverse Mercator is conformal; an equal-area projection would not be |
| Local suitability | The regency spans 113.58–114.35 °E | Mostly within UTM zone 49 (108–114 °E); the eastern sliver crossing 114 °E is the source of the residual distortion above |

::: {.en}
DGN95 / Indonesia TM-3 was considered, since it is the national system, but its
1.5°-wide zones place this regency awkwardly and it offers no practical
advantage here over UTM 49S at this level of distortion.

Points are then clipped to the window, removing
`r n_fmt(get_aud("outside study window")$n_removed)` detections that fell in the
buffered download box but outside the regency. Downloading a buffered box and
clipping afterwards (rather than requesting the exact polygon) keeps the
boundary decision in the analysis, where it is documented and testable, instead
of in the acquisition step.
:::

::: {.zh}
曾考虑 DGN95 / Indonesia TM-3(印尼国家坐标系),但其 1.5° 宽的分带使本县位置
尴尬,且在当前畸变水平下相对 UTM 49S 没有实际优势。

随后将点裁剪到窗口内,移除了
`r n_fmt(get_aud("outside study window")$n_removed)` 条落在缓冲下载框内但在县界之外
的探测。先下载缓冲框、事后裁剪(而不是直接请求精确多边形),是为了把边界决策**留在
分析环节**(在那里它有文档、可检验)而不是埋进数据获取环节。
:::

## <span class="en">The audit trail</span><span class="zh">审计轨迹</span>

```{r}
#| label: tbl-audit
#| tbl-cap: "<span class='en'>Complete record-level audit trail from raw download to analysed pattern.</span><span class='zh'>从原始下载到最终分析模式的完整记录级审计轨迹。</span>"
aud |>
  select(step, n_in, n_removed, pct_removed, n_out, reason) |>
  gt::gt() |>
  gt::cols_label(step = "Step", n_in = "In", n_removed = "Removed",
                 pct_removed = "%", n_out = "Out", reason = "Rationale") |>
  gt::fmt_number(columns = c(n_in, n_removed, n_out), decimals = 0) |>
  gt::fmt_number(columns = pct_removed, decimals = 1) |>
  gt::cols_width(step ~ gt::pct(17), reason ~ gt::pct(43)) |>
  gt::tab_style(style = gt::cell_text(weight = "bold"),
                locations = gt::cells_body(columns = n_out,
                                           rows = nrow(aud))) |>
  gt::tab_options(table.font.size = gt::pct(80), data_row.padding = gt::px(4))
```

::: {.en}
The pattern carried into analysis is **`r n_fmt(N_PX)` pixel-days**
(`r n_fmt(N_DET)` detections collapsed to unique pixel × local day) in a
`r n_fmt(AREA_KM2)` km² window over `r N_DAYS` days, a mean intensity
of `r sprintf("%.3f", LAM_BAR)` per km², or one fire-affected pixel-day per
`r sprintf("%.1f", 1/LAM_BAR)` km².
:::

::: {.zh}
进入分析的模式为 **`r n_fmt(N_PX)` 个像元-日**(由 `r n_fmt(N_DET)` 条探测折叠为
唯一的「像元 × 本地日」),窗口面积 `r n_fmt(AREA_KM2)` km²,时长
`r N_DAYS` 天,平均强度 `r sprintf("%.3f", LAM_BAR)` 个/km²,即每
`r sprintf("%.1f", 1/LAM_BAR)` km² 一个过火像元-日。
:::
