# <span class="en">Analytical assumptions and methodological decisions</span><span class="zh">分析假设与方法学决策</span> {#sec-assumptions}

::: {.en}
The objective of this section is not to claim that the assumptions hold. Most of
them do not hold exactly, and one of them (complete observation) fails badly
and demonstrably. The objective is to state each assumption, say whether it is
reasonable here, and say what its failure does to the conclusion that rests
on it.
:::

::: {.zh}
这一节**不是**要声称这些假设都成立。其中大多数并不严格成立,而有一条(完整
观测)明显且可证地失效了。它要做的是陈述每一条假设、说明它在此处是否合理,
并说明它失效会对依赖它的结论造成什么影响。
:::

## <span class="en">The foundational assumption, and how it fails</span><span class="zh">最基础的那个假设,以及它如何失效</span>

::: {.en}
Every method below treats the data as a realisation of a point process
$X$ on the window $W$, observed completely. Three parts of that need scrutiny.

**Events are points.** They are not; they are 375 m pixels, represented by their
centres. The approximation is harmless for questions posed at kilometre scales
and invalid below about 500 m (@sec-prep). Every interpretation below respects
that floor.

**The window is fixed and independent of the data.** Satisfied: the regency
boundary was fixed before the points were seen. This is why a convex or concave
hull of the detections was rejected as a window in @sec-window.

**Observation is complete.** *This fails.* @sec-crosssensor showed that only
`r sprintf("%.1f%%", 100*mean({a <- fire_all |> sf::st_drop_geometry() |> dplyr::filter(grepl("VIIRS", firms_source)) |> distinct(pixel_id, date_local, firms_source) |> count(pixel_id, date_local, name = "np"); a$np} == 3))`
of fire-affected pixel-days were seen by all three VIIRS satellites, and
@sec-fo-temporal shows one day in the series with no daytime detections at all.
Formally, what is observed is a *thinned* version of the fire process, with
retention probability $p(u, t)$ that depends on cloud, overpass geometry and
fire size.
:::

::: {.zh}
下文所有方法都把数据视为窗口 $W$ 上某个点过程 $X$ 的一次**完整观测**的实现。
其中有三个部分需要审视。

**事件是点。** 它们不是;它们是 375 m 像元,用其中心代表。这一近似对公里尺度的
问题无害,但在约 500 m 以下无效(@sec-prep)。下文每一处解读都遵守这一下界。

**窗口是固定的,且独立于数据。** 满足:县界在看到点之前就已固定。这正是
@sec-window 中把探测点的凸包/凹包排除在窗口候选之外的原因。

**观测是完整的。** *这一条失效。* @sec-crosssensor 表明,过火像元-日中只有
`r sprintf("%.1f%%", 100*mean({a <- fire_all |> sf::st_drop_geometry() |> dplyr::filter(grepl("VIIRS", firms_source)) |> distinct(pixel_id, date_local, firms_source) |> count(pixel_id, date_local, name = "np"); a$np} == 3))`
被三颗 VIIRS 卫星全部看到,而 @sec-fo-temporal 显示序列中有一天完全没有白天探测。
形式化地说,我们观测到的是火灾过程的一个**抽稀(thinned)**版本,其保留概率
$p(u, t)$ 取决于云、过境几何与火的规模。
:::

::: {.callout-important .en}
## What thinning does (and does not) do to the conclusions
Independent random thinning leaves the *inhomogeneous* second-order structure
unchanged: if $X$ has pair correlation $g(r)$, a thinned version has the same
$g(r)$ [@baddeley2015, ch. 7]. That is a genuinely useful robustness property,
and it is why the second-order conclusions in @sec-so survive an incomplete
observation process.

Thinning is *not* independent here, though. Cloud is spatially and temporally
autocorrelated over tens of kilometres and whole days. Two specific
consequences follow, and both are carried into the interpretation:

1. **First-order intensities are lower bounds**, and the ratio between high and
   low intensity areas is biased if cloud frequency differs between them.
2. **Day-to-day count variation is not fire-activity variation**, which is why
   @sec-fo-temporal reports the temporal analysis as being about the
   *observation-fire compound* rather than about fire.
:::

::: {.callout-important .zh}
## 抽稀会、以及不会对结论造成什么影响

**独立随机**抽稀不改变*非齐次*二阶结构:若 $X$ 的对相关函数为 $g(r)$,其抽稀版本
具有相同的 $g(r)$ [@baddeley2015, ch. 7]。这是一条确实有用的稳健性性质,也正是
@sec-so 中二阶结论能在不完整观测下依然成立的原因。

然而此处的抽稀**并非独立**。云在数十公里、整天的尺度上具有空间与时间自相关。
由此有两个具体后果,二者都被带入了解读:

1. **一阶强度只是下界**,而且如果高强度区与低强度区的云频率不同,两者之间的强度
   **比值**也会有偏。
2. **逐日计数变化不等于火灾活动变化**。这正是 @sec-fo-temporal 把时间分析表述为
   关于*观测, 火复合体*而不是关于火本身的原因。
:::

## <span class="en">Assumptions by method</span><span class="zh">各方法所需的假设</span>

```{r}
#| label: tbl-assumptions
#| tbl-cap: "<span class='en'>Assumptions required by each method, an assessment for this study, and the consequence if the assumption fails.</span><span class='zh'>各方法所需的假设、针对本研究的评估,以及该假设失效后的后果。(下表为英文原表)</span>"

tibble::tribble(
  ~Method, ~Assumption, ~Holds, ~`Consequence and mitigation`,

  "Quadrat $\\chi^2$ test", "Homogeneous Poisson process (constant intensity, independent points)",
  "**No**",
  "Rejection is uninformative: it cannot distinguish intensity variation from interaction. Reported only to demonstrate that fact, across four quadrat sizes.",

  "Fixed-bandwidth KDE", "A single bandwidth is appropriate everywhere; kernel is isotropic",
  "**Partly**",
  "Intensity spans two orders of magnitude, so one bandwidth must oversmooth peaks or undersmooth sparse areas. Mitigated with a bandwidth ladder and an adaptive-bandwidth estimate.",

  "KDE edge correction", "The process continues beyond $W$ in the same way",
  "**Partly**",
  "True on land borders, false on the southern coast, where there is no process beyond the boundary. Coastal intensities are therefore over-corrected upward.",

  "$G$, $F$, $J$ functions", "Stationarity and isotropy",
  "**No**",
  "Under strong inhomogeneity these functions confound intensity variation with interaction. Used descriptively; no interaction claim rests on them.",

  "Ripley's $K$ vs CSR", "Stationarity and isotropy",
  "**No**",
  "Guaranteed rejection. Included as an explicit demonstration of why an inhomogeneous null is required.",

  "$K_{inhom}$, $g_{inhom}$", "**Scale separation**: intensity varies only at scales $\\ge \\sigma_\\lambda$; interaction operates below it",
  "*Unverifiable*",
  "The central assumption of this report. Cannot be checked from one realisation, so results are reported across $\\sigma_\\lambda \\in \\{1, 2, 5, 10\\}$ km and the conclusion is stated as a function of it (@sec-so-kinhom).",

  "$K_{inhom}$ intensity estimate", "$\\hat\\lambda$ is not contaminated by the pattern it describes",
  "**Partly**",
  "Leave-one-out kernel estimation and `normpower = 2` reduce, but do not remove, the bias. Direction of bias is towards the null, so a positive finding is conservative.",

  "Isotropy of interaction", "Interaction depends on distance, not direction",
  "**Doubtful**",
  "Canals form an oriented grid, so fire may spread preferentially along them. An isotropic $g(r)$ would then understate the true along-canal correlation. Flagged as a limitation, not resolved.",

  "Knox space-time test", "Under the null, dates are exchangeable across locations",
  "**Yes, by construction**",
  "This is the strength of the test: it requires no model for either the spatial or the temporal marginal pattern. The cost is that 'interaction' here includes first-order space-time inseparability (a moving fire front), not only contagion.",

  "`ppm` (Poisson pseudolikelihood)", "Conditional independence of points given the trend",
  "**No**",
  "Point estimates of the trend remain consistent, but standard errors are **understated**, possibly severely. Coefficients are therefore interpreted for sign and magnitude, and the clustering is modelled explicitly in `kppm`.",

  "Thomas cluster process", "Parents are a homogeneous Poisson process; offspring are isotropic Gaussian scatter",
  "**Approximately**",
  "Anisotropic, canal-aligned spread violates the isotropic-offspring assumption, so the fitted `scale` should be read as an effective cluster radius rather than a physical one."
) |>
  gt::gt() |>
  gt::fmt_markdown(columns = c(Assumption, Holds, `Consequence and mitigation`)) |>
  gt::cols_width(Method ~ gt::pct(16), Assumption ~ gt::pct(25),
                 Holds ~ gt::pct(9), `Consequence and mitigation` ~ gt::pct(50)) |>
  gt::tab_options(table.font.size = gt::pct(78), data_row.padding = gt::px(6))
```

## <span class="en">Null models</span><span class="zh">零模型</span> {#sec-nullmodels}

::: {.en}
A second-order result means nothing without the reference it is measured
against. Five null models are used, each answering a different question.
:::

::: {.zh}
二阶结果若没有被测量所依据的参照,就毫无意义。分析共使用五个零模型,各自回答
一个不同的问题。
:::

```{r}
#| label: tbl-nulls
#| tbl-cap: "<span class='en'>The five reference models used in this report, what each holds fixed, and what rejecting it licenses.</span><span class='zh'>本报告使用的五个参照模型、各自固定了什么,以及拒绝它能够支持何种论断。(下表为英文原表)</span>"

tibble::tribble(
  ~Null, ~`Simulation mechanism`, ~`Holds fixed`, ~`Rejection licenses the claim that...`,
  "**N1** Complete spatial randomness",
  "`runifpoint(n, W)`", "$n$, $W$",
  "...the pattern is not uniform-random. Almost vacuous here.",

  "**N2** Inhomogeneous Poisson",
  "`rpoispp(lambda_hat)`", "$W$, the kernel intensity surface at $\\sigma_\\lambda$",
  "...clustering exists **beyond** intensity variation at scales above $\\sigma_\\lambda$. The substantive test.",

  "**N3** Random date labelling",
  "`sample(dates)` over fixed locations", "All locations, all dates, both marginals",
  "...where and when are dependent, genuine space-time interaction.",

  "**N4** Fitted inhomogeneous Poisson (`ppm`)",
  "`simulate(M1)`", "$W$, the covariate-driven trend",
  "...measured drainage/access geography does not account for the clustering.",

  "**N5** Fitted Thomas cluster process (`kppm`)",
  "`simulate(M3)`", "$W$, trend, and a clustering mechanism",
  "...even a trend-plus-clustering model is inadequate."
) |>
  gt::gt() |> gt::fmt_markdown(columns = everything()) |>
  gt::cols_width(Null ~ gt::pct(19), `Simulation mechanism` ~ gt::pct(20),
                 `Holds fixed` ~ gt::pct(24),
                 `Rejection licenses the claim that...` ~ gt::pct(37)) |>
  gt::tab_options(table.font.size = gt::pct(80), data_row.padding = gt::px(6))
```

## <span class="en">Methodological decisions</span><span class="zh">方法学决策</span> {#sec-decisions}

```{r}
#| label: tbl-decisions
#| tbl-cap: "<span class='en'>Every parameter that could materially change a result, its value, and why.</span><span class='zh'>所有可能实质性改变结果的参数、取值及理由。(下表为英文原表)</span>"

tibble::tribble(
  ~Decision, ~Value, ~Justification,
  "Coordinate reference system", "EPSG:32749 (WGS 84 / UTM 49S)",
  "Metre units, conformal, 0.26% area error over this window (@sec-crs).",

  "Point-event definition", "Unique 375 m pixel x local day",
  "Removes instrumental repeat-overpass duplication, retains genuine day-to-day persistence (@sec-event-def). All second-order results repeated on two alternative definitions.",

  "Temporal aggregation", "Local calendar day (WIB, UTC+07)",
  "Finer aggregation would describe the satellite orbit; two overpasses per day is the sampling limit (@sec-timezone).",

  "KDE bandwidth (headline)", "$\\sigma$ = 2,000 m",
  "Above the ~375-500 m sensor resolution floor, below the ~20 km kecamatan scale. Automatic selectors returned 337 m to 13,688 m; see @sec-fo-bandwidth for why two of them are rejected.",

  "KDE bandwidth (reported range)", "500, 1,000, 2,000, 5,000, 10,000 m",
  "Sensitivity ladder shown in full, so the reader can see which conclusions depend on the choice.",

  "Adaptive KDE", "Abramson rule, $h_0$ = 2,000 m",
  "Allows the bandwidth to shrink where data are dense, which a fixed bandwidth cannot [@abramson1982].",

  "Covariate raster resolution", "250 m",
  "Finer than the 375 m sensor pixel, so the covariate is never the coarser layer; coarse enough to refit models repeatedly.",

  "$r_{max}$ for $K$, $K_{inhom}$", "15 km",
  paste0("Below one quarter of the window's shorter dimension (~",
         sprintf("%.0f", diff(win$xrange)/1000), " km), the conventional ceiling for reliable estimation."),

  "$r_{max}$ for $G$, $F$", "3 km",
  "Nearest-neighbour distances are almost all far below this; larger $r$ adds only edge-corrected noise.",

  "$r_{max}$ for $g_{inhom}$", "8 km",
  "Interaction is expected to vanish well within this range, and $g$ is noisier than $K$.",

  "Edge correction, $K$ / $g$", "Ohser-Stoyan translation",
  paste0("Unbiased, and ~20x faster than the isotropic correction on this ",
         n_fmt(nrow(sf::st_coordinates(bnd))),
         "-vertex polygon, which is decisive at 199 simulations x 4 bandwidths."),

  "Edge correction, $G$ / $F$ / $J$", "Kaplan-Meier",
  "The recommended correction for distance distributions in an irregular window [@baddeley2015].",

  "Intensity normalisation in $K_{inhom}$", "`normpower = 2`",
  "Reduces sensitivity to error in $\\hat\\lambda$ [@baddeley2000]; the spatstat default changed in version 3.8-1, so it is set explicitly for reproducibility.",

  "Envelope type", "**Global**, 199 simulations",
  "Pointwise envelopes have no valid overall significance level because they are read at ~200 values of $r$ simultaneously [@baddeley2014envelope]. Global envelopes give an exact level of 2/(199+1) = 1%.",

  "Formal test statistics", "DCLF (integrated deviation) and MAD",
  "One $p$-value per hypothesis rather than a curve that invites selective reading [@myllymaki2017].",

  "Monte Carlo replicates", "999 (quadrat, Knox), 199 (envelopes), 499 (lag test)",
  "Set by the smallest $p$-value each test needs to resolve, balanced against runtime.",

  "Random seed", "`set.seed(1234)`",
  "Every Monte-Carlo result in this report is exactly reproducible."
) |>
  gt::gt() |> gt::fmt_markdown(columns = c(Value, Justification)) |>
  gt::cols_width(Decision ~ gt::pct(21), Value ~ gt::pct(20),
                 Justification ~ gt::pct(59)) |>
  gt::tab_options(table.font.size = gt::pct(79), data_row.padding = gt::px(6))
```
