# <span class="en">Second-order analysis: do events interact?</span><span class="zh">二阶分析:事件之间有交互吗?</span> {#sec-so}

::: {.en}
First-order analysis established that intensity varies enormously across the
window. Second-order analysis asks a different question: **given that variation,
do detections still attract one another?**

The answer turns out to depend almost entirely on what "given that variation"
means, and this section's main finding is *why*. Three reference models are
used in order of increasing stringency, and they do not agree. Understanding the
disagreement is more informative than any one of the three tests.
:::

::: {.zh}
一阶分析已确立强度在窗口内变化极大。二阶分析问的是另一个问题:**在给定这种变化
之后,探测点之间是否仍然相互吸引?**

结果表明,答案几乎完全取决于「给定这种变化」是什么意思,而这一节的**主要发现正是
「为什么」**。此处按严格程度递增使用三个参照模型,而它们**并不一致**。理解这种
不一致,比其中任何单个检验都更有信息量。
:::

## <span class="en">Parameters common to this section</span><span class="zh">本节通用参数</span> {#sec-so-params}

```{r}
#| label: so-params
#| code-fold: true
#| code-summary: "Estimation settings"

NSIM <- SO$nsim          # 199 simulations -> global envelope level 2/200 = 1%
CORR <- SO$corr          # Ohser-Stoyan translation correction
r_K  <- SO$r_K           # 0 to 15 km
r_G  <- SO$r_G           # 0 to 3 km

tibble::tibble(
  Setting = c("Analysis unit", "Simulations per envelope", "Envelope type",
              "Edge correction (K, L, g)", "Edge correction (G, F, J)",
              "r range (L)", "r range (G, F)", "Window bounding box"),
  Value = c(
    paste0("pixel-day, n = ", n_fmt(N_PX)),
    paste0(NSIM, " (+ ", NSIM, " for the reference mean where the null is not CSR)"),
    "global (constant width), α = 0.01",
    CORR, "Kaplan–Meier",
    paste0("0 – ", n_fmt(max(r_K)), " m"),
    paste0("0 – ", n_fmt(max(r_G)), " m"),
    sprintf("%.0f × %.0f km", diff(win$xrange)/1000, diff(win$yrange)/1000))
) |>
  gt::gt() |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(4))
```

::: {.en}
Two choices need a word of explanation.

**Why global envelopes.** A pointwise envelope is read at every one of ~200
distances, so its nominal 1% level applies to each $r$ separately and the
probability of *some* excursion under the null is far higher than 1%
[@baddeley2014envelope]. A global envelope has a constant width chosen so that
the probability of the observed curve leaving it *anywhere* in the $r$ range is
exactly $2/(199+1) = 1\%$. Every envelope in this section is global, and each is
accompanied by a single $p$-value from a Diggle–Cressie–Loosmore–Ford (DCLF) or
maximum-absolute-deviation (MAD) test, so that no conclusion depends on
selectively reading a curve [@myllymaki2017].

**Why $L$ rather than $K$.** A constant-width band applied to $K(r)$, whose
variance grows like $r^4$, is set by the largest distances and is effectively
blind at small $r$. All envelopes therefore use the variance-stabilised
$L(r) = \sqrt{K(r)/\pi}$, which under a Poisson process equals $r$. Plots below
show $L(r)$ **minus its reference curve**, so a horizontal line at zero is the
null and any departure is read directly in metres.
:::

::: {.zh}
有两项选择需要说明。

**为什么用全局包络。** 逐点包络会在约 200 个距离上分别被读取,因此其名义 1% 水平
只对每个 $r$ 单独成立,而在零模型下曲线*在某处*越界的概率远高于 1%
[@baddeley2014envelope]。全局包络的宽度是常数,并使观测曲线在整个 $r$ 范围内
*任何地方*越界的概率恰好为 $2/(199+1) = 1\%$。以下所有包络都是全局的,并且每个都
配有一个来自 Diggle–Cressie–Loosmore–Ford(DCLF)或最大绝对偏差(MAD)检验的单一
$p$ 值,从而使任何结论都不依赖于对曲线的选择性阅读 [@myllymaki2017]。

**为什么用 $L$ 而不用 $K$。** 把常宽带用在 $K(r)$ 上时,由于其方差按 $r^4$ 增长,
带宽实际由最大距离决定,在小 $r$ 处近乎失明。因此所有包络都使用方差稳定化的
$L(r) = \sqrt{K(r)/\pi}$,它在泊松过程下等于 $r$。下文各图显示的是 $L(r)$
**减去其参照曲线**,因此零水平线就是零模型,任何偏离都可以直接以米读取。
:::

## <span class="en">N1: against complete spatial randomness, and why it settles nothing</span><span class="zh">N1:对照完全空间随机性,以及它为何解决不了任何问题</span> {#sec-so-csr}

::: {.rq .en}
**Question tested.** *Is the pattern distinguishable from a uniform random
scatter of the same number of points in the same window?*
**Null model N1:** homogeneous Poisson, simulated by `runifpoint(n, W)`.
:::

::: {.rq .zh}
**所检验的问题。** *该模式能否与同一窗口内、同样点数的均匀随机散布区分开?*
**零模型 N1:** 齐次泊松过程,由 `runifpoint(n, W)` 模拟。
:::

```{r}
#| label: fig-gfj
#| fig-cap: "<span class='en'>Nearest-neighbour (G), empty-space (F) and J functions against complete spatial randomness. All three reject CSR decisively at every distance examined.</span><span class='zh'>最近邻(G)、空隙(F)与 J 函数对照完全空间随机性。三者在所考察的每个距离上都决定性地拒绝 CSR。</span>"
#| fig-alt: "Nearest-neighbour (G), empty-space (F) and J functions against complete spatial randomness. All three reject CSR decisively at every distance examined."
#| fig-width: 11
#| fig-height: 3.5

gf <- function(e, ttl, sub, ylab) {
  d <- as.data.frame(e)
  ggplot(d, aes(r / 1000)) +
    geom_ribbon(aes(ymin = lo, ymax = hi, fill = "band"), alpha = 0.9) +
    geom_line(aes(y = theo, colour = "theo"), linetype = "22", linewidth = 0.5) +
    geom_line(aes(y = obs, colour = "obs"), linewidth = 0.85) +
    scale_fill_manual(NULL, values = c(band = PAL$env),
                      labels = "Global envelope") +
    scale_colour_manual(NULL, values = c(obs = PAL$fire, theo = PAL$theo),
                        labels = c(obs = "Observed", theo = "CSR"),
                        breaks = c("obs", "theo")) +
    labs(title = ttl, subtitle = sub, x = "Distance r (km)", y = ylab) +
    theme_th(10)
}

gf(SO$env_G, "G: nearest-neighbour distances",
   "Observed far ABOVE CSR → neighbours much closer than random",
   "G(r)") +
gf(SO$env_F, "F: empty-space distances",
   "Observed far BELOW CSR → large empty gaps exist",
   "F(r)") +
gf(SO$env_J, "J = (1−G)/(1−F)",
   "J < 1 throughout → clustering, not inhibition",
   "J(r)") +
  patchwork::plot_layout(nrow = 1, guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-lcsr
#| fig-cap: "<span class='en'>L function against CSR, centred on the theoretical value. The global envelope is so narrow relative to the observed excess that it is barely visible.</span><span class='zh'>L 函数对照 CSR,以理论值为中心。相对观测到的超额,全局包络窄到几乎看不见。</span>"
#| fig-alt: "L function against CSR, centred on the theoretical value. The global envelope is so narrow relative to the observed excess that it is barely visible."
#| fig-width: 9
#| fig-height: 3.8

plot_env_centred(
  SO$env_L_csr,
  title = "N1: L(r) against complete spatial randomness",
  subtitle = paste0("DCLF u = ",
                    formatC(stat_of(SO$test_L_csr), format = "g", digits = 3),
                    ", p ", pv(SO$test_L_csr$p.value),
                    "  ·  MAD p ", pv(SO$mad_L_csr$p.value),
                    "  ·  ", NSIM, " simulations"),
  caption = "This rejection was a foregone conclusion: the first-order analysis already showed intensity varying ~40-fold.",
  ref_lab = "CSR")
```

```{r}
#| label: nnd-numbers
#| echo: false
nnd_ratio <- SO$nnd_obs / mean(SO$nnd_csr)
```

::: {.en}
Every statistic rejects CSR at the smallest attainable $p$-value
(`r pv(SO$test_L_csr$p.value)`, the floor for `r NSIM` simulations). The mean
nearest-neighbour distance is `r sprintf("%.0f", SO$nnd_obs)` m against
`r sprintf("%.0f", mean(SO$nnd_csr))` m expected under CSR (simulated
2.5–97.5% interval `r sprintf("%.0f", quantile(SO$nnd_csr, .025))`–`r sprintf("%.0f", quantile(SO$nnd_csr, .975))` m),  a ratio of `r sprintf("%.2f", nnd_ratio)`, so observed neighbours are about
`r sprintf("%.1f", 1/nnd_ratio)` times closer together than random.
:::

::: {.zh}
每一个统计量都在所能达到的最小 $p$ 值上拒绝 CSR
(`r pv(SO$test_L_csr$p.value)`,即 `r NSIM` 次模拟的下限)。平均最近邻距离为
`r sprintf("%.0f", SO$nnd_obs)` m,而 CSR 下的期望为
`r sprintf("%.0f", mean(SO$nnd_csr))` m(模拟 2.5–97.5% 区间
`r sprintf("%.0f", quantile(SO$nnd_csr, .025))`–`r sprintf("%.0f", quantile(SO$nnd_csr, .975))` m),比值为 `r sprintf("%.2f", nnd_ratio)`,即观测到的邻居之间的距离约为随机情形的
`r sprintf("%.1f", 1/nnd_ratio)` 分之一。
:::

::: {.callout-warning .en}
## This tells us almost nothing
The result is real but uninformative, and the reason is structural, not
statistical. CSR bundles two hypotheses (*constant intensity* and
*independent points*) into one null. The first-order analysis already
demonstrated that constant intensity is false, with intensity ranging over
roughly `r sprintf("%.0f", FO_K$int_check$ratio_max_to_mean[which(FO_K$sigma_ladder == FO_K$sigma_main)])`-fold
around the mean. Rejection of N1 was therefore guaranteed *before any
second-order statistic was computed*, and it cannot be attributed to interaction.
:::

::: {.callout-warning .zh}
## 这几乎没有告诉我们任何东西

这个结果是真实的,但没有信息量,而原因是**结构性的**而非统计性的。CSR 把两个假设(*常数强度*与*点相互独立*)捆进了一个零模型。一阶分析已经证明常数强度不成立:
强度围绕均值的变化幅度约
`r sprintf("%.0f", FO_K$int_check$ratio_max_to_mean[which(FO_K$sigma_ladder == FO_K$sigma_main)])`
倍。因此对 N1 的拒绝在*任何二阶统计量被计算之前*就已经注定,而且它**无法**被归因于
交互作用。
:::

## <span class="en">N2: against an inhomogeneous Poisson process</span><span class="zh">N2:对照非齐次泊松过程</span> {#sec-so-kinhom}

::: {.rq .en}
**Question tested.** *Once the large-scale intensity gradient is taken as given,
do detections still cluster beyond what that gradient explains?*
**Null model N2:** inhomogeneous Poisson with intensity fixed at the kernel
estimate $\hat\lambda_{\sigma_\lambda}$, simulated by `rpoispp(lambda_hat)`.
**Statistic:** $L_{inhom}(r)$, weighting each pair by
$1/(\hat\lambda(x_i)\hat\lambda(x_j))$ so that intensity variation is divided
out [@baddeley2000].
:::

::: {.rq .zh}
**所检验的问题。** *在把大尺度强度梯度视为已给定之后,探测点是否仍然聚集到超出
该梯度所能解释的程度?*
**零模型 N2:** 非齐次泊松过程,其强度固定为核估计
$\hat\lambda_{\sigma_\lambda}$,由 `rpoispp(lambda_hat)` 模拟。
**统计量:** $L_{inhom}(r)$,它对每一对点按
$1/(\hat\lambda(x_i)\hat\lambda(x_j))$ 加权,从而把强度变化除掉
[@baddeley2000]。
:::

::: {.en}
Two implementation decisions make the comparison legitimate, and both were
arrived at after a first attempt failed.

1. **A single fixed intensity.** `Linhom` will re-estimate $\hat\lambda$
   separately for each simulated pattern by leave-one-out kernel smoothing.
   That was tried first and rejected: it makes observed and simulated statistics
   incommensurable, and on this pattern it was numerically degenerate, the
   simulated $L_{inhom}$ collapsed to zero. Since the null model *is* "an
   inhomogeneous Poisson process with intensity $\hat\lambda$", $\hat\lambda$ is
   estimated once from the data and the same image is used throughout.
2. **`normpower = 2`**, the normalisation recommended to reduce sensitivity to
   error in $\hat\lambda$ [@baddeley2000]. The spatstat default changed in
   version 3.8-1, so it is set explicitly.
:::

::: {.zh}
有两项实现决策使这一比较得以成立,而两者都是在第一次尝试失败之后才确定的。

1. **使用单一固定强度。** `Linhom` 会用留一核平滑为**每个**模拟模式分别重新估计
   $\hat\lambda$。这一做法先被尝试、随后被否决:它使观测统计量与模拟统计量不可
   比较,并且在本模式上数值退化,模拟出的 $L_{inhom}$ 坍缩为零。既然零模型**就是**
   「强度为 $\hat\lambda$ 的非齐次泊松过程」,那么 $\hat\lambda$ 就从数据中估计
   **一次**,并在全过程中使用同一张影像。
2. **`normpower = 2`**,即推荐用于降低对 $\hat\lambda$ 估计误差敏感性的归一化方式
   [@baddeley2000]。该参数的 spatstat 默认值在 3.8-1 版中有过变更,因此此处显式
   设定。
:::

### <span class="en">The result: the answer is a function of $\sigma_\lambda$</span><span class="zh">结果:答案是 $\sigma_\lambda$ 的函数</span>

```{r}
#| label: fig-linhom
#| fig-cap: "<span class='en'>Inhomogeneous L against an inhomogeneous Poisson null at four intensity bandwidths. The observed curve leaves the global envelope only at σ_λ = 10 km.</span><span class='zh'>四种强度带宽下,非齐次 L 对照非齐次泊松零模型。观测曲线只在 σ_λ = 10 km 时离开全局包络。</span>"
#| fig-alt: "Inhomogeneous L against an inhomogeneous Poisson null at four intensity bandwidths. The observed curve leaves the global envelope only at σ_λ = 10 km."
#| fig-width: 11
#| fig-height: 6.4

lad <- SO$linhom_ladder
panels <- lapply(seq_along(lad), function(k) {
  x <- lad[[k]]; s <- env_summary(x$env)
  plot_env_centred(
    x$env,
    title = bquote(sigma[lambda] == .(n_fmt(x$sigma)) ~ "m"),
    subtitle = sprintf("DCLF p %s · %s",
                       pv(x$test$p.value),
                       ifelse(is.na(s$r_first), "inside the envelope throughout",
                              sprintf("outside over %.1f–%.1f km",
                                      s$r_first, s$r_last))),
    ref_lab = "Inhomogeneous Poisson") +
    theme_th(10) + theme(legend.position = "none")
})
patchwork::wrap_plots(panels, nrow = 2)
```

```{r}
#| label: tbl-linhom
#| tbl-cap: "<span class='en'>The N2 test as a function of the intensity bandwidth, alongside a count of the resolution cells the null intensity consumes. The two columns move together.</span><span class='zh'>N2 检验作为强度带宽的函数,并列出零模型强度所消耗的分辨单元数。这两列是同步变化的。(下表为英文原表)</span>"

A_KM2 <- area(win) / 1e6
lin_tab <- bind_rows(lapply(lad, function(x) {
  s <- env_summary(x$env)
  cells <- A_KM2 / (pi * (x$sigma / 1000)^2)
  tibble(`σ_λ (m)` = x$sigma,
         `Resolution cells in the null` = cells,
         `Points per cell` = N_PX / cells,
         `DCLF p` = x$test$p.value,
         `MAD p` = x$mad$p.value,
         `Outside envelope` = ifelse(is.na(s$r_first), "—",
                                     sprintf("%.1f–%.1f km", s$r_first, s$r_last)))
}))
lin_tab |>
  gt::gt() |>
  gt::fmt_number(columns = c(`σ_λ (m)`, `Resolution cells in the null`),
                 decimals = 0) |>
  gt::fmt_number(columns = `Points per cell`, decimals = 2) |>
  gt::fmt_number(columns = c(`DCLF p`, `MAD p`), decimals = 4) |>
  gt::tab_footnote(
    footnote = gt::md(paste0(
      "*Resolution cells* is the window area divided by the kernel footprint ",
      "$\\pi\\sigma_\\lambda^2$, a deliberately rough count of how many ",
      "independent values the kernel intensity estimate is free to choose. ",
      "It is an order-of-magnitude heuristic, not an exact effective degrees ",
      "of freedom.")),
    locations = gt::cells_column_labels(columns = `Resolution cells in the null`)) |>
  gt::tab_options(table.font.size = gt::pct(82), data_row.padding = gt::px(6))
```

::: {.en}
The pattern in that table is the central methodological finding of this report.
:::

::: {.zh}
上表中的规律正是这项分析在方法学上的核心发现。
:::

```{r}
#| label: fig-power
#| fig-cap: "<span class='en'>Why the N2 test reaches opposite conclusions at different bandwidths: the flexibility of the null intensity, not the data, decides the answer.</span><span class='zh'>N2 检验在不同带宽下得出相反结论的原因:决定答案的是零模型强度的灵活度,而不是数据。</span>"
#| fig-alt: "Why the N2 test reaches opposite conclusions at different bandwidths: the flexibility of the null intensity, not the data, decides the answer."
#| fig-width: 9.4
#| fig-height: 4.2

pw <- lin_tab |>
  transmute(sigma = `σ_λ (m)`, cells = `Resolution cells in the null`,
            p = `DCLF p`) |>
  bind_rows(tibble(sigma = NA_real_, cells = VA$n_par_M1,
                   p = VA$gof_M1_test$p.value))
pw$label <- c(sprintf("σ_λ = %s m", n_fmt(pw$sigma[1:4])),
              "covariate model (N4, §7)")

ggplot(pw, aes(cells, p)) +
  annotate("rect", xmin = 1, xmax = Inf, ymin = 0.05, ymax = Inf,
           fill = PAL$grey_l, alpha = 0.35) +
  annotate("text", x = 3.2, y = 0.5, label = "cannot reject the null",
           hjust = 0, size = 3.1, colour = PAL$grey) +
  geom_hline(yintercept = 0.05, linetype = "22", colour = PAL$theo) +
  geom_line(data = ~dplyr::filter(.x, !is.na(sigma)), colour = PAL$fire,
            linewidth = 0.7) +
  geom_point(aes(colour = is.na(sigma)), size = 3.4) +
  # geom_label_repel rather than text: these labels sit over a line and two
  # reference rules, and an opaque background keeps both readable.
  ggrepel::geom_label_repel(aes(label = label), size = 3.1, colour = "#292524",
                            seed = 5, min.segment.length = 0.1,
                            box.padding = 0.65, point.padding = 0.35,
                            label.size = 0, label.padding = unit(0.13, "lines"),
                            fill = alpha("white", 0.9),
                            segment.size = 0.3, segment.colour = PAL$grey) +
  geom_hline(yintercept = 0, colour = NA) +
  scale_x_log10(labels = label_comma()) +
  scale_y_log10(labels = label_number(accuracy = 0.0001)) +
  scale_colour_manual(values = c(`FALSE` = PAL$fire, `TRUE` = PAL$water),
                      guide = "none") +
  annotate("segment", x = 2715, xend = 2715, y = 0.0015, yend = 1,
           linetype = "33", colour = PAL$water) +
  annotate("text", x = 2715, y = 0.0013, label = "n = 2,715 points",
           size = 2.9, colour = PAL$water, vjust = 1) +
  labs(title = "Power depends on how many parameters the null is allowed",
       subtitle = paste0("The same data, the same statistic, the same 199 ",
                         "simulations — only the null's flexibility changes"),
       x = "Resolution cells / parameters in the null intensity (log scale)",
       y = "DCLF p-value (log scale)") +
  theme_th(11)
```

::: {.interp .en}
**Resolution against sample size.** At $\sigma_\lambda = 1$ km the kernel intensity has
roughly `r n_fmt(round(lin_tab[["Resolution cells in the null"]][1]))` resolution
cells for `r n_fmt(N_PX)` points, **fewer than one point per cell**. Such a null
is close to saturated: it can reproduce almost any pattern, so failing to reject
it (p = `r sprintf("%.3f", lin_tab[["DCLF p"]][1])`) carries little information.
As $\sigma_\lambda$ grows, the null becomes parsimonious and the $p$-value falls
monotonically to `r sprintf("%.4f", lin_tab[["DCLF p"]][4])` at 10 km, where
excess clustering appears over
`r sprintf("%.1f–%.1f km", env_summary(lad[[4]]$env)$r_first, env_summary(lad[[4]]$env)$r_last)`.
The covariate model of @sec-va, with only `r VA$n_par_M1` parameters,
sits at the far left and is rejected decisively. **Implication:** whether the
~2 km structure in this pattern is called "intensity variation" or "clustering"
is a **modelling choice**, not something these data can settle.
*But note:* "resolution cells" is a rough heuristic for the null's flexibility,
not an exact degrees-of-freedom calculation.
:::

::: {.interp .zh}
**分辨单元数与样本量。** 在 $\sigma_\lambda = 1$ km 时,核强度约有
`r n_fmt(round(lin_tab[["Resolution cells in the null"]][1]))` 个分辨单元,而点只有
`r n_fmt(N_PX)` 个,**每个单元不到一个点**。这样的零模型接近**饱和**:它几乎能
再现任何模式,因此「未能拒绝它」(p = `r sprintf("%.3f", lin_tab[["DCLF p"]][1])`)
所携带的信息很少。随着 $\sigma_\lambda$ 增大,零模型变得简约,$p$ 值单调下降到
10 km 时的 `r sprintf("%.4f", lin_tab[["DCLF p"]][4])`,此时在
`r sprintf("%.1f–%.1f km", env_summary(lad[[4]]$env)$r_first, env_summary(lad[[4]]$env)$r_last)`
范围内出现超额聚集。@sec-va 中仅有 `r VA$n_par_M1` 个参数的协变量模型位于
最左端,并被决定性地拒绝。**含义:** 本模式中那个约 2 km 的结构,称之为「强度变化」
还是「聚集」,是一个**建模选择**,而不是这些数据能够判定的事实。
*但请注意:*「分辨单元」只是对零模型灵活度的粗略启发式度量,不是精确的自由度计算。
:::

::: {.callout-important .en}
## The honest statement of the N2 result
At the bandwidth chosen for the headline intensity surface
($\sigma_\lambda = `r n_fmt(SO$sigma_main)`$ m), **the observed pattern is not
distinguishable from an inhomogeneous Poisson process**
(DCLF $p =$ `r sprintf("%.3f", lad[[which(SO$sig_ladder == SO$sigma_main)]]$test$p.value)`,
MAD $p =$ `r sprintf("%.3f", lad[[which(SO$sig_ladder == SO$sigma_main)]]$mad$p.value)`,
observed curve inside the 1% global envelope at every distance).

It would be wrong to report this as "no clustering". It is a statement about a
test with very little power, for two identifiable reasons:

1. **The null is nearly saturated** at this bandwidth, as above.
2. **The inverse-intensity weights are unbounded.** $L_{inhom}$ weights pairs by
   $1/(\hat\lambda_i\hat\lambda_j)$, and at $\sigma_\lambda = 2$ km the fitted
   intensity at the data points spans a factor of about
   `r {lamx <- density(ppp_px, sigma = SO$sigma_main, edge = TRUE, positive = TRUE, dimyx = c(512,256))[win, drop = FALSE]; v <- lamx[ppp_px]; sprintf("%.0f", max(v)/min(v))}`.
   A simulated pair falling in the sparse north therefore contributes an
   enormous amount, giving the null distribution very heavy tails. The
   consequence is visible in @fig-pcf, where the global envelope for $g_{inhom}$
   is roughly ±200 around a null value of 1.

The way out is not a better bandwidth. It is a **null model with few
parameters**, which is what @sec-va provides.
:::

::: {.callout-important .zh}
## 对 N2 结果的诚实表述

在主强度面所选用的带宽下($\sigma_\lambda = `r n_fmt(SO$sigma_main)`$ m),
**观测到的模式与非齐次泊松过程无法区分**
(DCLF $p =$ `r sprintf("%.3f", lad[[which(SO$sig_ladder == SO$sigma_main)]]$test$p.value)`,
MAD $p =$ `r sprintf("%.3f", lad[[which(SO$sig_ladder == SO$sigma_main)]]$mad$p.value)`,
观测曲线在每个距离上都位于 1% 全局包络之内)。

把这说成「没有聚集」是**错误的**。这是关于一个**检验力极低**的检验的陈述,而原因有
两条、都可以指认:

1. **在该带宽下零模型近乎饱和**,如上所述。
2. **逆强度权重是无界的。** $L_{inhom}$ 按 $1/(\hat\lambda_i\hat\lambda_j)$ 对
   点对加权,而在 $\sigma_\lambda = 2$ km 时,数据点处的拟合强度跨度约为
   `r {lamx <- density(ppp_px, sigma = SO$sigma_main, edge = TRUE, positive = TRUE, dimyx = c(512,256))[win, drop = FALSE]; v <- lamx[ppp_px]; sprintf("%.0f", max(v)/min(v))}`
   倍。因此一对落在稀疏北部的模拟点会贡献一个极大的量,使零分布具有非常重的尾部。
   其后果在 @fig-pcf 中可见:$g_{inhom}$ 的全局包络在零值 1 附近大约是 ±200。

出路**不是**更好的带宽,而是一个**参数很少的零模型**。这正是 @sec-va 所提供的。
:::

## <span class="en">Describing the structure, rather than testing it</span><span class="zh">描述这个结构,而不是检验它</span> {#sec-so-pcf}

::: {.en}
Since the N2 test cannot resolve the question, the pair-correlation function is
used here **descriptively**: it says how much more pairing there is at distance
$r$ than the fitted 2 km intensity implies, without asking whether that excess
is significant.
:::

::: {.zh}
既然 N2 检验无法解决这个问题,此处就把对相关函数**当作描述性工具**使用:它说明
在距离 $r$ 上的成对程度比拟合的 2 km 强度所隐含的高出多少,而不追问这一超额是否
显著。
:::

```{r}
#| label: fig-pcf
#| fig-cap: "<span class='en'>Inhomogeneous pair-correlation function relative to the σ_λ = 2 km intensity. The observed curve is shown with the global envelope, which is far too wide to be informative, itself a diagnostic of the weight-explosion problem.</span><span class='zh'>相对 σ_λ = 2 km 强度的非齐次对相关函数。图中同时给出全局包络,但它宽到完全没有信息量。这本身就是权重爆炸问题的诊断。</span>"
#| fig-alt: "Inhomogeneous pair-correlation function relative to the σ_λ = 2 km intensity. The observed curve is shown with the global envelope, which is far too wide to be informative, itself a diagnostic of the weight-explosion problem."
#| fig-width: 9.4
#| fig-height: 4.2

# g(r) is plotted on a log scale, which requires positive values. The r = 0
# endpoint and the lower envelope can both be zero or negative for a
# kernel-smoothed pair-correlation estimate, so those are handled explicitly
# rather than silently dropped by the scale.
gd <- as.data.frame(SO$env_g) |>
  dplyr::filter(r > 0, is.finite(obs), obs > 0) |>
  dplyr::mutate(lo_p = pmax(lo, min(obs) / 3), hi_p = pmax(hi, lo_p * 1.001))

ggplot(gd, aes(r / 1000)) +
  geom_ribbon(aes(ymin = lo_p, ymax = hi_p, fill = "band"), alpha = 0.55) +
  geom_hline(yintercept = 1, linetype = "22", colour = PAL$theo,
             linewidth = 0.5) +
  geom_line(aes(y = obs, colour = "obs"), linewidth = 0.95) +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = -Inf, ymax = Inf,
           fill = "#78716c", alpha = 0.22) +
  annotate("segment", x = 0.5, xend = 0.5, y = -Inf, yend = Inf,
           colour = "#57534e", linewidth = 0.4, linetype = "12") +
  annotate("text", x = 0.58, y = max(gd$hi_p), label = "below sensor resolution",
           hjust = 0, vjust = 1.4, size = 2.9, colour = "#57534e") +
  scale_fill_manual(NULL, values = c(band = PAL$env),
                    labels = "Global envelope (uninformatively wide)") +
  scale_colour_manual(NULL, values = c(obs = PAL$fire), labels = "Observed") +
  scale_y_log10(labels = label_number(accuracy = 0.1)) +
  labs(title = expression(paste("Inhomogeneous pair correlation ", g[inhom](r))),
       subtitle = paste0("σ_λ = ", n_fmt(SO$sigma_main),
                         " m · log scale · dashed line at 1 is the null value"),
       x = "Distance r (km)", y = expression(g[inhom](r))) +
  theme_th(11)
```

```{r}
#| label: tbl-pcf
#| tbl-cap: "<span class='en'>Observed inhomogeneous pair correlation at selected distances, relative to the σ_λ = 2 km intensity. Reported descriptively: the envelope is too wide to support a significance claim.</span><span class='zh'>若干选定距离上观测到的非齐次对相关值(相对 σ_λ = 2 km 强度)。仅作描述性报告:包络太宽,不足以支持显著性论断。(下表为英文原表)</span>"
tibble(`r (m)` = c(500, 750, 1000, 2000, 3000, 5000, 8000)) |>
  mutate(`g_inhom(r)` = approx(gd$r, gd$obs, `r (m)`)$y,
         `Excess pairing` = sprintf("%.1f× the null", `g_inhom(r)`)) |>
  gt::gt() |>
  gt::fmt_number(columns = `g_inhom(r)`, decimals = 2) |>
  gt::fmt_number(columns = `r (m)`, decimals = 0) |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(5))
```

::: {.interp .en}
**Excess pairing at short range.** Relative to the 2 km kernel intensity there is still
substantial excess pairing at short range (`r sprintf("%.1f×", approx(gd$r, gd$obs, 500)$y)` the null at 500 m,
`r sprintf("%.1f×", approx(gd$r, gd$obs, 1000)$y)` at 1 km,
`r sprintf("%.1f×", approx(gd$r, gd$obs, 2000)$y)` at 2 km) decaying towards
but not reaching 1 by 8 km. **Evidence:** the *shape* is a clean monotone decay,
which is what a clustered process produces. **But the envelope spans roughly
±200**, so none of this is statistically distinguishable from the null: the
inverse-intensity weights make the sampling distribution enormously dispersed.
**Implication:** the descriptive scale of the excess is a few kilometres, and
establishing it as real requires the parametric route of @sec-va, which
independently estimates a cluster scale of
`r n_fmt(round(VA$clustpar[["scale"]], -1))` m. *One qualification:* nothing inside the
grey strip may be interpreted at all; and `divisor = "d"` makes $g$ intrinsically
noisy at very small $r$.
:::

::: {.interp .zh}
**短程上的超额成对。** 相对 2 km 的核强度,短距离上仍存在可观的超额成对(500 m 处为
零模型的 `r sprintf("%.1f×", approx(gd$r, gd$obs, 500)$y)`,1 km 处
`r sprintf("%.1f×", approx(gd$r, gd$obs, 1000)$y)`,2 km 处
`r sprintf("%.1f×", approx(gd$r, gd$obs, 2000)$y)`)并向 1 衰减,但到 8 km 仍未
达到 1。**证据:** 其*形状*是干净的单调衰减,而这正是聚集过程所产生的形态。
**但包络跨度约为 ±200**,因此这些都无法在统计上与零模型区分:逆强度权重使抽样分布
极度离散。**含义:** 该超额在描述意义上的尺度是数公里,而要确立它是真实的,需要
@sec-va 的参数化路径,后者独立估计出的簇群尺度为
`r n_fmt(round(VA$clustpar[["scale"]], -1))` m。*一项限定:* 灰色条带内的任何内容都不可
解读;另外 `divisor = "d"` 使 $g$ 在极小 $r$ 处本身就很嘈杂。
:::

## <span class="en">Is the N2 non-result an artefact of our other choices?</span><span class="zh">N2 的「非结果」是其他选择造成的假象吗?</span> {#sec-so-robustness}

```{r}
#| label: tbl-robust
#| tbl-cap: "<span class='en'>Robustness of the N2 result to the definition of a point event and to the study window. All variants agree, including the one that is immune to pixel-level duplication.</span><span class='zh'>N2 结果对「点事件定义」与「研究窗口」的稳健性。所有变体结论一致,包括那个对像元级重复完全免疫的变体。(下表为英文原表)</span>"

t_unit <- bind_rows(lapply(names(SO$unit_check), function(nm) {
  x <- SO$unit_check[[nm]]; s <- env_summary(x$env)
  tibble(Check = "Point-event definition", Variant = nm, n = x$n,
         `DCLF p` = x$test$p.value, `% r outside` = s$pct_out)
}))

t_win <- SO$window_check |>
  transmute(Check = "Study window",
            Variant = ifelse(inner_buffer_km == 0, "full regency",
                             paste0("shrunk by ", inner_buffer_km, " km")),
            n = n, `DCLF p` = dclf_p, `% r outside` = pct_r_outside)

bind_rows(t_unit, t_win) |>
  gt::gt(groupname_col = "Check") |>
  gt::fmt_number(columns = n, decimals = 0) |>
  gt::fmt_number(columns = `DCLF p`, decimals = 4) |>
  gt::fmt_number(columns = `% r outside`, decimals = 1) |>
  gt::tab_options(table.font.size = gt::pct(84), data_row.padding = gt::px(5),
                  row_group.font.weight = "bold")
```

```{r}
#| label: tbl-normpower
#| tbl-cap: "<span class='en'>Sensitivity of the observed L_inhom to the normalisation power, whose spatstat default changed in version 3.8-1.</span><span class='zh'>观测 L_inhom 对归一化幂次的敏感性;该参数的 spatstat 默认值在 3.8-1 版中有过变更。(下表为英文原表)</span>"
SO$np_check |>
  rename(`normpower` = normpower, `L at 500 m` = L_at_500,
         `L at 2 km` = L_at_2000, `L at 10 km` = L_at_10000) |>
  gt::gt() |>
  gt::fmt_number(columns = -normpower, decimals = 0) |>
  gt::tab_options(table.font.size = gt::pct(85), data_row.padding = gt::px(4))
```

::: {.en}
**Point-event definition.** All three units give the same answer at
$\sigma_\lambda = 2$ km, inside the envelope, $p$ between
`r sprintf("%.3f", min(vapply(SO$unit_check, function(x) x$test$p.value, 0)))` and
`r sprintf("%.3f", max(vapply(SO$unit_check, function(x) x$test$p.value, 0)))`.
The **event** unit matters most: it is already de-duplicated in space *and*
time, so it cannot be affected by one fire occupying several pixels. Its
agreement confirms that the N2 non-result is a property of the test, not of
pixel-level duplication.

It should be noted that the event unit is itself imperfect. Single-linkage
chaining merged
`r sprintf("%.0f%%", 100*max(fire_ev$n_px)/sum(fire_ev$n_px))` of all pixel-days
into one component, because at this fire density there is no clean separation
between discrete events, which is itself evidence of a spatially contiguous
burning field.

**Study window.** Shrinking the window by 5 and 10 km removes the boundary zone
altogether, a stronger check on edge effects than any correction formula. The
result is unchanged, so nothing here is manufactured by the translation
correction or by the coastal boundary.

**Normalisation.** `normpower = 1` and `2` change the magnitude of
$L_{inhom}$ but not its shape or sign, so the choice is not load-bearing.
:::

::: {.zh}
**点事件定义。** 三种单元在 $\sigma_\lambda = 2$ km 下给出相同答案,都位于包络
之内,$p$ 介于
`r sprintf("%.3f", min(vapply(SO$unit_check, function(x) x$test$p.value, 0)))` 与
`r sprintf("%.3f", max(vapply(SO$unit_check, function(x) x$test$p.value, 0)))` 之间。
其中**事件**单元最关键:它在空间**和**时间上都已去重,因此不可能受「一场火占据多个
像元」的影响。它的一致性印证了 N2 的「非结果」是**检验本身**的性质,而不是像元级
重复造成的。

需要说明的是,事件单元本身并不完美。单连接链式合并把全部像元-日的
`r sprintf("%.0f%%", 100*max(fire_ev$n_px)/sum(fire_ev$n_px))` 并进了一个连通分量,
因为在本例的火密度下,离散事件之间没有清晰的分界,而这本身就是「火场在空间上连续
成片」的证据。

**研究窗口。** 把窗口内缩 5 km 与 10 km 会完全移除边界带,这是比任何校正公式都更强
的边界效应检查。结果不变,因此此处没有任何结论是由平移校正或海岸边界制造出来的。

**归一化。** `normpower = 1` 与 `2` 改变 $L_{inhom}$ 的**量级**,但不改变其形状或
符号,因此这一选择不承重。
:::

## <span class="en">N3: space–time interaction</span><span class="zh">N3:时空交互</span> {#sec-so-knox}

::: {.rq .en}
**Question tested.** *Are detections that are close in space also close in
time, beyond what their separate spatial and temporal patterns imply?*
**Null model N3:** random labelling, the observed locations and the observed
multiset of dates are both held fixed, and dates are randomly re-assigned to
locations (999 permutations) [@knox1964].
:::

::: {.rq .zh}
**所检验的问题。** *空间上靠近的探测点在时间上是否也靠近,其程度是否超出各自的
空间格局与时间格局所隐含的水平?*
**零模型 N3:** 随机贴标,观测到的位置与观测到的日期多重集**都保持固定**,只把
日期随机重新分配给各位置(999 次置换)[@knox1964]。
:::

::: {.en}
This null has a property none of the others has: **it requires no model for
either marginal pattern**, and therefore no bandwidth and no covariates. The
spatial configuration is taken as observed, the daily totals are taken as
observed, and only their pairing is randomised. That makes it the strongest
instrument available here, both because it sidesteps the scale-separation
problem that crippled N2, and because @sec-fo-temporal showed the daily totals
are contaminated by the observation process, which this test *conditions on*
rather than models.
:::

::: {.zh}
这个零模型有一个其他任何零模型都不具备的性质:**它不需要为任一边际格局建模**,
因此既不需要带宽,也不需要协变量。空间构型按观测取用,逐日总量按观测取用,只有
两者的**配对关系**被随机化。这使它成为此处最有力的工具,既因为它绕开了让 N2 瘫痪
的尺度分离问题,也因为 @sec-fo-temporal 已表明逐日总量受观测过程污染,而本检验是
**以它们为条件**,而不是对它们建模。
:::

```{r}
#| label: fig-knox
#| fig-cap: "<span class='en'>Space-time interaction. Left: observed count of space-time-close pairs relative to the permutation mean, for nine threshold combinations. Right: the excess by exact temporal lag, at two spatial radii.</span><span class='zh'>时空交互。左:九种阈值组合下,时空邻近对的观测数相对置换均值之比。右:两种空间半径下,按精确时间滞后给出的超额。</span>"
#| fig-alt: "Space-time interaction. Left: observed count of space-time-close pairs relative to the permutation mean, for nine threshold combinations. Right: the excess by exact temporal lag, at two spatial radii."
#| fig-width: 10.6
#| fig-height: 4.4

p_k <- SO$knox_res |>
  mutate(space = paste0(space_m / 1000, " km"),
         time = factor(paste0("≤ ", time_d, " day", ifelse(time_d == 1, "", "s")))) |>
  ggplot(aes(time, ratio, fill = space)) +
  geom_hline(yintercept = 1, linetype = "22", colour = PAL$theo) +
  geom_col(position = position_dodge(0.8), width = 0.72) +
  geom_text(aes(label = sprintf("%.2f", ratio)),
            position = position_dodge(0.8), vjust = -0.35, size = 2.6,
            colour = PAL$grey) +
  scale_fill_manual("Spatial threshold",
                    values = c(`1 km` = PAL$fire, `2 km` = PAL$fire_l,
                               `5 km` = PAL$grey_l)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.14))) +
  labs(title = "Observed / expected space-time-close pairs",
       subtitle = "Expected = mean over 999 random re-labellings of the dates",
       x = "Temporal threshold", y = "Ratio") +
  theme_th(10)

p_l <- bind_rows(
  SO$st_by_lag_1km |> mutate(radius = "within 1 km"),
  SO$st_by_lag     |> mutate(radius = "within 5 km")) |>
  ggplot(aes(lag_days, ratio, colour = radius)) +
  geom_hline(yintercept = 1, linetype = "22", colour = PAL$theo) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.1) +
  scale_colour_manual(NULL, values = c(`within 1 km` = PAL$fire,
                                       `within 5 km` = PAL$water)) +
  scale_x_continuous(breaks = 0:5) +
  labs(title = "Excess by temporal lag",
       subtitle = "Ratio of observed to permuted pair counts at each exact lag",
       x = "Temporal lag (days)", y = "Observed / expected") +
  theme_th(10)

p_k + p_l
```

```{r}
#| label: tbl-knox
#| tbl-cap: "<span class='en'>Monte-Carlo Knox test. p is the proportion of 999 date permutations producing at least as many space-time-close pairs as observed.</span><span class='zh'>蒙特卡洛 Knox 检验。p 为 999 次日期置换中,时空邻近对数不少于观测值的比例。(下表为英文原表)</span>"
SO$knox_res |>
  transmute(`Spatial threshold` = paste0(space_m / 1000, " km"),
            `Temporal threshold` = paste0("≤ ", time_d, " d"),
            `Pairs within spatial threshold` = n_pairs_space,
            Observed = observed, `Expected (permuted)` = sim_mean,
            Ratio = ratio, `p` = p_value) |>
  gt::gt() |>
  gt::fmt_number(columns = c(`Pairs within spatial threshold`, Observed,
                             `Expected (permuted)`), decimals = 0) |>
  gt::fmt_number(columns = Ratio, decimals = 2) |>
  gt::fmt_number(columns = p, decimals = 3) |>
  gt::tab_options(table.font.size = gt::pct(83), data_row.padding = gt::px(4))
```

```{r}
#| label: knox-numbers
#| echo: false
k00 <- SO$knox_res |> dplyr::filter(space_m == 1000, time_d == 0)
k_all_sig <- all(SO$knox_res$p_value < 0.05)
k_max <- SO$knox_res |> dplyr::slice_max(ratio, n = 1)
```

::: {.interp .en}
**Nine thresholds, one answer.** All nine threshold combinations reject at the
permutation floor ($p$ `r pv(min(SO$knox_res$p_value))`), and the excess is
sharply graded: within 1 km on the same day the observed count is
`r sprintf("%.2f×", k00$ratio)` the permuted expectation, against
`r sprintf("%.2f×", SO$knox_res$ratio[SO$knox_res$space_m==5000 & SO$knox_res$time_d==0])`
at 5 km. The right panel shows the interaction is *localised*, not merely
positive: at 1 km the ratio falls from
`r sprintf("%.2f", SO$st_by_lag_1km$ratio[1])` at lag 0 through 1 into a
**deficit** by lag 3 (`r sprintf("%.2f", SO$st_by_lag_1km$ratio[4])`). Nearby
pairs are packed into a two-to-three-day window *at the expense of* longer lags.
**Evidence:** the null holds both marginals fixed, and unlike N2 it uses no
bandwidth and no covariates. **Implication:** where and when are genuinely
dependent, fire predicts fire nearby for about two days. *Read with care:* this
includes a moving front and multi-day smouldering, not only contagion; the
lag-0 excess involves distinct locations, so not smouldering alone.
:::

::: {.interp .zh}
**九组阈值,一个答案。** 九种阈值组合全部在置换下限上被拒绝
($p$ `r pv(min(SO$knox_res$p_value))`),而超额呈明显梯度:1 km 内同日的观测数为
置换期望的 `r sprintf("%.2f×", k00$ratio)`,而 5 km 处为
`r sprintf("%.2f×", SO$knox_res$ratio[SO$knox_res$space_m==5000 & SO$knox_res$time_d==0])`。
右图显示这一交互是*局域的*,而不仅仅是正的:在 1 km 上,比值从滞后 0 的
`r sprintf("%.2f", SO$st_by_lag_1km$ratio[1])` 穿过 1,到滞后 3 时转为**亏缺**
(`r sprintf("%.2f", SO$st_by_lag_1km$ratio[4])`)。邻近点对被压缩进一个两到三天的
窗口内,而这是**以更长滞后为代价**的。**证据:** 该零模型固定了两个边际,而且与
N2 不同,它不使用带宽、也不使用协变量。**含义:** 「哪里」与「何时」确实相互依赖
,火会在约两天内预示附近再有火。*阅读时请留意:* 这里面既包含推进的火线,也包含同一片地面
的跨日阴燃,而不只是传染;但滞后 0 的超额涉及的是**不同位置**,因此并非仅由阴燃
造成。
:::

## <span class="en">Where this leaves the second-order question</span><span class="zh">二阶问题最终停在哪里</span>

```{r}
#| label: tbl-so-summary
#| tbl-cap: "<span class='en'>The three second-order reference models and what each established.</span><span class='zh'>三个二阶参照模型及各自确立了什么。(下表为英文原表)</span>"
tibble::tribble(
  ~Null, ~Result, ~`What it establishes`,
  "**N1** CSR",
  sprintf("rejected, p %s", pv(SO$test_L_csr$p.value)),
  "Nothing usable, rejection was guaranteed by the first-order result.",

  "**N2** inhomogeneous Poisson, kernel intensity",
  sprintf("not rejected at σ_λ = 1–2 km (p %s at 2 km); rejected at σ_λ = 10 km (p %s)",
          sprintf("%.3f", lad[[2]]$test$p.value),
          sprintf("%.4f", lad[[4]]$test$p.value)),
  "That the spatial question is **not identifiable** from these data: the answer is set by the null's flexibility, not by the pattern.",

  "**N3** random date labelling",
  sprintf("rejected at all nine thresholds, strongest %.2f× at %s km / ≤%s d",
          k_max$ratio, k_max$space_m/1000, k_max$time_d),
  "Genuine space–time interaction, with no bandwidth, no covariates and no model for either marginal."
) |>
  gt::gt() |> gt::fmt_markdown(columns = everything()) |>
  gt::cols_width(Null ~ gt::pct(22), Result ~ gt::pct(34),
                 `What it establishes` ~ gt::pct(44)) |>
  gt::tab_options(table.font.size = gt::pct(83), data_row.padding = gt::px(7))
```

::: {.en}
The spatial second-order question is left open by N2 *by construction*. It is
resolved in @sec-va by replacing the kernel null with a parametric one.
:::

::: {.zh}
空间二阶问题被 N2 *在构造上*留成了开放问题。它在 @sec-va 中通过把核零模型替换为
参数化零模型而得到解决。
:::
