# <span class="en">Reproducibility appendix</span><span class="zh">可复现性附录</span> {#sec-repro}

## <span class="en">How to reproduce this report</span><span class="zh">如何复现本报告</span>

::: {.en}
The repository is a self-contained Quarto project. From a clean clone:
:::

::: {.zh}
本仓库是一个自包含的 Quarto 项目。从一份干净的克隆开始:
:::

```bash
# 1. Acquire the data (choose ONE route)
#    Route A - open 7-day FIRMS files, no credentials needed. Note that this
#    returns whatever the most recent seven days happen to be, so re-running it
#    today will NOT reproduce the September 2026 window analysed here.
Rscript scripts/01b_assemble_public7d.R

#    Route B - authenticated FIRMS Area API, arbitrary date range. This is the
#    route that reproduces (or extends) the exact window.
export FIRMS_MAP_KEY=<your key>          # or write it to .firms_key
Rscript scripts/01_download_firms.R 2026-09-04 2026-09-11

# 2. Boundaries and contextual layers
Rscript scripts/02_osm_context.R

# 3. Preparation and analysis
Rscript scripts/03_prepare.R
Rscript scripts/04_covariates.R
Rscript scripts/05_firstorder.R
Rscript scripts/05b_stkde.R
Rscript scripts/06_secondorder.R
Rscript scripts/07_valueadded.R

# 4. Render the report and the executive summary
quarto render
```

::: {.callout-note .en}
## A note on exact reproducibility of the data
The open 7-day FIRMS files are a *rolling window*: they always contain the last
seven days. The raw CSV actually used here is committed to the repository
(`data/raw/firms_pulangpisau_2026-09-04_2026-09-11.csv`), so steps 2–4 reproduce
every number in this report exactly. Step 1 is documented so that the
acquisition is transparent and so that the analysis can be re-run over a
different or longer window, not because re-running it would return the same
records.
:::

::: {.callout-note .zh}
## 关于数据可精确复现性的说明

开放的 7 日 FIRMS 文件是一个*滚动窗口*:它始终只包含最近七天。本研究实际使用的原始
CSV 已提交进仓库(`data/raw/firms_pulangpisau_2026-09-04_2026-09-11.csv`),因此
步骤 2–4 可以**精确**复现本报告中的每一个数字。步骤 1 之所以写出来,是为了让数据
获取过程透明、并让分析能在不同或更长的窗口上重跑,而**不是**因为重跑它会返回相同的
记录。
:::

## <span class="en">Pipeline structure</span><span class="zh">管线结构</span>

::: {.en}
Rectangles are scripts; rounded boxes are data artefacts.
:::

::: {.zh}
矩形为脚本,圆角框为数据产物。
:::

```{mermaid}
%%| label: fig-pipeline
flowchart TD
  A1["01 / 01b<br/>FIRMS acquisition"] --> D1(["data/raw/<br/>firms_*.csv"])
  A2["02_osm_context.R"] --> D2(["data/raw/osm/"])
  B0(["COD-AB ADM1-3<br/>(HDX / BPS)"]) --> P1
  D1 --> P1["03_prepare.R<br/><i>QA, CRS, units</i>"]
  P1 --> D3(["ppp_det · ppp_pxday · ppp_event<br/>win · fire_sf · qa_audit"])
  D2 --> C1["04_covariates.R"]
  B0 --> C1
  C1 --> D4(["covars.rds<br/>d_canal · d_river · d_road · d_settle"])
  D3 --> F1["05_firstorder.R<br/><i>KDE, quadrat, rhohat</i>"]
  D4 --> F1
  D3 --> F2["05b_stkde.R<br/><i>daily + ST kernels</i>"]
  D3 --> S1["06_secondorder.R<br/><i>G/F/J, L, Linhom, Knox</i>"]
  D3 --> V1["07_valueadded.R<br/><i>ppm, kppm</i>"]
  D4 --> V1
  F1 --> R(["index.qmd<br/>slides.qmd"])
  F2 --> R
  S1 --> R
  V1 --> R
```

## <span class="en">Caching policy</span><span class="zh">缓存策略</span>

::: {.en}
Monte-Carlo envelopes dominate the runtime of this analysis. They are wrapped in
a small helper so that the **code that ran stays visible in the report** while
repeated renders do not repeat the computation:
:::

::: {.zh}
蒙特卡洛包络占据了本分析的绝大部分运行时间。它们被包在一个小助手函数里,使得
**实际运行过的代码仍然显示在报告中**,同时重复渲染不会重复计算:
:::

```{r}
#| label: cache-helper-shown
#| eval: false
#| file: R/utils.R
```

::: {.en}
Deleting `data/cache/` forces a full recomputation; because
`set.seed(1234)` is set once at the top of every script, the recomputed values
are identical.
:::

::: {.zh}
删除 `data/cache/` 会强制完全重算;由于每个脚本开头都设定了一次
`set.seed(1234)`,重算得到的数值完全相同。
:::

## <span class="en">Session information</span><span class="zh">会话信息</span>

```{r}
#| label: sessioninfo
#| code-fold: true
#| code-summary: "Full session information"

si <- sessionInfo()
pkgs <- c("sf", "terra", "spatstat", "spatstat.geom", "spatstat.explore",
          "spatstat.model", "spatstat.random", "sparr", "dplyr", "ggplot2",
          "tmap", "gt", "patchwork", "lubridate", "dbscan", "igraph",
          "osmdata", "quarto", "knitr")
tibble::tibble(
  Package = pkgs,
  Version = vapply(pkgs, function(p)
    tryCatch(as.character(utils::packageVersion(p)),
             error = function(e) "not installed"), character(1))
) |>
  gt::gt() |>
  gt::tab_header(title = gt::md(paste0("**", si$R.version$version.string, "** on ",
                                       si$running))) |>
  gt::tab_source_note(gt::md(paste0(
    "GEOS ", sf::sf_extSoftVersion()[["GEOS"]],
    " &middot; GDAL ", sf::sf_extSoftVersion()[["GDAL"]],
    " &middot; PROJ ", sf::sf_extSoftVersion()[["PROJ"]]))) |>
  gt::tab_options(table.font.size = gt::pct(82), data_row.padding = gt::px(3),
                  container.height = gt::px(420))
```

## <span class="en">Analysis scripts</span><span class="zh">分析脚本</span> {#sec-repro-scripts}

::: {.en}
The complete, unedited source of each script, inserted from disk at render time
so that it cannot drift from what actually ran.
:::

::: {.zh}
每个脚本的完整、未经编辑的源码,在渲染时从磁盘插入,因此不可能与实际运行的版本
产生偏移。
:::

::: {.panel-tabset}

### 01 acquisition (API)

```{r}
#| eval: false
#| file: scripts/01_download_firms.R
```

### 01b acquisition (open)

```{r}
#| eval: false
#| file: scripts/01b_assemble_public7d.R
```

### 02 OSM context

```{r}
#| eval: false
#| file: scripts/02_osm_context.R
```

### 03 preparation

```{r}
#| eval: false
#| file: scripts/03_prepare.R
```

### 04 covariates

```{r}
#| eval: false
#| file: scripts/04_covariates.R
```

### 05 first order

```{r}
#| eval: false
#| file: scripts/05_firstorder.R
```

### 05b spatio-temporal

```{r}
#| eval: false
#| file: scripts/05b_stkde.R
```

### 06 second order

```{r}
#| eval: false
#| file: scripts/06_secondorder.R
```

### 07 value added

```{r}
#| eval: false
#| file: scripts/07_valueadded.R
```

:::
