---
title: "Hands-on Exercise 5a: Global Measures of Spatial Autocorrelation"
author: "Ji Guofang"
date: "last-modified"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

A choropleth map can look clustered without being clustered. The eye is a poor judge of randomness. It finds pattern in noise, and it is easily led by whichever class breaks the cartographer happened to choose. Whether similar values really do sit next to one another more often than chance allows is a statistical question, and the statistics that answer it are the global measures of spatial autocorrelation.

This exercise computes those measures for the 88 counties of Hunan province, using GDP per capita in 2012 as the variable of interest and the **spdep** package throughout. Three instruments are applied to the same data. Moran's I is a cross-product statistic that behaves like a correlation coefficient between each value and the average of its neighbours. Geary's C is a squared-difference statistic that compares the dissimilarity of neighbouring pairs with the dissimilarity of all pairs. The spatial correlogram recomputes either statistic at successive orders of contiguity, and so shows how far the dependence reaches.

Each of the first two is evaluated twice, once against its analytical sampling distribution and once against a Monte Carlo permutation distribution. The two routes rest on different assumptions, so agreement between them is informative in itself. By the end of the exercise the following are in place:

- geospatial data imported with the appropriate functions of **sf**,
- a csv file imported with the appropriate function of **readr**,
- a relational join performed with the appropriate function of **dplyr**,
- global spatial autocorrelation statistics computed with the appropriate functions of **spdep**, and
- a statistically defensible reading of what those statistics say.

The local counterpart of these measures, which asks not whether the province clusters but *where*, is the subject of [Hands-on Exercise 5b](hand-on-ex5b.qmd).

## Getting Started

### The analytical question

Equal distribution of development across a province is a standard objective of local government and planning. The analytical task that follows from it is a chain of three questions, each asked only if the previous one has been answered:

1.  Is development evenly distributed geographically? If the answer is **no**, then
2.  is there a sign of spatial clustering? And if the answer to that is **yes**, then
3.  where are those clusters?

This exercise settles the first two for GDP per capita in Hunan province, People's Republic of China. The third is left to [Exercise 5b](hand-on-ex5b.qmd). A global statistic is by construction a single number for the whole map, so it cannot locate anything.

### The Study Area and Data

Two data sets are used:

| Dataset | Description |
|----|----|
| `Hunan` | County boundary layer of Hunan province, China, in ESRI shapefile format |
| `Hunan_2012.csv` | Selected local development indicators for Hunan in 2012 |

These are the same two files used in [Hands-on Exercise 4](../hand-on-ex4/hand-on-ex4.qmd), carried over unchanged, so the neighbour structure built below is identical to the Queen contiguity list constructed there.

### Setting the Analytical Tools

Four packages are used:

| Package | Role |
|----|----|
| **sf** | Importing and handling geospatial data |
| **spdep** | Spatial weights, and the global autocorrelation statistics and their permutation tests |
| **tmap** | Cartographic quality thematic maps |
| **tidyverse** | Data import, transformation and wrangling |

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

## Getting the Data Into R Environment

### Import shapefile into R environment

`st_read()` of **sf** imports the Hunan shapefile as a simple feature object:

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

The header reports 88 polygon features with seven attribute fields, in geographic coordinates on WGS 84. The coordinates are deliberately left unprojected. Every statistic in this document is computed from a contiguity relation, and whether two counties share a boundary does not depend on the coordinate system. [Exercise 5b](hand-on-ex5b.qmd) does project the layer, because the distance bands used there do depend on it.

### Import csv file into R environment

`read_csv()` of **readr** imports the indicator table. The output is a plain data frame with no geometry:

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### Performing relational join

`left_join()` of **dplyr** attaches the indicator fields to the attribute table of the `hunan` simple feature object, and `select()` keeps the identifying fields and GDPPC:

```{r}
hunan <- left_join(hunan, hunan2012, by = "County") %>%
  select(1:4, 7, 15)
```

The join key is written out rather than left to automatic matching, and `hunan` is named first so that the result keeps its geometry column and remains an `sf` object.

### Visualising Regional Development Indicator

Before any test is run, the variable is mapped. Two choropleths of the same column are drawn, one classified by equal interval and one by quantile, because the pair shows that the first question above cannot be settled by looking:

```{r}
#| fig-width: 12
#| fig-height: 6
equal <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_polygons(fill = "GDPPC",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "GDPPC",
                position = tm_pos_in("left", "bottom"))) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile classification")

tmap_arrange(equal, quantile, asp = 1, ncol = 2)
```

The two maps are drawn from identical numbers and tell different stories. GDPPC in Hunan ranges from 8,497 to 88,656 with a median of 20,433, so the distribution is strongly right-skewed. Under equal interval that skew puts 58 of the 88 counties into the lowest class and only nine into the top three, so most of the province is drawn in one pale colour and the map reads as broadly uniform. Under quantile the same data fill all five classes by construction, and a dark north-eastern corner appears around Changsha.

Neither map is wrong, and that is the difficulty. A classification choice has changed the apparent answer to question 1, which is why a statistic that does not depend on class breaks is needed.

## Global Measures of Spatial Autocorrelation

### Computing Contiguity Spatial Weights

Every measure below is a weighted sum over pairs of counties, so the neighbour relation has to be fixed before anything can be computed. `poly2nb()` of **spdep** builds a neighbours list from regions with contiguous boundaries. Its `queen` argument defaults to `TRUE`, which returns first order neighbours under the Queen criterion, where a single shared point is enough to make two counties neighbours.

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

There are 88 area units, 448 non-zero links, and an average of 5.09 neighbours per county. The most connected unit has 11 neighbours, and two units have one neighbour each. No unit is isolated, which matters here: a county with no neighbours contributes nothing to the cross-products that make up these statistics, and has to be handled explicitly.

### Row-standardised weights matrix

The neighbour list records *who* is a neighbour. The weights matrix records *how much* each neighbour counts, and `nb2listw()` converts one into the other. With `style = "W"` each of a county's neighbours receives the fraction 1/(number of neighbours), so the weights in every row sum to one and the spatial lag of a county is the plain average of its neighbours' values.

```{r}
rswm_q <- nb2listw(wm_q,
                   style = "W",
                   zero.policy = TRUE)
rswm_q
```

::: callout-note
## What the arguments do

The input to `nb2listw()` must be an object of class `nb`. Two arguments carry most of the meaning:

-   **`style`** takes `"W"`, `"B"`, `"C"`, `"U"`, `"minmax"` or `"S"`. `B` is basic binary coding; `W` is row standardised, summing over all links to *n*; `C` is globally standardised, also summing to *n*; `U` is `C` divided by the number of neighbours, summing to unity; and `S` is the variance-stabilising scheme of Tiefelsdorf et al. (1999, pp. 167-168).
-   **`zero.policy = TRUE`** inserts a zero-length weights vector for any region with no neighbours, which produces a lag value of zero for that region. Whether zero is a sensible lag for an isolated unit is a substantive question, not a technical one. No unit is isolated here, so the argument never takes effect, and is set only so that the code does not depend on that remaining true.
:::

The printed weights constants are the quantities the analytical tests are built from: *S~0~* = 88, the sum of all weights, which equals *n* exactly because each row sums to one; *S~1~* = 37.86; and *S~2~* = 365.91. Row standardisation carries one well-known cost. Counties on the provincial edge have fewer neighbours, so each of the neighbours they do have takes a larger weight, and their lagged values rest on a thinner base than interior counties'. `style = "B"` is the more robust alternative, and `"W"` is kept here for interpretability.

## Global Measures of Spatial Autocorrelation: Moran's I

Moran's I compares each value's deviation from the provincial mean with the deviations of its neighbours. It is bounded approximately by −1 and +1. Positive values mean that similar values sit together, negative values mean that high and low values alternate, and the expectation under no spatial pattern is not zero but −1/(*n*−1), here −0.0115.

### Moran's I test

```{r}
moran.test(hunan$GDPPC,
           listw = rswm_q,
           zero.policy = TRUE,
           na.action = na.omit)
```

The observed I is 0.3007, against an expectation of −0.0115 and a variance of 0.004348 under the randomisation hypothesis. The standard deviate is 4.7351 and the one-sided p-value is 1.095 × 10^-6^, so the null hypothesis of spatial randomness is rejected at any conventional level. The sign is positive, which means the clustering is of like with like: counties with high GDPPC tend to border counties with high GDPPC, and the same holds at the low end. Development in Hunan is therefore **not** evenly distributed, and there **is** a sign of spatial clustering, which answers the first two of the three questions above.

The magnitude carries as much information as the p-value. An I of 0.30 is moderate rather than overwhelming. Neighbourhood explains a real but partial share of the variation in county income, and a great deal of that variation remains local to individual counties.

### Computing Monte Carlo Moran's I

The test above draws its p-value from an analytical approximation to the sampling distribution of I. That approximation is asymptotic and assumes a normality or randomisation null. A permutation test avoids the assumption entirely: `moran.mc()` reassigns the 88 observed GDPPC values to the 88 counties at random, recomputes I, and repeats, building the null distribution from the data rather than from theory. With `nsim = 999` the observed value is compared against 999 simulations, for 1000 values in total.

```{r}
set.seed(1234)
bperm <- moran.mc(hunan$GDPPC,
                  listw = rswm_q,
                  nsim = 999,
                  zero.policy = TRUE,
                  na.action = na.omit)
bperm
```

`set.seed()` is called first so that the permutations, and therefore the printed p-value, are reproducible across renders.

The observed I of 0.30075 has rank 1000 of 1000, meaning no permutation of the data produced a value as high. The pseudo p-value is 1/1000 = 0.001, the smallest value 999 simulations can deliver. That figure is a floor rather than a measurement: the true p-value is smaller than 0.001 by an unknown amount, and the analytical test's 1.1 × 10^-6^ is the better estimate of how much smaller.

The conclusion matches the analytical one, which is the point of running both. Had they disagreed, the randomisation-based p-value would have been the one to trust, since it makes no distributional assumption.

### Visualising Monte Carlo Moran's I

The simulated statistics are examined directly, rather than only through the p-value derived from them. The mean and variance of the 999 simulated values come first:

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

The simulated mean is −0.01505, against the theoretical expectation of −0.01149, and the simulated variance is 0.004372 against the theoretical 0.004348. The permutation distribution has reproduced the analytical moments to three decimal places. That is empirical confirmation that the assumptions behind `moran.test()` hold for this data set.

The distribution is then plotted with `hist()` and `abline()` of R Graphics:

```{r}
#| fig-width: 8
#| fig-height: 5
hist(bperm$res,
     freq = TRUE,
     breaks = 20,
     xlab = "Simulated Moran's I")
abline(v = 0,
       col = "red")
```

The simulated values form a roughly symmetric distribution centred just below zero and running from −0.183 to 0.276. The observed I of 0.30075 is the single bar at the far right, outside the range of all 999 simulations, which is what rank 1000 means.

The red line is drawn at zero rather than at the expectation of −0.0115. On this scale the difference is invisible, but it matters in principle. The null value of Moran's I for a finite sample is slightly negative, so treating zero as the benchmark introduces a small conservative bias into any visual reading.

The base graphics call answers the question directly. The same distribution drawn with **ggplot2** carries both reference values explicitly, which makes the comparison easier to read:

```{r}
#| fig-width: 8
#| fig-height: 5
ggplot(data.frame(sim = bperm$res[1:999]),
       aes(x = sim)) +
  geom_histogram(bins = 20,
                 fill = "grey75",
                 colour = "white") +
  geom_vline(xintercept = -1 / (length(hunan$GDPPC) - 1),
             colour = "grey30",
             linetype = "dashed") +
  geom_vline(xintercept = bperm$statistic,
             colour = "red",
             linewidth = 0.8) +
  annotate("text",
           x = bperm$statistic,
           y = Inf,
           vjust = 1.8,
           hjust = 1.1,
           label = "observed I = 0.301",
           colour = "red",
           size = 3.5) +
  labs(x = "Simulated Moran's I",
       y = "Frequency",
       title = "Permutation distribution of Moran's I",
       subtitle = "999 simulations; dashed line is the expectation -1/(n-1)") +
  theme_minimal()
```

The gap between the red line and the right-hand tail of the grey distribution is the visual form of the test result. The observed statistic is not merely in the tail; it is beyond it.

## Global Measures of Spatial Autocorrelation: Geary's C

Moran's I is built from cross-products of deviations from the mean, so it responds to whether neighbouring counties fall on the same side of the provincial average. Geary's C is built from the squared differences between neighbouring values directly, so it responds to how far apart neighbours are in absolute terms. Geary's C is therefore the more sensitive to local differences and Moran's I the more sensitive to global structure, and the two need not agree.

The scale is also inverted. Geary's C has an expectation of exactly 1 under spatial randomness. Values **below** 1 indicate positive spatial autocorrelation, and values above 1 indicate negative autocorrelation.

### Geary's C test

```{r}
geary.test(hunan$GDPPC, listw = rswm_q)
```

The observed C is 0.6907, against an expectation of 1 and a variance of 0.007336. The standard deviate is 3.6108 and the p-value is 1.526 × 10^-4^. The alternative hypothesis is printed as *Expectation greater than statistic*, which is the direction corresponding to positive autocorrelation on this inverted scale. C is significantly below 1, so neighbouring counties differ from one another by less than randomly paired counties do. That is the conclusion Moran's I reached, obtained here from a different functional form.

The agreement is a genuine finding rather than a formality. Moran's I and Geary's C are inversely related but not equivalent, and a data set in which a few extreme local contrasts sit inside an otherwise smooth field can produce a significant I with an unremarkable C. Hunan does not: the clustering is visible both to a statistic that looks at the mean and to one that looks at differences.

### Computing Monte Carlo Geary's C

As with Moran's I, the analytical result is checked against a permutation distribution, this time with `geary.mc()`:

```{r}
set.seed(1234)
bperm <- geary.mc(hunan$GDPPC,
                  listw = rswm_q,
                  nsim = 999)
bperm
```

The observed C of 0.69072 has rank 1 of 1000, and the pseudo p-value is again 0.001. The two statements are consistent. On the Geary scale the evidence of clustering lies in the lower tail, so being the smallest of the 1000 values is the strongest possible result.

This is worth spelling out, because the printed line `alternative hypothesis: greater` looks at first as though the wrong tail is being tested. Inside `geary.mc()` the p-value is computed from `nsim - rank`, so it falls as the observed rank falls. Inside `moran.mc()` it is computed so that it falls as the observed rank rises. The two functions carry the same label and read opposite ends of their respective distributions, which is correct, since a low C and a high I are the same substantive finding.

### Visualising the Monte Carlo Geary's C

The moments of the simulated distribution are computed as before:

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

The simulated mean is 1.0044 against a theoretical expectation of exactly 1, and the simulated variance is 0.007436 against the theoretical 0.007336. Again the match is close, and again it is a reason to trust the analytical test.

```{r}
#| fig-width: 8
#| fig-height: 5
hist(bperm$res, freq = TRUE, breaks = 20, xlab = "Simulated Geary c")
abline(v = 1, col = "red")
```

The simulated values run from 0.714 to 1.272 and centre tightly on 1, where the red reference line is drawn. Here the line sits at the true expectation rather than near it, because for Geary's C the expectation is exactly 1 regardless of *n*.

The observed C of 0.69072 falls just outside the left edge of the simulated range. Compared with the Moran's I histogram the margin is narrower: the observed value is barely beyond the most extreme simulation rather than well beyond it. That is the same information the two standard deviates carry, 3.61 for Geary's C against 4.74 for Moran's I. Both reject the null, but Moran's I rejects it more emphatically, which is the expected ordering when the clustering is driven by a compact cluster of very high values rather than by a uniform smoothness across the whole map.

## Spatial Correlogram

Both statistics so far have used one neighbour definition, namely counties that touch. But spatial dependence does not stop at a shared boundary, and its reach is itself a property worth measuring. A spatial correlogram recomputes a global statistic at increasing orders of contiguity: lag 1 is counties that touch, lag 2 is counties two boundaries away, and so on. Plotting the statistic against lag shows how quickly dependence decays, and at what distance it disappears or reverses.

Correlograms are not as fundamental as variograms, the keystone concept of geostatistics, but for exploratory and descriptive work they provide richer information, since each point comes with its own significance test.

### Compute Moran's I correlogram

`sp.correlogram()` of **spdep** computes a six-lag correlogram of GDPPC using Moran's I, and `plot()` of base graphics draws it:

```{r}
#| fig-width: 8
#| fig-height: 5
MI_corr <- sp.correlogram(wm_q,
                          hunan$GDPPC,
                          order = 6,
                          method = "I",
                          style = "W")
plot(MI_corr)
```

The plot alone does not support a complete interpretation, because not every plotted value is statistically significant. The full analysis report is needed:

```{r}
print(MI_corr)
```

The six estimates are 0.3008, 0.2060, 0.0668, 0.0299, −0.1530 and −0.1187, and the significance codes separate them into three regimes.

Lags 1 and 2 are strongly positive and highly significant, at p = 2.19 × 10^-6^ and 2.03 × 10^-6^. Dependence is not confined to counties that share a boundary. It survives a second step outward at two-thirds of its original strength.

Lags 3 and 4 fade to nothing. Lag 3 is 0.0668 with p = 0.040, statistically significant at 5% but substantively negligible, and lag 4 is 0.0299 with p = 0.226, which is indistinguishable from the null. Positive dependence has run out somewhere between the second and third ring of neighbours.

Lags 5 and 6 turn significantly negative, at −0.1530 with p = 5.98 × 10^-5^ and −0.1187 with p = 0.0089. This is neither noise nor a failure of the method. At five or six steps from any given county one has crossed most of the province, so the comparison being made is essentially between the prosperous north-east and the poorer west and south. Distant pairs are systematically dissimilar, which is the signature of a single large gradient across the study area rather than of many small clusters.

Read together, the three regimes give a description that no single statistic could. Hunan's development has a local neighbourhood effect reaching about two counties, embedded in a province-wide north-east to south-west gradient.

One caution attaches to the p-values in this table. Each is computed for its own lag with no correction for the fact that six tests are being read together, and the estimates are not independent of one another, since the lag-2 neighbours of a county are defined relative to its lag-1 neighbours. The pattern of the sequence is more informative than any individual p-value in it.

### Compute Geary's C correlogram and plot

The same call with `method = "C"` produces the Geary's C correlogram:

```{r}
#| fig-width: 8
#| fig-height: 5
GC_corr <- sp.correlogram(wm_q,
                          hunan$GDPPC,
                          order = 6,
                          method = "C",
                          style = "W")
plot(GC_corr)
```

```{r}
print(GC_corr)
```

The Geary estimates are 0.6907, 0.7630, 0.9397, 1.0098, 1.2008 and 1.0773. On the inverted scale they tell the same story as the Moran sequence, mirrored about 1.

Lags 1 and 2 are well below 1 and highly significant, at p = 3.05 × 10^-4^ and 7.22 × 10^-4^: neighbours, and neighbours of neighbours, are more alike than random pairs. Lags 3 and 4 sit at 0.94 and 1.01 with p = 0.389 and 0.876, indistinguishable from the null. Lag 5 rises to 1.2008 with p = 7.59 × 10^-4^, confirming the significant negative dependence the Moran correlogram found at the same lag.

The two sequences differ in one place. At lag 6 Geary's C is 1.0773 with p = 0.310, which is not significant, where Moran's I was −0.1187 with p = 0.0089. The disagreement follows from what each statistic measures. At six steps the pairs being compared are the far corners of the province, which fall on opposite sides of the provincial mean, and Moran's I registers that reliably because it works from deviations about the mean. Geary's C works from the squared differences between the pairs themselves, and at that range those differences are large but variable, so the same structure produces a weaker signal.

Where the two correlograms agree, at lags 1, 2 and 5, the finding can be treated as a property of the data rather than of the statistic chosen to measure it.

## Reference

-   Kam, T. S. *R for Geospatial Data Science and Analytics*, [Chapter 9: Global Measures of Spatial Autocorrelation](https://r4gdsa.netlify.app/chap09).
-   Anselin, L. (1995) ["Local Indicators of Spatial Association - LISA"](https://onlinelibrary.wiley.com/doi/10.1111/j.1538-4632.1995.tb00338.x), *Geographical Analysis*, Vol. 27, No. 2, pp. 93-115.
-   Bivand, R. S. and Wong, D. W. S. (2018) ["Comparing implementations of global and local indicators of spatial association"](https://link.springer.com/article/10.1007/s11749-018-0599-x), *TEST*, Vol. 27, pp. 716-748.
-   Tiefelsdorf, M., Griffith, D. A. and Boots, B. (1999) "A variance-stabilizing coding scheme for spatial link matrices", *Environment and Planning A*, Vol. 31, pp. 165-180.
-   [spdep: Spatial Dependence - Weighting Schemes, Statistics](https://cran.r-project.org/web/packages/spdep/index.html)
