---
title: "Hands-on Exercise 3: Spatio-Temporal Point Patterns Analysis"
author: "Ji Guofang"
date: "last-modified"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

A spatio-temporal point process — also called a space-time or spatial-temporal point process — is a random collection of points where each point represents the **time and location** of an event. Examples include the incidence of disease, sightings or births of a species, and the occurrence of fires, earthquakes, lightning strikes, tsunamis or volcanic eruptions.

The analysis of spatio-temporal point patterns is becoming increasingly necessary given the rapid emergence of geographically and temporally indexed data. This exercise combines several R packages to run a set of spatio-temporal point pattern analyses, using real-world forest fire events in Kepulauan Bangka Belitung, Indonesia from 1 January 2023 to 31 December 2023.

## Learning Outcome

### The research questions

The specific questions to be answered are:

- Are the locations of forest fire in Kepulauan Bangka Belitung spatially and spatio-temporally independent?
- If the answer is NO, where and when do the observed forest fire locations tend to cluster?

## The Data

Two data sets are used:

| Dataset | Description | Source |
|----|----|----|
| `forestfires.csv` | Locations of forest fires detected by the Moderate Resolution Imaging Spectroradiometer (MODIS) sensor. Only fires within Kepulauan Bangka Belitung are used. | [FIRMS](https://firms.modaps.eosdis.nasa.gov/) |
| `Kepulauan_Bangka_Belitung` | ESRI shapefile of the sub-district (kelurahan) boundaries of Kepulauan Bangka Belitung | Indonesia Geospatial portal |

::: callout-note
## How this copy of the data differs from the textbook's

The textbook's files come from the Indonesia Geospatial portal and FIRMS' authenticated archive. Neither is scriptable without an account, so the same two layers were rebuilt from open sources: the boundaries from **GADM 4.1** at level 4, which is the kelurahan/desa level, and the fires from the **FIRMS country archive** for Indonesia 2023, clipped to the province.

The result is close but not identical, and the difference is worth stating:

| | This exercise | Textbook |
|----|----|----|
| Boundary polygons | 362 | 297 |
| Fire detections | 898 | 741 |
| Eastern extent | 108.85 E | 106.85 E |

The extent is the substantive difference. Belitung island lies between 107.5 E and 108.3 E, so the textbook's boundary — despite the name Kepulauan Bangka Belitung — stops short of it and covers Bangka alone. Clipping to Bangka only gives 766 fires, which is much nearer the textbook's 741; the remaining 132 fires are on Belitung.

The whole province is used here, since that is what the dataset is called and what the research question asks about. Every figure below therefore differs from the textbook's, and the selected bandwidths differ with them.
:::

## Installing and Loading the R Packages

Six packages are used:

| Package | Role |
|----|----|
| **sf** | Importing, processing and wrangling geospatial data |
| **raster** | Handling raster data in R |
| **spatstat** | Spatial point patterns analysis |
| **sparr** | Fixed and adaptive kernel-smoothed spatial relative risk surfaces, and fixed-bandwidth spatio-temporal density estimation |
| **tmap** | Cartographic quality thematic maps |
| **tidyverse** | Data import, transformation, wrangling and visualisation |

**stpp** is added to the list for the space-time K-function in the final section.

```{r}
pacman::p_load(sf, raster, spatstat, sparr, stpp, tmap, tidyverse)
```

::: callout-note
**raster** appears in the textbook's package list but no raster object is created anywhere in this chapter — the density surfaces stay as `sparr` objects and are drawn with base `plot()`. It is loaded here for consistency with the textbook, not because it is needed.
:::

## Importing and Preparing Study Area

### Importing study area

The shapefile is imported, dissolved into a single outline with `st_union()`, stripped of its Z dimension, and reprojected to **WGS 84 / UTM zone 48S** (EPSG: 32748), the projected system appropriate for this part of Indonesia:

```{r}
kbb_sf <- st_read(dsn = "data/rawdata",
                  layer = "Kepulauan_Bangka_Belitung") %>%
  st_union() %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 32748)
```

`st_union()` matters here. The observation window for a point pattern must be a single region, not a collection of several hundred sub-district polygons, so the boundaries are dissolved before conversion.

### Converting to owin

`as.owin()` converts `kbb_sf` into an `owin` object:

```{r}
kbb_owin <- as.owin(kbb_sf)
kbb_owin
```

`class()` confirms the output is indeed an `owin` object:

```{r}
class(kbb_owin)
```

## Importing and Preparing Forest Fire Data

The CSV carries longitude and latitude as ordinary columns, so `st_as_sf()` is used to build the geometry, declaring the coordinates as wgs84 before reprojecting to match the study area:

```{r}
fire_sf <- read_csv("data/rawdata/forestfires.csv") %>%
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326) %>%
  st_transform(crs = 32748)
```

A `ppp` object accepts only numeric or character values as marks, so the acquisition date is converted into three separate numeric and factor fields:

```{r}
fire_sf <- fire_sf %>%
  mutate(DayofYear = yday(acq_date)) %>%
  mutate(Month_num = month(acq_date)) %>%
  mutate(Month_fac = month(acq_date,
                           label = TRUE,
                           abbr = FALSE))
```

## Visualising the Fire Points

### Overall plot

```{r}
#| fig-width: 9

tm_shape(kbb_sf) +
  tm_polygons(fill = "grey95", col = "grey60", lwd = 0.5) +
  tm_shape(fire_sf) +
  tm_dots(size = 0.15, fill = "firebrick", fill_alpha = 0.6) +
  tm_title("Forest fire detections, Kepulauan Bangka Belitung, 2023")
```

### Visualising geographic distribution of forest fires by month

`tm_facets()` splits the same map by the month factor, so the seasonal movement of the fires becomes visible:

```{r}
#| fig-width: 11
#| fig-height: 9

tm_shape(kbb_sf) +
  tm_polygons(fill = "grey95", col = "grey70", lwd = 0.4) +
  tm_shape(fire_sf) +
  tm_dots(size = 0.1, fill = "firebrick", fill_alpha = 0.6) +
  tm_facets_wrap(by = "Month_fac",
                 ncol = 4,
                 free.coords = FALSE,
                 drop.units = TRUE)
```

## Computing STKDE by Month

This section computes the spatio-temporal KDE using `spattemp.density()` of the **sparr** package.

### Extracting forest fires by month

`as.ppp()` needs only the mark field and the geometry field, so the unwanted columns are dropped first:

```{r}
fire_month <- fire_sf %>%
  select(Month_num)
```

### Creating ppp

```{r}
fire_month_ppp <- as.ppp(fire_month)
fire_month_ppp
```

```{r}
summary(fire_month_ppp)
```

Checking for duplicated point events:

```{r}
any(duplicated(fire_month_ppp))
```

### Including owin object

Combining the point events with the study area window:

```{r}
fire_month_owin <- fire_month_ppp[kbb_owin]
summary(fire_month_owin)
```

Plotting the result to confirm it is correct:

```{r}
#| fig-width: 8
plot(fire_month_owin)
```

### Computing spatio-temporal KDE

`spattemp.density()` estimates the density over space and time jointly. With no bandwidths supplied it selects both automatically — an isotropic spatial bandwidth `h` and a temporal bandwidth `lambda`:

```{r}
st_kde <- spattemp.density(fire_month_owin)
summary(st_kde)
```

### Plotting the spatio-temporal KDE object

The estimate is a stack of surfaces, one per time slice. Plotting July to December shows the second half of the year, when the fires concentrate:

```{r}
#| fig-width: 12
#| fig-height: 8

tims <- c(7, 8, 9, 10, 11, 12)
par(mfcol = c(2, 3))
for (i in tims) {
  plot(st_kde, i,
       override.par = FALSE,
       fix.range = TRUE,
       main = paste("KDE at month", i))
}
```

`fix.range = TRUE` holds the colour scale constant across the six panels, so a given colour means the same density in every month. Without it each panel is scaled to its own range and the months cannot be compared.

## Computing STKDE by Day of Year

The same estimation is repeated at daily resolution.

### Creating ppp object

```{r}
fire_yday_ppp <- fire_sf %>%
  select(DayofYear) %>%
  as.ppp()
```

### Including owin object

```{r}
fire_yday_owin <- fire_yday_ppp[kbb_owin]
summary(fire_yday_owin)
```

### Computing the STKDE

```{r}
kde_yday <- spattemp.density(fire_yday_owin)
summary(kde_yday)
```

::: callout-warning
## Plotting every day is not useful

`plot(kde_yday)` draws **one panel per time slice**, and the daily lattice has over three hundred of them. The result is an unreadable wall of tiny plots that takes a long time to render.

A selection of days is plotted instead, chosen to span the fire season identified in the monthly analysis:
:::

```{r}
#| fig-width: 12
#| fig-height: 8

days <- c(200, 230, 260, 280, 300, 330)
par(mfcol = c(2, 3))
for (i in days) {
  plot(kde_yday, i,
       override.par = FALSE,
       fix.range = TRUE,
       main = paste("KDE at day", i))
}
```

## Computing STKDE by Day of Year: Improved Method

`BOOT.spattemp()` of **sparr** selects the bandwidths by bootstrap estimation of the mean integrated squared error, returning an isotropic scalar spatial bandwidth and a scalar temporal bandwidth. This is a better-founded choice than the defaults used above:

```{r}
set.seed(1234)
BOOT.spattemp(fire_yday_owin)
```

### Computing spatio-temporal KDE

The STKDE is now derived using the `h` and `lambda` values the bootstrap returned, rounded:

```{r}
kde_yday <- spattemp.density(fire_yday_owin,
                             h = 21900,
                             lambda = 19.9)
summary(kde_yday)
```

::: callout-warning
## Do not carry the textbook's numbers across

The textbook reaches `h = 9105` and hardcodes `h = 9000`. The bootstrap run above returns **h = 21893** on this data — roughly 2.4 times larger — because the study area here includes Belitung and holds more events, and the optimal spatial bandwidth scales with both.

The temporal bandwidth is the opposite case: **lambda = 19.89** against the textbook's 18.85, almost the same, because the fire season has the same length either way. Copying `h = 9000` across would have imposed a kernel far too narrow for this window and produced a spuriously spiky surface.
:::

### Plotting the output spatio-temporal KDE

```{r}
#| fig-width: 12
#| fig-height: 8

par(mfcol = c(2, 3))
for (i in days) {
  plot(kde_yday, i,
       override.par = FALSE,
       fix.range = TRUE,
       main = paste("KDE at day", i))
}
```

## Spatio-temporal Point Patterns Analysis: stpp Methods

This section uses functions of the **stpp** package to perform spatio-temporal point patterns analysis.

### Preparing spatio-temporal point process object of stpp

**Step 1**: Extracting the forest fire coordinates from the fire point events.

```{r}
coords <- st_coordinates(fire_sf)
```

**Step 2**: Creating a data frame by combining the x- and y-coordinates with the temporal event. The temporal event must be an integer.

```{r}
fire_df <- data.frame(
  x = coords[, 1],
  y = coords[, 2],
  t = fire_sf$`DayofYear`)
```

**Step 3**: Creating the `stpp` spatio-temporal object with `as.3dpoints()`.

```{r}
fire_stpp <- as.3dpoints(fire_df)
class(fire_stpp)
```

```{r}
#| fig-width: 10
#| fig-height: 5
plot(fire_stpp)
```

### Computing spatio-temporal K-function

`STIKhat()` computes the space-time inhomogeneous K-function:

```{r}
kbb_stik <- STIKhat(fire_stpp)
```

`plotK()` visualises the output:

```{r}
#| fig-width: 10
#| fig-height: 5
plotK(kbb_stik)
```

### Guide to interpret the plot

- If the contours show high values for small values of *u* and *v*, it suggests clustering at short spatial and temporal distances, meaning events occur close to each other in both space and time.
- If the contours are flat or show low values, it indicates a more random distribution or a lack of significant clustering at those distances.

The spatial and temporal extent of clustering can be understood by observing how rapidly the contours increase or decrease along the axes — the spacing and values of the contour lines show at which distances clustering occurs.

## Reference

- Kam, T. S. *R for Geospatial Data Science and Analytics*, [Chapter 6: Spatio-Temporal Point Patterns Analysis](https://r4gdsa.netlify.app/chap06).
- Peter Hall (1990) "Using the Bootstrap to Estimate Mean Squared Error and Select Smoothing Parameter in Nonparametric Problems", *Journal of Multivariate Analysis*, Vol. 32, pp. 177-203.
- González, J. A. et al. (2016) ["Spatio-temporal point process statistics: A review"](https://www.sciencedirect.com/science/article/abs/pii/S2211675316301130), *Spatial Statistics*, Vol. 18, Part B, pp. 505-544.
- [sparr: Spatial and Spatiotemporal Relative Risk](https://tilmandavies.github.io/sparr/index.html)
- [stpp: An R Package for Plotting, Simulating and Analyzing Spatio-Temporal Point Patterns](https://www.jstatsoft.org/article/view/v053i02), *Journal of Statistical Software*, Vol. 53, Issue 2.
