---
title: "Hands-on Exercise 2b: 2nd Order Spatial Point Patterns Analysis"
author: "Ji Guofang"
date: "last-modified"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

::: {.en}
[Hands-on Exercise 2a](hand-on-ex2a.qmd) used first-order methods, which describe how the *intensity* of a point process varies across space, meaning where events are dense and where they are sparse. Kernel density estimation answers "how many events per unit area, here?"

Second-order methods ask a different question: given the intensity, do events attract or repel one another, and **at what distance**? The Clark-Evans test at the end of the previous exercise condensed the whole pattern into a single number and so could not say. The four summary functions in this exercise, G, F, K and L, are functions of distance, so they can.

The null hypothesis tested throughout is **complete spatial randomness (CSR)**: the childcare centres are randomly distributed over the study area, following a homogeneous Poisson process. The significance level is 0.001.
:::

::: {.zh}
[练习 2a](hand-on-ex2a.qmd) 用的是一阶方法，描述点过程的*强度*如何随空间变化，也就是哪里事件密、哪里事件疏。核密度估计回答的是"此处每单位面积有多少个事件"。

二阶方法问的是另一个问题：在强度已给定的前提下，事件之间是相互吸引还是相互排斥，而且**在什么距离上**如此？上一份练习末尾的 Clark-Evans 检验把整个格局压成一个数，因此答不了这个问题。本练习的四个汇总函数 G、F、K、L 都是距离的函数，可以。

全程检验的零假设是**完全空间随机（CSR）**：托儿中心在研究区内随机分布，服从齐性泊松过程。显著性水平取 0.001。
:::

## The Data

| Dataset | Source | Format |
|----|----|----|
| Child Care Services | [data.gov.sg](https://data.gov.sg/) | GeoJSON, point |
| Master Plan 2019 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/) | GeoJSON, polygon |

## Installing and Loading the R Packages

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

::: {.en}
**terra** is not needed here, because no density surface is converted to a raster in this exercise.
:::

::: {.zh}
本练习不需要 **terra**，因为没有任何密度面要转成栅格。
:::

## Data Import and Preparation

::: {.en}
The cleaned subzone layer saved in the previous exercise is reused:
:::

::: {.zh}
沿用上一份练习保存下来的清理后分区图层：
:::

```{r}
mpsz_cl <- read_rds("data/rds/mpsz_cl.rds")
```

```{r}
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

::: {.en}
Converting to a `ppp` and jittering the duplicates, exactly as in Hands-on Exercise 2a. `st_geometry()` is used so the pattern is unmarked; passing the whole `sf` object would turn the sixteen attribute columns into marks and stop `duplicated()` from seeing coincident locations. The summary functions below assume a simple point pattern, so removing the 187 duplicates is a precondition rather than a nicety:
:::

::: {.zh}
转成 `ppp` 并对重合点做抖动，做法与练习 2a 完全相同。这里传 `st_geometry()` 是为了得到无标记的点格局；若把整个 `sf` 对象传进去，十六个属性列会变成标记，`duplicated()` 就看不出坐标重合了。下面的汇总函数都假定点格局是简单的，所以去掉那 187 个重合点是前提条件，而不是锦上添花：
:::

```{r}
childcare_ppp <- as.ppp(st_geometry(childcare_sf))

set.seed(1234)
childcare_ppp_jit <- rjitter(childcare_ppp,
                             retry = TRUE,
                             nsim = 1,
                             drop = TRUE)

any(duplicated(childcare_ppp_jit))
```

### The two study areas

::: {.en}
The analysis is run on two planning areas rather than the whole country. Second-order functions assume the process is homogeneous, meaning that intensity is roughly constant across the window, and that assumption is untenable for Singapore as a whole, where the centre and the outlying areas differ by an order of magnitude. Within one planning area it is defensible:
:::

::: {.zh}
分析放在两个规划区而不是全国做。二阶函数假定过程是齐性的，即强度在整个窗口内大致恒定；放在整个新加坡这个假定站不住，市中心与外围地区的强度差了一个数量级。限定在单个规划区内，这个假定才说得过去：
:::

```{r}
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")

childcare_ck_ppp <- childcare_ppp_jit[as.owin(ck)]
childcare_tm_ppp <- childcare_ppp_jit[as.owin(tm)]

c(`Choa Chu Kang` = npoints(childcare_ck_ppp),
  Tampines        = npoints(childcare_tm_ppp))
```

```{r}
#| fig-width: 12
#| fig-height: 5

par(mfrow = c(1, 2))
plot(childcare_ck_ppp, main = "Choa Chu Kang")
plot(childcare_tm_ppp, main = "Tampines")
```

::: callout-note
## Reading a Monte Carlo envelope

Each function below is accompanied by an `envelope()` plot. The procedure simulates many point patterns under CSR, computes the same function on each, and shades the region between the pointwise minimum and maximum.

The interpretation is the same in every case: where the observed curve lies **inside** the grey envelope, the pattern is indistinguishable from randomness at that distance. Where it strays outside, CSR is rejected, and which side it exits on says whether the departure is clustering or regularity.

`set.seed()` is called before every envelope so the simulations are reproducible.
:::

## Second-order Spatial Point Patterns Analysis

::: {.en}
All four functions below are *distance-dependent* summaries of the same pattern, and each is tested against CSR with a Monte Carlo envelope. They differ in what they measure from, which is why all four are worked through here rather than one being reported on its own.
:::

::: {.zh}
下面四个函数都是同一个点格局的*随距离变化*的汇总，每一个都用蒙特卡洛包络对照 CSR 检验。它们的差别在于从什么出发测量，所以这里四个全部走一遍，而不是只报其中一个。
:::

## Analysing Spatial Point Process Using G-Function

::: {.en}
The G-function measures the distribution of distances from an event to its **nearest neighbour**. G(r) is the proportion of events whose nearest neighbour lies within distance *r*.

A curve rising **faster** than expected under CSR means nearest neighbours are closer than chance allows, i.e. **clustering**. A curve rising more slowly means events keep their distance, i.e. **regularity**.
:::

::: {.zh}
G 函数度量的是每个事件到其**最近邻**的距离分布。G(r) 表示最近邻落在距离 *r* 以内的事件所占比例。

曲线比 CSR 预期上升得**更快**，说明最近邻比随机情形更靠近，即**聚集**；上升得更慢，说明事件彼此保持距离，即**规则分布**。
:::

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

G_CK <- Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim = c(0, 500))
```

::: {.en}
Testing against CSR with 999 simulations:
:::

::: {.zh}
用 999 次模拟对照 CSR 检验：
:::

```{r}
#| fig-width: 8

set.seed(1234)
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
plot(G_CK.csr)
```

### Tampines planning area

```{r}
#| fig-width: 8

G_tm <- Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

```{r}
#| fig-width: 8

set.seed(1234)
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
plot(G_tm.csr)
```

## Analysing Spatial Point Process Using F-Function

::: {.en}
The F-function is the complement of G. Instead of measuring from events, it measures from **arbitrary locations** in the window to the nearest event. F(r) is the proportion of such locations with an event within distance *r*.

The two are sensitive to different things. G describes how events sit relative to each other; F describes how much empty space the pattern leaves. A clustered pattern leaves large gaps, so its F curve rises more slowly than under CSR, the opposite direction to its G curve. Reporting both guards against a conclusion that is an artefact of one measurement's blind spot.
:::

::: {.zh}
F 函数是 G 的补。它不从事件出发，而是从窗口内**任意位置**量到最近的事件。F(r) 表示这类位置中，距离 *r* 以内有事件的比例。

两者敏感的东西不同。G 描述事件彼此之间如何落位，F 描述格局留下了多少空白。聚集的格局会留下大片空隙，所以它的 F 曲线比 CSR 下上升得更慢，方向与 G 曲线相反。两个都报，可以避免结论只是某一种测量方式盲区造成的假象。
:::

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

F_CK <- Fest(childcare_ck_ppp)
plot(F_CK)
```

#### Performing Complete Spatial Randomness Test

```{r}
#| fig-width: 8

set.seed(1234)
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
plot(F_CK.csr)
```

### Tampines planning area

```{r}
#| fig-width: 8

F_tm <- Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

```{r}
#| fig-width: 8

set.seed(1234)
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
plot(F_tm.csr)
```

## Analysing Spatial Point Process Using K-Function

::: {.en}
G and F both look only at the nearest event, which throws away most of the pattern. Ripley's K-function counts **all** events within distance *r* of each event, averaged over events and standardised by intensity. It therefore describes the pattern across the whole range of distances at once.

The plots below are drawn as `K(d) - r` rather than K(d) itself. Under CSR, K(r) = πr², and subtracting the expectation flattens that curve onto the horizontal axis, which makes departures far easier to see than a comparison of two rising curves.
:::

::: {.zh}
G 和 F 都只看最近的那一个事件，把格局的大部分信息丢掉了。Ripley 的 K 函数统计每个事件在距离 *r* 以内的**所有**事件，对全部事件取平均并按强度标准化，因此能一次性描述整个距离范围上的格局。

下面的图画的是 `K(d) - r` 而不是 K(d) 本身。CSR 下 K(r) = πr²，减去期望值可把该曲线压平到横轴上，比起对比两条上升曲线，偏离要容易看得多。
:::

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)")
```

```{r}
#| fig-width: 8

set.seed(1234)
K_ck.csr <- envelope(childcare_ck_ppp, Kest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(K_ck.csr, . - r ~ r, xlab = "d", ylab = "K(d)-r")
```

### Tampines planning area

```{r}
#| fig-width: 8

K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)", xlim = c(0, 1000))
```

```{r}
#| fig-width: 8

set.seed(1234)
K_tm.csr <- envelope(childcare_tm_ppp, Kest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(K_tm.csr, . - r ~ r, xlab = "d", ylab = "K(d)-r", xlim = c(0, 500))
```

::: callout-note
## Why `glocal = TRUE` and only 99 simulations

`glocal = TRUE` requests a *global* envelope. A pointwise envelope is a separate test at every distance, so with hundreds of distances some will fall outside by chance alone and the apparent significance is inflated. A global envelope has the stated coverage across the whole curve simultaneously, which is what a claim about the pattern actually requires.

The simulation count drops to 99 because K is far more expensive to compute than G, since it counts every pair of events at every distance rather than one distance per event. With `rank = 1` and 99 simulations the envelope is the pointwise extreme of those runs, giving a test at the 0.01 level.
:::

## Analysing Spatial Point Process Using L-Function

::: {.en}
K(r) grows as r², so its variance grows with distance too, and the envelope fans out at larger *r* even when nothing is happening. The L-function is the square-root transformation that removes this:
:::

::: {.zh}
K(r) 随 r² 增长，它的方差也随距离增长，所以即便什么都没发生，包络在 *r* 较大处也会张开。L 函数就是消除这一点的平方根变换：
:::

$$L(r) = \sqrt{\frac{K(r)}{\pi}}$$

::: {.en}
Under CSR, L(r) = r, so plotting `L(d) - r` again flattens the expectation onto the horizontal axis, but now with roughly constant variance across the whole range, which makes the envelope width comparable at every distance. This is why L is normally the function reported.
:::

::: {.zh}
CSR 下 L(r) = r，所以画 `L(d) - r` 同样把期望压平到横轴，但此时整个范围上的方差大致恒定，包络宽度在各距离上可比。这就是通常以 L 作为上报函数的原因。
:::

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

L_ck <- Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)")
```

```{r}
#| fig-width: 8

set.seed(1234)
L_ck.csr <- envelope(childcare_ck_ppp, Lest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(L_ck.csr, . - r ~ r, xlab = "d", ylab = "L(d)-r")
```

### Tampines planning area

```{r}
#| fig-width: 8

L_tm <- Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)", xlim = c(0, 1000))
```

```{r}
#| fig-width: 8

set.seed(1234)
L_tm.csr <- envelope(childcare_tm_ppp, Lest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(L_tm.csr, . - r ~ r, xlab = "d", ylab = "L(d)-r", xlim = c(0, 500))
```

## Summary

| Function | Measures from | Clustering shows as |
|----|----|----|
| **G** | each event to its nearest event | curve **above** the envelope |
| **F** | arbitrary locations to the nearest event | curve **below** the envelope |
| **K** | each event to all events within *r* | curve **above** the envelope |
| **L** | as K, variance-stabilised | curve **above** the envelope |

::: {.en}
The direction reverses for F because it measures empty space rather than proximity: a clustered pattern leaves larger gaps, so a randomly chosen location is *further* from the nearest event than CSR predicts. Reading F in the same direction as G therefore inverts the conclusion.
:::

::: {.zh}
F 的方向之所以反过来，是因为它量的是空白而不是邻近：聚集的格局留下更大的空隙，于是随机取到的位置到最近事件的距离比 CSR 预期的*更远*。把 F 按 G 的方向去读，结论会正好读反。
:::

::: {.en}
All four are **global** statistics: they characterise the pattern as a whole and cannot say *where* any clustering sits. That question requires local methods, so none of the four supports a claim about a particular part of the map.
:::

::: {.zh}
这四个都是**全局**统计量：它们刻画的是整体格局，说不出聚集究竟*在哪里*。那个问题要靠局部方法，所以四者都不足以支撑关于地图上某一具体部位的论断。
:::

## Reference

- Kam, T. S. *R for Geospatial Data Science and Analytics*, [Chapter 5: 2nd Order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap05.html).
- Baddeley, A., Rubak, E. and Turner, R. (2015) *Spatial Point Patterns: Methodology and Applications with R*. Chapman and Hall/CRC.
- O'Sullivan, D. and Unwin, D. (2010) *Geographic Information Analysis*, 2nd Edition. Chapters 5-6.
- Yuan, Y., Qiang, Y., Bin Asad, K., and Chow, T. E. (2020). [Point Pattern Analysis](https://gistbok-ltb.ucgis.org/page/30/concept/9213). *The GIS&T Body of Knowledge*.
