---
title: "Hands-on Exercise 2b: 2nd Order Spatial Point Patterns Analysis"
author: "Ji Guofang"
date: "last-modified"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

[Hands-on Exercise 2a](hand-on-ex2a.qmd) used first-order methods, which describe how the *intensity* of a point process varies across space, meaning where events are dense and where they are sparse. Kernel density estimation answers "how many events per unit area, here?"

Second-order methods ask a different question: given the intensity, do events attract or repel one another, and **at what distance**? The Clark-Evans test at the end of the previous exercise condensed the whole pattern into a single number and so could not say. The four summary functions in this exercise, G, F, K and L, are functions of distance, so they can.

The null hypothesis tested throughout is **complete spatial randomness (CSR)**: the childcare centres are randomly distributed over the study area, following a homogeneous Poisson process. The significance level is 0.001.

## The Data

| Dataset | Source | Format |
|----|----|----|
| Child Care Services | [data.gov.sg](https://data.gov.sg/) | GeoJSON, point |
| Master Plan 2019 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/) | GeoJSON, polygon |

## Installing and Loading the R Packages

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

**terra** is not needed here, because no density surface is converted to a raster in this exercise.

## Data Import and Preparation

The cleaned subzone layer saved in the previous exercise is reused:

```{r}
mpsz_cl <- read_rds("data/rds/mpsz_cl.rds")
```

```{r}
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

Converting to a `ppp` and jittering the duplicates, exactly as in Hands-on Exercise 2a. `st_geometry()` is used so the pattern is unmarked; passing the whole `sf` object would turn the sixteen attribute columns into marks and stop `duplicated()` from seeing coincident locations. The summary functions below assume a simple point pattern, so removing the 187 duplicates is a precondition rather than a nicety:

```{r}
childcare_ppp <- as.ppp(st_geometry(childcare_sf))

set.seed(1234)
childcare_ppp_jit <- rjitter(childcare_ppp,
                             retry = TRUE,
                             nsim = 1,
                             drop = TRUE)

any(duplicated(childcare_ppp_jit))
```

### The two study areas

The analysis is run on two planning areas rather than the whole country. Second-order functions assume the process is homogeneous, meaning that intensity is roughly constant across the window, and that assumption is untenable for Singapore as a whole, where the centre and the outlying areas differ by an order of magnitude. Within one planning area it is defensible:

```{r}
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")

childcare_ck_ppp <- childcare_ppp_jit[as.owin(ck)]
childcare_tm_ppp <- childcare_ppp_jit[as.owin(tm)]

c(`Choa Chu Kang` = npoints(childcare_ck_ppp),
  Tampines        = npoints(childcare_tm_ppp))
```

```{r}
#| fig-width: 12
#| fig-height: 5

par(mfrow = c(1, 2))
plot(childcare_ck_ppp, main = "Choa Chu Kang")
plot(childcare_tm_ppp, main = "Tampines")
```

::: callout-note
## Reading a Monte Carlo envelope

Each function below is accompanied by an `envelope()` plot. The procedure simulates many point patterns under CSR, computes the same function on each, and shades the region between the pointwise minimum and maximum.

The interpretation is the same in every case: where the observed curve lies **inside** the grey envelope, the pattern is indistinguishable from randomness at that distance. Where it strays outside, CSR is rejected, and which side it exits on says whether the departure is clustering or regularity.

`set.seed()` is called before every envelope so the simulations are reproducible.
:::

## Second-order Spatial Point Patterns Analysis

All four functions below are *distance-dependent* summaries of the same pattern, and each is tested against CSR with a Monte Carlo envelope. They differ in what they measure from, which is why the textbook works through all four rather than reporting one.

## Analysing Spatial Point Process Using G-Function

The G-function measures the distribution of distances from an event to its **nearest neighbour**. G(r) is the proportion of events whose nearest neighbour lies within distance *r*.

A curve rising **faster** than expected under CSR means nearest neighbours are closer than chance allows, i.e. **clustering**. A curve rising more slowly means events keep their distance, i.e. **regularity**.

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

G_CK <- Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim = c(0, 500))
```

Testing against CSR with 999 simulations:

```{r}
#| fig-width: 8

set.seed(1234)
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
plot(G_CK.csr)
```

### Tampines planning area

```{r}
#| fig-width: 8

G_tm <- Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

```{r}
#| fig-width: 8

set.seed(1234)
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
plot(G_tm.csr)
```

## Analysing Spatial Point Process Using F-Function

The F-function is the complement of G. Instead of measuring from events, it measures from **arbitrary locations** in the window to the nearest event. F(r) is the proportion of such locations with an event within distance *r*.

The two are sensitive to different things. G describes how events sit relative to each other; F describes how much empty space the pattern leaves. A clustered pattern leaves large gaps, so its F curve rises more slowly than under CSR, the opposite direction to its G curve. Reporting both guards against a conclusion that is an artefact of one measurement's blind spot.

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

F_CK <- Fest(childcare_ck_ppp)
plot(F_CK)
```

#### Performing Complete Spatial Randomness Test

```{r}
#| fig-width: 8

set.seed(1234)
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
plot(F_CK.csr)
```

### Tampines planning area

```{r}
#| fig-width: 8

F_tm <- Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

```{r}
#| fig-width: 8

set.seed(1234)
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
plot(F_tm.csr)
```

## Analysing Spatial Point Process Using K-Function

G and F both look only at the nearest event, which throws away most of the pattern. Ripley's K-function counts **all** events within distance *r* of each event, averaged over events and standardised by intensity. It therefore describes the pattern across the whole range of distances at once.

The plots below are drawn as `K(d) - r` rather than K(d) itself. Under CSR, K(r) = πr², and subtracting the expectation flattens that curve onto the horizontal axis, which makes departures far easier to see than a comparison of two rising curves.

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

K_ck <- Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)")
```

```{r}
#| fig-width: 8

set.seed(1234)
K_ck.csr <- envelope(childcare_ck_ppp, Kest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(K_ck.csr, . - r ~ r, xlab = "d", ylab = "K(d)-r")
```

### Tampines planning area

```{r}
#| fig-width: 8

K_tm <- Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . - r ~ r, ylab = "K(d)-r", xlab = "d(m)", xlim = c(0, 1000))
```

```{r}
#| fig-width: 8

set.seed(1234)
K_tm.csr <- envelope(childcare_tm_ppp, Kest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(K_tm.csr, . - r ~ r, xlab = "d", ylab = "K(d)-r", xlim = c(0, 500))
```

::: callout-note
## Why `glocal = TRUE` and only 99 simulations

`glocal = TRUE` requests a *global* envelope. A pointwise envelope is a separate test at every distance, so with hundreds of distances some will fall outside by chance alone and the apparent significance is inflated. A global envelope has the stated coverage across the whole curve simultaneously, which is what a claim about the pattern actually requires.

The simulation count drops to 99 because K is far more expensive to compute than G, since it counts every pair of events at every distance rather than one distance per event. With `rank = 1` and 99 simulations the envelope is the pointwise extreme of those runs, giving a test at the 0.01 level.
:::

## Analysing Spatial Point Process Using L-Function

K(r) grows as r², so its variance grows with distance too, and the envelope fans out at larger *r* even when nothing is happening. The L-function is the square-root transformation that removes this:

$$L(r) = \sqrt{\frac{K(r)}{\pi}}$$

Under CSR, L(r) = r, so plotting `L(d) - r` again flattens the expectation onto the horizontal axis, but now with roughly constant variance across the whole range, which makes the envelope width comparable at every distance. This is why L is normally the function reported.

### Choa Chu Kang planning area

```{r}
#| fig-width: 8

L_ck <- Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)")
```

```{r}
#| fig-width: 8

set.seed(1234)
L_ck.csr <- envelope(childcare_ck_ppp, Lest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(L_ck.csr, . - r ~ r, xlab = "d", ylab = "L(d)-r")
```

### Tampines planning area

```{r}
#| fig-width: 8

L_tm <- Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . - r ~ r, ylab = "L(d)-r", xlab = "d(m)", xlim = c(0, 1000))
```

```{r}
#| fig-width: 8

set.seed(1234)
L_tm.csr <- envelope(childcare_tm_ppp, Lest,
                     nsim = 99, rank = 1, glocal = TRUE)
plot(L_tm.csr, . - r ~ r, xlab = "d", ylab = "L(d)-r", xlim = c(0, 500))
```

## Summary

| Function | Measures from | Clustering shows as |
|----|----|----|
| **G** | each event to its nearest event | curve **above** the envelope |
| **F** | arbitrary locations to the nearest event | curve **below** the envelope |
| **K** | each event to all events within *r* | curve **above** the envelope |
| **L** | as K, variance-stabilised | curve **above** the envelope |

The direction reverses for F because it measures empty space rather than proximity: a clustered pattern leaves larger gaps, so a randomly chosen location is *further* from the nearest event than CSR predicts. Reading F in the same direction as G therefore inverts the conclusion.

All four are **global** statistics: they characterise the pattern as a whole and cannot say *where* any clustering sits. That question requires local methods, so none of the four supports a claim about a particular part of the map.

## Reference

- Kam, T. S. *R for Geospatial Data Science and Analytics*, [Chapter 5: 2nd Order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap05.html).
- Baddeley, A., Rubak, E. and Turner, R. (2015) *Spatial Point Patterns: Methodology and Applications with R*. Chapman and Hall/CRC.
- O'Sullivan, D. and Unwin, D. (2010) *Geographic Information Analysis*, 2nd Edition. Chapters 5-6.
- Yuan, Y., Qiang, Y., Bin Asad, K., and Chow, T. E. (2020). [Point Pattern Analysis](https://gistbok-ltb.ucgis.org/page/30/concept/9213). *The GIS&T Body of Knowledge*.
