---
title: "Hands-on Exercise 2a: 1st Order Spatial Point Patterns Analysis"
author: "Ji Guofang"
date: "last-modified"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

Spatial Point Pattern Analysis evaluates the pattern or distribution of a set of points on a surface. The points may be the locations of events such as crimes, traffic accidents or disease onset, or of business services such as coffee shops and bus stops.

This exercise uses the **spatstat** package to discover the spatial point processes of childcare centres in Singapore. The questions driving the analysis are:

- Are the childcare centres in Singapore randomly distributed across the country?
- If not, where are the locations with a higher concentration of childcare centres?

## The Data

| Dataset | Source | Format |
|----|----|----|
| Child Care Services | [data.gov.sg](https://data.gov.sg/) | GeoJSON, point |
| Master Plan 2019 Subzone Boundary (No Sea) | [data.gov.sg](https://data.gov.sg/) | GeoJSON, polygon |

## Installing and Loading the R Packages

```{r}
pacman::p_load(sf, terra, spatstat, tmap, tidyverse)
```

| Package | Role in this exercise |
|----|----|
| **sf** | Importing and handling the vector layers |
| **spatstat** | The point pattern analysis engine: `ppp`, `owin`, KDE, bandwidth selection |
| **terra** | Converting spatstat's `im` output into a raster **tmap** can draw |
| **tmap** | Plotting the point map and the density surfaces |
| **tidyverse** | Data import and wrangling |

`spatstat` is a meta-package. Loading it pulls in `spatstat.geom` (the `ppp` and `owin` classes), `spatstat.random` (simulation), `spatstat.explore` (density estimation and the summary functions used in Hands-on Exercise 2b), and others. Everything below comes from one of these.

## Importing and Wrangling Geospatial Data Sets

### The subzone boundary

```{r}
mpsz_sf <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

The GeoJSON release of this layer is used rather than the KML one the textbook imports. KML stores every attribute inside a single `Description` field as an HTML table, so parsing them out needs **rvest** and a custom `extract_kml_field()` helper. The GeoJSON release carries `REGION_N`, `PLN_AREA_N`, `SUBZONE_N` and `SUBZONE_C` as ordinary columns, so that step is not required here and **rvest** is not loaded.

The offshore islands are excluded. They are separated from the mainland by open water, and leaving them in would stretch the observation window across areas no childcare centre could occupy, distorting every density estimate that follows:

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")

c(before = nrow(mpsz_sf), after = nrow(mpsz_cl))
```

Saving the cleaned layer so that Hands-on Exercise 2b can reuse it:

```{r}
write_rds(mpsz_cl, "data/rds/mpsz_cl.rds")
```

### The childcare centres

```{r}
childcare_sf <- st_read("data/geospatial/ChildCareServices.geojson") %>%
  st_zm(drop = TRUE, what = "ZM") %>%
  st_transform(crs = 3414)
```

The import reports 1925 point features with proper attribute columns, among them `NAME`, the address fields and `DESCRIPTION`, so as with the subzone layer no HTML parsing is needed.

`st_zm()` drops any Z (and M) dimension. This GeoJSON release is already plain XY, so the call does nothing here, but it is kept as a guard: the KML release of the same dataset carries a Z coordinate fixed at 0, and spatstat's `ppp` class is strictly two-dimensional, so the conversion below fails if a third coordinate survives. Keeping the call means the document works with either download format.

### Mapping the geospatial data sets

```{r}
#| fig-width: 9

tm_shape(mpsz_cl) +
  tm_polygons(fill = "grey95", col = "grey70", lwd = 0.4) +
  tm_shape(childcare_sf) +
  tm_dots(size = 0.05, fill = "firebrick") +
  tm_title("Childcare centres in Singapore")
```

The same map in interactive mode, which allows zooming into individual centres:

```{r}
tmap_mode("view")

tm_shape(childcare_sf) +
  tm_dots(size = 0.02) +
  tm_view(set_zoom_limits = c(11, 14))
```

```{r}
tmap_mode("plot")
```

## Geospatial Data Wrangling

spatstat does not work with `sf` objects. A point pattern must be a `ppp` object, which bundles the event coordinates together with an observation window (`owin`) describing the region within which the events could have occurred.

### Converting sf data frames to ppp class

```{r}
childcare_ppp <- as.ppp(st_geometry(childcare_sf))
childcare_ppp
```

::: callout-warning
## Pass the geometry, not the whole sf object

`as.ppp(childcare_sf)` also works, but it converts all sixteen attribute columns into **marks** and returns a *marked* point pattern. Two things then go wrong, neither of them loudly:

- `plot()` stops drawing the points. Given a data frame of marks it draws one panel per mark column and prints the attribute values as text, producing an unreadable wall of overlapping labels instead of a map.
- `duplicated()` compares marks as well as coordinates. Two centres at the same address with different names are then not counted as duplicates, and the check in the next section reports zero where the true figure is 187.

`st_geometry()` strips the attributes and yields an unmarked pattern, which is what every method in this exercise needs. None of them use the attributes.
:::

```{r}
#| fig-width: 8
plot(childcare_ppp)
```

```{r}
summary(childcare_ppp)
```

#### Handling duplicated points

Duplicates are a genuine problem rather than a formality. Much of point pattern theory assumes a *simple* point pattern, in which no two events share a location; where duplicates exist, several statistical methods either ignore the repeats or fail outright.

```{r}
any(duplicated(childcare_ppp))
```

`multiplicity()` reports how many times each point is repeated, so the number of points involved can be counted:

```{r}
sum(multiplicity(childcare_ppp) > 1)
```

The duplication is substantial: 187 points sit on a location already occupied, and 338 points are involved in a coincidence, so the 1925 centres occupy only 1738 distinct locations. Multiple centres registered at the same block or the same mall share one address point.

There are three ways to deal with duplicates. Deleting them is simplest but discards real centres. Attaching weights to the retained points keeps the information but complicates every subsequent analysis. Jittering, which displaces each duplicate by a tiny random amount so that no two points coincide, keeps all the events and is the approach used here:

```{r}
set.seed(1234)
childcare_ppp_jit <- rjitter(childcare_ppp,
                             retry = TRUE,
                             nsim = 1,
                             drop = TRUE)

any(duplicated(childcare_ppp_jit))
```

`retry = TRUE` makes the function try again if a displacement happens to land on another point, so the result is guaranteed duplicate-free.

::: callout-warning
## Seed the jitter, not just the envelopes

`set.seed()` is called before `rjitter()`, which the textbook does not do, since it seeds only before the simulation envelopes. That is not enough, because `rjitter()` displaces every point by a random amount, not merely the duplicated ones. Without a seed, every re-render produces a slightly different pattern, and with it different Clark-Evans statistics, different selected bandwidths, and different G, F, K and L curves.

Rendering this document twice without the seed gave *R* = 0.55568 and then *R* = 0.55431 from identical code. The conclusions are unaffected at that magnitude, but any number quoted in the text stops matching the output beneath it. Hands-on Exercise 2b seeds the same call with the same value, so both documents work from the identical jittered pattern.
:::

### Creating owin object

An `owin` confines the analysis to the region where events could actually occur. Without one, spatstat uses the bounding box of the points, which for Singapore would include a large amount of sea and Malaysian territory. Density is events divided by area, so the wrong window gives the wrong answer everywhere.

```{r}
sg_owin <- as.owin(mpsz_cl)
sg_owin
```

```{r}
#| fig-width: 8
plot(sg_owin)
```

### Combining point events object and owin object

Subsetting a `ppp` by an `owin` keeps only the events inside it and, more importantly, records that window as the pattern's domain:

```{r}
childcareSG_ppp <- childcare_ppp_jit[sg_owin]
childcareSG_ppp
```

```{r}
#| fig-width: 8
plot(childcareSG_ppp)
```

## Clark-Evans Test for Nearest Neighbour Analysis

The Clark-Evans test gives a single-number summary of whether the pattern is clustered, random or regular. It compares the observed mean nearest-neighbour distance against the value expected under complete spatial randomness (CSR).

The test statistic *R* is that ratio: **R < 1** indicates clustering, **R ≈ 1** randomness, and **R > 1** regularity.

The hypotheses are:

- **H₀**: the distribution of childcare services is randomly distributed.
- **H₁**: the distribution of childcare services is not randomly distributed.

The null hypothesis is rejected if the p-value is smaller than 0.001.

### Performing the Clark-Evans test without CSR

```{r}
clarkevans.test(childcareSG_ppp,
                correction = "none",
                clipregion = "sg_owin",
                alternative = c("clustered"),
                nsim = 99)
```

The header of that output reports a Z-test rather than a Monte Carlo test, even though `nsim = 99` was passed. The signature explains why:

```
clarkevans.test(X, ..., correction, clipregion = NULL, alternative = ...,
                method = c("asymptotic", "MonteCarlo"), nsim = 999)
```

`method` defaults to `"asymptotic"`, so `nsim` does nothing unless `method = "MonteCarlo"` is set explicitly, and no warning is raised. The `clipregion` argument is likewise inert with `correction = "none"`, since it is only consulted by the guard-area correction.

### Performing the Clark-Evans test with CSR

Asking for the simulated test gives the same *R* but a different p-value, because the asymptotic approximation is not reliable for a strongly clustered pattern:

```{r}
set.seed(1234)
clarkevans.test(childcareSG_ppp,
                correction = "none",
                alternative = c("clustered"),
                method = "MonteCarlo",
                nsim = 999)
```

Both tests reject H₀. *R* = 0.552 is well below 1, so the childcare centres are clustered rather than randomly distributed across Singapore.

The two p-values are not comparable at face value. The Z-test reports *p* < 2.2e-16 while the Monte Carlo test reports *p* = 0.001, which is not weaker evidence but the smallest value 999 simulations can produce, since the p-value is (1 + number of simulations at least as extreme) / (1 + nsim). Not one of the 999 CSR simulations came close to the observed *R*. Increasing `nsim` would lower that floor without changing the conclusion.

## Kernel Density Estimation Method

KDE estimates the intensity of the point process, meaning the expected number of events per unit area, as a smooth surface, by centring a kernel function on each event and summing the contributions.

The single most consequential choice is the **bandwidth** (`sigma`), the width of that kernel. Too small and the surface degenerates into a spike at every event; too large and every local structure is smoothed away.

### Working with automatic bandwidth selection method

`density()` accepts a bandwidth selection function in place of a number. `bw.diggle()` chooses the bandwidth by cross-validation, minimising a mean-square-error criterion:

```{r}
kde_childcareSG_bw <- density(childcareSG_ppp,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")
```

```{r}
#| fig-width: 8
plot(kde_childcareSG_bw)
```

`edge = TRUE` applies an edge correction. Without it, events near the coastline appear artificially sparse, because part of their kernel falls outside the window and is simply lost.

The chosen bandwidth can be retrieved directly:

```{r}
bw.diggle(childcareSG_ppp)
```

Note the density values on the colour bar. The layer is in svy21, so the coordinate unit is the metre and the intensity is reported as events per square metre, in numbers so small they are awkward to read.

### Rescaling KDE values

`rescale.ppp()` converts the pattern to kilometres, which puts the intensity on a readable scale. This changes the *units*, not the geometry: the pattern is identical, only reported differently.

```{r}
childcareSG_ppp.km <- rescale.ppp(childcareSG_ppp, 1000, "km")
childcareSG_ppp.km
```

Re-running the same estimation on the rescaled pattern produces an identical surface with sensible numbers on the legend:

```{r}
#| fig-width: 8

kde_childcareSG.bw <- density(childcareSG_ppp.km,
                              sigma = bw.diggle,
                              edge = TRUE,
                              kernel = "gaussian")

plot(kde_childcareSG.bw)
```

### Working with different automatic bandwidth methods

spatstat offers four selectors, and they do not agree:

```{r}
c(bw.diggle = bw.diggle(childcareSG_ppp.km),
  bw.CvL    = bw.CvL(childcareSG_ppp.km),
  bw.ppl    = bw.ppl(childcareSG_ppp.km))
```

```{r}
bw.scott(childcareSG_ppp.km)
```

`bw.CvL` returns a far larger bandwidth than the other two, and `bw.scott` returns one value per axis. Baddeley and colleagues recommend `bw.ppl` when the pattern consists predominantly of tight clusters, on the grounds that it tends to produce more sensible results in that case, while `bw.diggle` is preferable when the aim is to detect a single tight cluster against a sparse background.

Comparing the two most commonly used selectors side by side:

```{r}
#| fig-width: 12
#| fig-height: 5

kde_childcareSG.ppl <- density(childcareSG_ppp.km,
                               sigma = bw.ppl,
                               edge = TRUE,
                               kernel = "gaussian")

par(mfrow = c(1, 2))
plot(kde_childcareSG.bw, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

### Working with different kernel methods

The Gaussian kernel is the default, but not the only option. The kernel determines how an event's influence decays with distance:

```{r}
#| fig-width: 12
#| fig-height: 9

par(mfrow = c(2, 2))
for (k in c("gaussian", "epanechnikov", "quartic", "disc")) {
  plot(density(childcareSG_ppp.km,
               sigma = bw.ppl,
               edge = TRUE,
               kernel = k),
       main = k)
}
```

The four surfaces are strikingly similar. This is the standard finding in density estimation: the bandwidth matters a great deal, the kernel shape hardly at all.

## Fixed and Adaptive KDE

### Computing KDE by using fixed bandwidth

Instead of a selector, `sigma` can be given a number. Because the pattern was rescaled to kilometres, `sigma = 0.6` means a 600-metre kernel:

```{r}
#| fig-width: 8

kde_childcareSG_600 <- density(childcareSG_ppp.km,
                               sigma = 0.6,
                               edge = TRUE,
                               kernel = "gaussian")

plot(kde_childcareSG_600, main = "Fixed bandwidth, sigma = 0.6 km")
```

This is where the earlier rescaling matters. Had the pattern stayed in metres, `sigma = 0.6` would mean a 60-centimetre kernel and the surface would be meaningless. The number is always in the pattern's current units.

### Computing KDE by using adaptive bandwidth

A fixed bandwidth is a poor fit for Singapore, where density varies enormously between the city centre and the outlying areas. Adaptive estimation varies the bandwidth with the local intensity, narrow where events are dense and wide where they are sparse:

```{r}
#| fig-width: 12
#| fig-height: 5

kde_childcareSG_adaptive <- adaptive.density(childcareSG_ppp.km,
                                             method = "kernel")

par(mfrow = c(1, 2))
plot(kde_childcareSG.bw, main = "Fixed bandwidth")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth")
```

## Plotting Cartographic Quality KDE Map

### Converting gridded output into raster

spatstat returns an `im` (pixel image) object, which base `plot()` renders but **tmap** cannot. `terra::rast()` converts it to a `SpatRaster`.

The **unrescaled** estimate is used here, not `kde_childcareSG.bw`:

```{r}
kde_childcareSG_raster <- rast(kde_childcareSG_bw)
kde_childcareSG_raster
```

::: callout-warning
## Do not rasterise the kilometre version

Converting the km-rescaled image instead gives a raster whose extent runs from 2.67 to 55.94, which are kilometre numbers. Labelling that with `EPSG:3414`, whose unit is the metre, places the surface within 56 metres of the svy21 origin, roughly a thousandfold too small and nowhere near Singapore, so it will not overlay the subzone layer.

Rescaling is useful for reading intensity values off a legend, but the raster has to be built from the pattern in the units its CRS actually declares.
:::

### Assigning projection systems

The conversion carries the pixel values and the extent but not the coordinate reference system, which has to be assigned:

```{r}
crs(kde_childcareSG_raster) <- "EPSG:3414"
kde_childcareSG_raster
```

### Plotting KDE map with tmap

```{r}
#| fig-width: 9

tm_shape(kde_childcareSG_raster) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(title = "Density")) +
  tm_shape(mpsz_cl) +
  tm_borders(col = "white", lwd = 0.3) +
  tm_title("Kernel density of childcare centres") +
  tm_scalebar()
```

## First Order SPPA at the Planning Area Level

A national density surface is dominated by the overall settlement pattern. Restricting the analysis to individual planning areas makes local differences visible.

### Geospatial data wrangling

#### Extracting study area

```{r}
pg <- mpsz_cl %>% filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>% filter(PLN_AREA_N == "JURONG WEST")
```

```{r}
#| fig-width: 12
#| fig-height: 9

par(mfrow = c(2, 2))
plot(st_geometry(pg), main = "Punggol")
plot(st_geometry(tm), main = "Tampines")
plot(st_geometry(ck), main = "Choa Chu Kang")
plot(st_geometry(jw), main = "Jurong West")
```

#### Creating owin object

```{r}
pg_owin <- as.owin(pg)
tm_owin <- as.owin(tm)
ck_owin <- as.owin(ck)
jw_owin <- as.owin(jw)
```

#### Combining point events object and owin object

```{r}
childcare_pg_ppp <- childcare_ppp_jit[pg_owin]
childcare_tm_ppp <- childcare_ppp_jit[tm_owin]
childcare_ck_ppp <- childcare_ppp_jit[ck_owin]
childcare_jw_ppp <- childcare_ppp_jit[jw_owin]
```

Rescaling each to kilometres:

```{r}
childcare_pg_ppp.km <- rescale.ppp(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km <- rescale.ppp(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km <- rescale.ppp(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km <- rescale.ppp(childcare_jw_ppp, 1000, "km")
```

The number of centres in each:

```{r}
c(Punggol         = npoints(childcare_pg_ppp),
  Tampines        = npoints(childcare_tm_ppp),
  `Choa Chu Kang` = npoints(childcare_ck_ppp),
  `Jurong West`   = npoints(childcare_jw_ppp))
```

```{r}
#| fig-width: 12
#| fig-height: 9

par(mfrow = c(2, 2))
plot(childcare_pg_ppp.km, main = "Punggol")
plot(childcare_tm_ppp.km, main = "Tampines")
plot(childcare_ck_ppp.km, main = "Choa Chu Kang")
plot(childcare_jw_ppp.km, main = "Jurong West")
```

### Clark and Evans Test

#### Choa Chu Kang planning area

```{r}
clarkevans.test(childcare_ck_ppp,
                correction = "none",
                clipregion = NULL,
                alternative = c("two.sided"),
                nsim = 999)
```

#### Tampines planning area

```{r}
clarkevans.test(childcare_tm_ppp,
                correction = "none",
                clipregion = NULL,
                alternative = c("two.sided"),
                nsim = 999)
```

The two areas do not give the same answer.

| Planning area | *R* | *p* | At α = 0.001 |
|----|----|----|----|
| Choa Chu Kang | 0.868 | 0.0293 | fail to reject H₀ |
| Tampines | 0.679 | 2.9e-11 | reject H₀, clustered |

Tampines is clearly clustered. Choa Chu Kang is the more interesting case. Its *R* of 0.868 is below 1, and at the conventional α = 0.05 the result would count as significant, but the threshold stated for this exercise is 0.001 and 0.0293 does not meet it. On the stated criterion the Choa Chu Kang pattern cannot be distinguished from complete spatial randomness.

These two calls also use the default `method = "asymptotic"`, so both are Z-tests and the `nsim = 999` argument is doing nothing, exactly as in the national test above.

The Clark-Evans statistic condenses the whole pattern into one number, so it says *whether* the pattern departs from randomness but not *at what distance* it does so. A pattern can be clustered at 200 m and regular at 2 km, and *R* will report only the net effect. That is precisely the limitation the Choa Chu Kang result runs into. Answering the distance question requires the summary functions G, F, K and L, which are the subject of [Hands-on Exercise 2b](hand-on-ex2b.qmd).

### Computing KDE surfaces by planning area

`bw.diggle` is used throughout so the four surfaces are produced by the same rule, even though the bandwidth it selects differs from area to area:

```{r}
#| fig-width: 12
#| fig-height: 9

par(mfrow = c(2, 2))
plot(density(childcare_pg_ppp.km, sigma = bw.diggle,
             edge = TRUE, kernel = "gaussian"), main = "Punggol")
plot(density(childcare_tm_ppp.km, sigma = bw.diggle,
             edge = TRUE, kernel = "gaussian"), main = "Tampines")
plot(density(childcare_ck_ppp.km, sigma = bw.diggle,
             edge = TRUE, kernel = "gaussian"), main = "Choa Chu Kang")
plot(density(childcare_jw_ppp.km, sigma = bw.diggle,
             edge = TRUE, kernel = "gaussian"), main = "Jurong West")
```

For a fair visual comparison, the same fixed bandwidth is applied to all four. A given colour then means the same intensity in every panel:

```{r}
#| fig-width: 12
#| fig-height: 9

par(mfrow = c(2, 2))
plot(density(childcare_pg_ppp.km, sigma = 0.25,
             edge = TRUE, kernel = "gaussian"), main = "Punggol")
plot(density(childcare_tm_ppp.km, sigma = 0.25,
             edge = TRUE, kernel = "gaussian"), main = "Tampines")
plot(density(childcare_ck_ppp.km, sigma = 0.25,
             edge = TRUE, kernel = "gaussian"), main = "Choa Chu Kang")
plot(density(childcare_jw_ppp.km, sigma = 0.25,
             edge = TRUE, kernel = "gaussian"), main = "Jurong West")
```

## Reference

- Kam, T. S. *R for Geospatial Data Science and Analytics*, [Chapter 4: 1st Order Spatial Point Patterns Analysis Methods](https://r4gdsa.netlify.app/chap04.html).
- Baddeley, A., Rubak, E. and Turner, R. (2015) *Spatial Point Patterns: Methodology and Applications with R*. Chapman and Hall/CRC.
- [spatstat](https://spatstat.org/) project resources.
- Yin, P. (2020). [Kernels and Density Estimation](https://gistbok-ltb.ucgis.org/page/30/concept/9214). *The GIS&T Body of Knowledge*.
