---
title: "Hands-on Exercise 1: Geospatial Data Science with R"
author: "Ji Guofang"
date: "24 January 2023"
date-modified: "last-modified"
format:
  html:
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

In this hands-on exercise, I learn how to handle geospatial data in R using the **sf** package. Geospatial Data Science involves importing, wrangling, integrating, and processing geographically referenced datasets. By the end of this exercise, I am able to:

- install and load **sf** and **tidyverse** packages into R environment,
- import geospatial data using appropriate functions of **sf** package,
- import aspatial data using **readr** package,
- explore simple feature data frames using **Base R** and **sf** functions,
- assign or transform coordinate systems using **sf** functions,
- convert aspatial data into a sf data frame,
- perform geoprocessing tasks using **sf** package,
- perform data wrangling using **dplyr**, and
- perform Exploratory Data Analysis (EDA) using **ggplot2**.

## Data Acquisition

The following datasets are used in this exercise:

- **Master Plan 2014 Subzone Boundary (Web)** from [data.gov.sg](https://data.gov.sg/)
- **Pre-Schools Location** from [data.gov.sg](https://data.gov.sg/)
- **Cycling Path** from [LTADataMall](https://www.mytransport.sg/content/mytransport/home/dataMall.html)
- **Singapore Airbnb listing data** from [Inside Airbnb](http://insideairbnb.com/get-the-data.html)

The geospatial files (`MP14_SUBZONE_WEB_PL`, `CyclingPathGazette`, and `PreSchoolsLocation.kml`) are placed in the `data/geospatial` sub-folder. The aspatial file `listings.csv` is placed in the `data/aspatial` sub-folder.

## Getting Started

Two R packages are used in this exercise:

- **sf** for importing, managing, and processing geospatial data
- **tidyverse** for data science tasks including importing, wrangling, and visualising data

Within tidyverse, the following packages are specifically used: **readr** (importing csv), **tidyr** (manipulating data), **dplyr** (transforming data), and **ggplot2** (visualising data).

```{r}
pacman::p_load(sf, tidyverse)
```

`p_load()` from the **pacman** package is used to install (if not yet installed) and load the packages into R.

## Importing Geospatial Data

### Importing polygon feature data in shapefile format

`st_read()` is used to import `MP14_SUBZONE_WEB_PL` shapefile as a polygon feature data frame. The `dsn` argument specifies the data path and `layer` provides the shapefile name (no file extension needed).

```{r}
mpsz = st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

The GeoJSON file is in **wgs84** (EPSG: 4326) by default. `st_transform(crs = 3414)` is applied immediately upon import to reproject it to the **svy21** projected coordinate system (EPSG: 3414).

### Importing polyline feature data in shapefile format

```{r}
cyclingpath = st_read(dsn = "data/geospatial",
                      layer = "CyclingPathGazette")
```

`cyclingpath` contains 4651 features and 19 fields, represented as multilinestring objects in **svy21** projected coordinates.

### Importing GIS data in GeoJSON format

For GeoJSON files, the full file path including the `.geojson` extension is provided directly to `st_read()`.

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.geojson")
```

`preschool` is a point feature data frame with point features and location fields. It is in **wgs84** coordinate system. The projection transformation to svy21 will be performed later.

## Checking the Content of A Simple Feature Data Frame

### Working with `st_geometry()`

`st_geometry()` retrieves the geometry list-column and prints basic information such as geometry type, geographic extent, and coordinate system.

```{r}
st_geometry(mpsz)
```

### Working with `glimpse()`

`glimpse()` from **dplyr** reveals the data type of each field in the data frame.

```{r}
glimpse(mpsz)
```

From the output, `FMEL-UPD_D` is a **date** field, and `X_ADDR`, `Y_ADDR`, `SHAPE_L`, and `SHAPE_AREA` are **double-precision** numeric fields.

### Working with `head()`

`head()` from Base R displays the first few records of the feature object. The `n` argument controls how many rows to show.

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data

`plot()` provides a quick way to visualise a sf object. By default, it produces a multi-plot of all attributes.

```{r}
plot(mpsz)
```

To plot only the geometry:

```{r}
plot(st_geometry(mpsz))
```

To plot by a specific attribute:

```{r}
plot(mpsz["PLN_AREA_N"])
```

Attempting to overlay the preschool points on the mpsz layer:

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

::: callout-note
The preschool points do not appear on the mpsz layer at this stage because `preschool` is still in **wgs84** while `mpsz` has been transformed to **svy21**. The coordinate systems do not match, so the points fall outside the visible extent. The transformation will be applied in the next section.
:::

## Working with Projection

### Checking the coordinate system of a simple feature data frame

`st_crs()` is used to inspect the coordinate reference system of a sf object. Since `mpsz` was already transformed to svy21 during import, the EPSG code should be 3414:

```{r}
st_crs(mpsz)
```

The output confirms `mpsz` is in **svy21** (EPSG: 3414).

::: callout-note
In some workflows, particularly when importing shapefiles that were incorrectly labelled, the EPSG code may show as 9001 instead of 3414. In that case, `st_set_crs(mpsz, 3414)` can be used to assign the correct label. However, this only changes the label and does **not** mathematically reproject the coordinates. If the data is truly in WGS84 (as with GeoJSON files from data.gov.sg), `st_transform(crs = 3414)` must be used instead.
:::

### Transforming the projection of preschool from wgs84 to svy21

`st_set_crs()` only reassigns the label without mathematically reprojecting the coordinates. To actually reproject data from one CRS to another, `st_transform()` must be used:

```{r}
preschool <- st_transform(preschool, crs = 3414)
```

Verifying the transformation:

```{r}
#| echo: false
st_geometry(preschool)
```

The preschool data is now in svy21. The bounding box values confirm this: the coordinates are well beyond the 0–360 range of decimal degrees used by geographic CRS.

Now overlaying the preschool layer on mpsz works correctly:

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

The composite map now displays as expected.

## Importing and Converting An Aspatial Data

### Importing the aspatial data

`listings.csv` from Inside Airbnb is imported using `read_csv()` from **readr**:

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

Checking the imported data:

```{r}
list(listings)
```

The `listings` tibble has 4252 rows and 16 columns. The `latitude` and `longitude` fields are in decimal degree format, indicating the data is likely in **wgs84** Geographic Coordinate System.

### Creating a simple feature data frame from an aspatial data frame

`st_as_sf()` converts the tibble to a simple feature data frame. The `coords` argument takes the x-coordinate (longitude) first, then y-coordinate (latitude). The data is then reprojected from wgs84 (EPSG: 4326) to svy21 (EPSG: 3414):

```{r}
listings_sf <- st_as_sf(listings,
                        coords = c("longitude", "latitude"),
                        crs = 4326) %>%
  st_transform(crs = 3414)
```

Examining the result:

```{r}
glimpse(listings_sf)
```

A new `geometry` column has been added, while `longitude` and `latitude` columns have been dropped.

Plotting `listings_sf` on top of `mpsz`:

```{r}
#| echo: false
plot(st_geometry(mpsz))
plot(st_geometry(listings_sf), add = TRUE)
```

## Geoprocessing with sf package

### Use case 1: Land acquisition analysis

#### The scenario

The authority plans to upgrade existing cycling paths and needs to acquire 5 metres of reserved land on both sides of each path. The task is to determine the extent and total area of land to be acquired.

#### The solution

First, `st_buffer()` is used to compute 5-metre buffers around cycling paths. The cycling path data is first transformed to svy21:

```{r}
#| echo: false
cyclingpath <- st_transform(cyclingpath, crs = 3414)
```

```{r}
buffer_cycling <- st_buffer(cyclingpath, dist = 5, nQuadSegs = 30)
```

Calculating the area of each buffer using the tidyverse approach:

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

Total land area to be acquired:

```{r}
sum(buffer_cycling$AREA)
```

To visualise the buffer within a specific planning subzone (Tampines West), `filter()` is used to extract the subzone polygon, followed by `st_intersection()` to clip the buffers:

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")
```

```{r}
buffer_cycling_selected <- st_intersection(buffer_cycling, mpsz_selected)
```

```{r}
#| echo: false
plot(st_geometry(buffer_cycling_selected))
```

### Use case 2: Number of pre-schools by planning subzone

#### The scenario

The authority requires a count of pre-schools in each planning subzone to support forward planning.

#### The solution

`st_intersects()` identifies which pre-schools fall within each planning subzone, and `lengths()` counts them:

```{r}
mpsz$`PreSch Count` <- lengths(st_intersects(mpsz, preschool))
```

Summary statistics of the new field:

```{r}
summary(mpsz$`PreSch Count`)
```

The planning subzone with the most pre-schools:

```{r}
top_n(mpsz, 1, `PreSch Count`)
```

Computing the area of each planning subzone:

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

Computing pre-school density (per km²):

```{r}
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count` / Area * 1000000)
```

## Exploratory Data Analysis (EDA)

Plotting the distribution of pre-school density using **ggplot2**:

```{r}
ggplot(data = mpsz,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "There are many planning subzones with a single pre-school, on the other hand,\nthere are seven planning subzones with at least 30 or more pre-schools",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```

Plotting the relationship between pre-school density and pre-school count:

```{r}
ggplot(data = mpsz,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black",
             fill = "light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "Relationship between Pre-school Density and Count",
       x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```

The scatterplot shows a general positive relationship: planning subzones with higher pre-school density also tend to have a higher absolute count of pre-schools.
