---
title: "Hands-on Exercise 1b: Thematic Mapping and GeoVisualisation with R"
author: "Ji Guofang"
date: "December 23, 2022"
date-modified: "last-modified"
format:
  html:
    toc: true
    toc-depth: 5
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

Thematic mapping uses map symbols to visualise selected properties of geographic features that are not naturally visible, such as population, temperature, crime rate, and property prices.

Geovisualisation combines cartography, computer science, and information visualisation to explore, analyse, and communicate geospatial data, enhancing spatial understanding and knowledge discovery.

In this hands-on exercise, I learn how to plot functional and truthful choropleth maps using the **tmap** package in R.

## Getting Started

The following R packages are used in this exercise:

- **tmap** for thematic map creation
- **sf** for handling geospatial data
- **tidyverse**, which includes **readr**, **tidyr** and **dplyr**, for data import and wrangling

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing Data into R

### The Data

Two datasets are used:

- **Master Plan 2019 Subzone Boundary (No Sea) (GeoJSON)** from [data.gov.sg](https://data.gov.sg/), the geospatial data providing Singapore's planning subzone boundaries based on URA Master Plan 2019.
- **Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2024** (`respopagesextod2024.csv`) from [Department of Statistics, Singapore](https://www.singstat.gov.sg/), the aspatial data whose PA and SZ fields georeference it to the subzone boundaries.

### Importing Geospatial Data into R

`st_read()` is used to import the GeoJSON file, and the data is immediately transformed to the SVY21 projected coordinate system (EPSG: 3414). Unlike KML, the GeoJSON format already encodes attribute fields as proper columns, so no additional parsing is needed.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

Examining the `mpsz` data frame:

```{r}
mpsz
```

Only the first ten records are displayed by default because `mpsz` is a simple feature (`sf`) data frame, and sf objects print a limited number of rows to keep output manageable.

### Importing Attribute Data into R

The population CSV file is imported using `read_csv()`:

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

### Data Preparation

A data table with 2024 values is prepared with the variables: `PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, and `DEPENDENCY`.

- **YOUNG**: age groups 0–4 to 20–24
- **ECONOMY ACTIVE**: age groups 25–29 to 60–64
- **AGED**: age group 65 and above
- **TOTAL**: all age groups
- **DEPENDENCY**: ratio of (YOUNG + AGED) to ECONOMY ACTIVE

#### Data wrangling

`pivot_wider()` from **tidyr** and `mutate()`, `filter()`, `group_by()`, `select()` from **dplyr** are used. Column names are referenced explicitly in `rowSums()` instead of positional indices, as `pivot_wider()` sorts age group columns alphabetically (causing `5_to_9` to appear after `55_to_59`), which makes positional indexing unreliable:

```{r}
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG,
              values_from = POP) %>%
  mutate(YOUNG = rowSums(select(., `0_to_4`, `5_to_9`, `10_to_14`,
                                   `15_to_19`, `20_to_24`),
                         na.rm = TRUE)) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(select(., `25_to_29`, `30_to_34`,
                                              `35_to_39`, `40_to_44`,
                                              `45_to_49`, `50_to_54`,
                                              `55_to_59`, `60_to_64`),
                                    na.rm = TRUE)) %>%
  mutate(`AGED` = rowSums(select(., `65_to_69`, `70_to_74`, `75_to_79`,
                                    `80_to_84`, `85_to_89`, `90_and_over`),
                          na.rm = TRUE)) %>%
  mutate(`TOTAL` = rowSums(select(., `0_to_4`:`90_and_over`),
                           na.rm = TRUE)) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`) / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`)
```

#### Joining attribute data and geospatial data

The `PA` and `SZ` values are converted to uppercase to match `SUBZONE_N` and `PLN_AREA_N`, and rows with no economy active population are removed:

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ),
            .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

`left_join()` joins the geospatial and attribute data using `SUBZONE_N` and `SZ` as the common identifier. Using `mpsz` as the left table ensures the output remains a simple feature data frame:

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                           by = c("SUBZONE_N" = "SZ"))
```

Saving the joined data frame as an RDS file for future use:

```{r}
dir.create("data/rds/", recursive = TRUE, showWarnings = FALSE)
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

## Choropleth Mapping with *tmap*

Two approaches are available for creating choropleth maps in **tmap**:

1.  Quick plot using `qtm()`
2.  Fully customisable map using tmap layer elements

### Quick choropleth map with `qtm()`

`qtm()` provides a concise way to produce a standard choropleth map. `tmap_mode("plot")` sets the output to a static map:

```{r}
tmap_mode("plot")
qtm(shp = mpsz_pop2024,
    fill = "DEPENDENCY")
```

### Customisable choropleth map using tmap elements

For finer control over cartographic quality, tmap's individual drawing elements are used:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

#### Drawing a base map

`tm_shape()` defines the input data, and `tm_polygons()` draws the polygon layer with default styling (light gray fill, dark gray borders):

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons()
```

#### Choropleth map using `tm_polygons()`

Assigning a variable to `fill` within `tm_polygons()` produces a choropleth map. The default classification method is *pretty* and the default colour scheme is `blues3`:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY")
```

#### Using `tm_fill()` and `tm_borders()`

`tm_polygons()` is a wrapper of `tm_fill()` and `tm_borders()`. Using them separately provides more control. `tm_fill()` alone (no borders):

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY")
```

Adding borders with `tm_borders()`:

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders()
```

Customising border appearance with `col`, `lwd`, and `lty`:

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill("DEPENDENCY") +
  tm_borders(col = "grey60",
             lwd = 0.1,
             lty = "dashed")
```

### Data Classification Methods

**tmap** supports ten classification methods: *fixed*, *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher*, and *jenks*. The `style` argument in `tm_scale_intervals()` is used to specify the method.

#### Quantile classification (5 classes)

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

#### Equal interval classification (5 classes)

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The quantile method produces more evenly distributed class sizes compared to equal interval, which can be skewed by data distribution.

#### Custom break points

Descriptive statistics are checked first to inform break point choices:

```{r}
summary(mpsz_pop2024$DEPENDENCY)
```

Custom breaks are set at 0.60, 0.70, 0.80, and 0.90, with minimum 0 and maximum 1.00:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00))) +
  tm_borders(fill_alpha = 0.5)
```

### Colour Scheme

#### Using ColourBrewer palette

The `values` argument in `tm_scale_intervals()` accepts ColourBrewer palette names:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour direction, prefix the palette name with `-`:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

#### Cartographic furniture

`tm_compass()`, `tm_scalebar()`, `tm_grid()`, and `tm_credits()` add map furniture:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5)) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

### Map Layout

#### Map legend

`fill.legend` and `tm_legend()` customise the legend position and title:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_pos_auto_in() +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_scalebar() +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: data.gov.sg & singstat",
             position = c("left", "bottom"))
```

#### Map style

`tmap_style()` applies predefined layout themes. The *natural* style is used below:

```{r}
tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

Resetting to the default white style:

```{r}
tmap_style("white")
```

## Small Multiple Choropleth Maps

Small multiple maps (facet maps) place many maps side-by-side to show how spatial patterns change with respect to another variable. In **tmap**, they can be created three ways.

### By assigning multiple values to an aesthetic argument

Assigning two variables to `fill` generates side-by-side choropleth maps:

```{r}
#| fig-width: 12

tm_shape(mpsz_pop2024) +
  tm_polygons(
    fill = c("YOUNG", "AGED"),
    fill.legend = tm_legend(
      position = tm_pos_in("right", "bottom")),
    fill.scale = tm_scale_intervals(
      style = "equal",
      n = 5,
      values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tmap_style("natural")
```

### By arranging maps in a grid layout with `tmap_arrange()`

Individual maps are created separately and then arranged:

```{r}
youngmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "YOUNG",
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of young population")

agedmap <- tm_shape(mpsz_pop2024) +
  tm_polygons(fill = "AGED",
              fill.legend = tm_legend(
                position = tm_pos_in("right", "bottom"),
                item.height = 0.8),
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of aged population")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

### By defining a group-by variable in `tm_facets()`

`tm_facets()` splits the map by a categorical variable:

```{r}
tm_shape(mpsz_pop2024) +
  tm_fill(fill = "DEPENDENCY",
          fill.scale = tm_scale_intervals(
            style = "quantile",
            values = "brewer.blues")) +
  tm_facets(by = "REGION_N",
            nrow = 2,
            ncol = 3,
            free.coords = TRUE,
            drop.units = TRUE) +
  tm_layout(legend.show = TRUE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(fill_alpha = 0.5)
```

## Mapping Spatial Object Meeting a Selection Criterion

`filter()` from **dplyr** is used to focus the choropleth map on a specific region:

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.greens"),
              fill.legend = tm_legend()) +
  tm_borders(fill_alpha = 0.5)
```

## Complementing Thematic Map with Statistical Chart

Maps and statistical charts complement each other: maps show spatial patterns while charts display numerical distributions. **tmap** supports embedding statistical charts directly into map visualisations using the `fill.chart` argument.

Adding a box plot to the Central Region dependency map:

```{r}
mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION") %>%
  tm_shape() +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.greens"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders() +
  tm_layout(asp = 0.8)
```

Outliers in the dependency distribution are highlighted on the choropleth map using red borders and subzone labels:

```{r}
mpsz_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")

stats <- boxplot.stats(mpsz_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- mpsz_selected[mpsz_selected$DEPENDENCY %in% outlier_vals, ]

tm_shape(mpsz_selected) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                values = "brewer.blues"),
              fill.legend = tm_legend(),
              fill.chart = tm_chart_box()) +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_text("SUBZONE_N", col = "red", size = 0.7) +
  tm_layout(asp = 0.8)
```

## Creating Interactive Maps

`tmap_mode("view")` switches to interactive mode, allowing users to zoom, pan, and click on features for popup information.

Basic interactive map of the Central Region:

```{r}
region_selected <- mpsz_pop2024 %>%
  filter(REGION_N == "CENTRAL REGION")
region_bbox <- st_bbox(region_selected)

stats <- boxplot.stats(region_selected$DEPENDENCY)
outlier_vals <- stats$out
outlier_sf <- region_selected[region_selected$DEPENDENCY %in% outlier_vals, ]

tmap_mode("view")
tm_shape(region_selected,
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name"       = "SUBZONE_N",
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2)
tmap_mode("plot")
```

To prevent users from zooming too far out of the study area, `set_zoom_limits` restricts the zoom range:

```{r}
tmap_mode("view")
tm_shape(region_selected,
         bbox = region_bbox) +
  tm_fill("DEPENDENCY",
          id = "SUBZONE_N",
          popup.vars = c(
            "Name"       = "SUBZONE_N",
            "Dependency" = "DEPENDENCY")) +
  tm_borders() +
  tm_shape(outlier_sf) +
  tm_borders(col = "red", lwd = 2) +
  tm_view(set_zoom_limits = c(12, 14))
tmap_mode("plot")
```

Setting `set_zoom_limits = c(12, 14)` restricts how far users can zoom in (level 14) and out (level 12), keeping the focus on the Central Region throughout exploration.

## Reference

### tmap package

- [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)
- [tmap on CRAN](https://cran.r-project.org/web/packages/tmap/index.html)
- [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

### Geospatial data wrangling

- [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)
- [Simple Features for R: Standardized Support for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)

### Data wrangling

- [dplyr](https://dplyr.tidyverse.org/)
- [tidyr: Tidy Data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
