---
title: "Hands-on Exercise 1: Geospatial Data Science with R"
author: "Ji Guofang"
date: "24 January 2023"
date-modified: "last-modified"
format:
  html:
    toc-depth: 4
    toc-expand: true
execute:
  echo: true
  eval: true
  warning: false
  freeze: auto
editor: visual
---

## Overview

::: {.en}
In this hands-on exercise, I learn how to handle geospatial data in R using the **sf** package. Geospatial Data Science involves importing, wrangling, integrating, and processing geographically referenced datasets. By the end of this exercise, I am able to:

- install and load **sf** and **tidyverse** packages into R environment,
- import geospatial data using appropriate functions of **sf** package,
- import aspatial data using **readr** package,
- explore simple feature data frames using **Base R** and **sf** functions,
- assign or transform coordinate systems using **sf** functions,
- convert aspatial data into a sf data frame,
- perform geoprocessing tasks using **sf** package,
- perform data wrangling using **dplyr**, and
- perform Exploratory Data Analysis (EDA) using **ggplot2**.
:::

::: {.zh}
本次练习学习用 **sf** 包在 R 中处理地理空间数据。地理空间数据科学所做的事，是把带有地理参照的数据集导入、整理、整合并加工。练习结束时应当能够：

- 在 R 环境中安装并载入 **sf** 与 **tidyverse**；
- 用 **sf** 的相应函数导入地理空间数据；
- 用 **readr** 导入非空间数据；
- 用 Base R 与 **sf** 的函数查看简单要素数据框；
- 用 **sf** 的函数指定或转换坐标系统；
- 把非空间数据转换成 sf 数据框；
- 用 **sf** 完成地理处理任务；
- 用 **dplyr** 做数据整理；
- 用 **ggplot2** 做探索性数据分析。
:::

## Data Acquisition

::: {.en}
The following datasets are used in this exercise:

- **Master Plan 2014 Subzone Boundary (Web)** from [data.gov.sg](https://data.gov.sg/)
- **Pre-Schools Location** from [data.gov.sg](https://data.gov.sg/)
- **Cycling Path** from [LTADataMall](https://www.mytransport.sg/content/mytransport/home/dataMall.html)
- **Singapore Airbnb listing data** from [Inside Airbnb](http://insideairbnb.com/get-the-data.html)

The geospatial files (`MP14_SUBZONE_WEB_PL`, `CyclingPathGazette`, and `PreSchoolsLocation.kml`) are placed in the `data/geospatial` sub-folder. The aspatial file `listings.csv` is placed in the `data/aspatial` sub-folder.
:::

::: {.zh}
本练习使用以下数据集：

- **Master Plan 2014 Subzone Boundary (Web)**，来自 [data.gov.sg](https://data.gov.sg/)
- **Pre-Schools Location**，来自 [data.gov.sg](https://data.gov.sg/)
- **Cycling Path**，来自 [LTADataMall](https://www.mytransport.sg/content/mytransport/home/dataMall.html)
- **Singapore Airbnb listing data**，来自 [Inside Airbnb](http://insideairbnb.com/get-the-data.html)

三个地理空间文件（`MP14_SUBZONE_WEB_PL`、`CyclingPathGazette`、`PreSchoolsLocation.kml`）放在 `data/geospatial` 子目录下，非空间文件 `listings.csv` 放在 `data/aspatial` 子目录下。
:::

## Getting Started

::: {.en}
Two R packages are used in this exercise:

- **sf** for importing, managing, and processing geospatial data
- **tidyverse** for data science tasks including importing, wrangling, and visualising data

Within tidyverse, the following packages are specifically used: **readr** (importing csv), **tidyr** (manipulating data), **dplyr** (transforming data), and **ggplot2** (visualising data).
:::

::: {.zh}
本练习用到两个 R 包：

- **sf**，负责地理空间数据的导入、管理与加工；
- **tidyverse**，负责导入、整理、可视化等数据科学任务。

tidyverse 当中具体用到的是：**readr**（读 csv）、**tidyr**（整形）、**dplyr**（变换）、**ggplot2**（绘图）。
:::

```{r}
pacman::p_load(sf, tidyverse)
```

::: {.en}
`p_load()` from the **pacman** package is used to install (if not yet installed) and load the packages into R.
:::

::: {.zh}
**pacman** 包的 `p_load()` 会检查这些包是否已安装，未安装则先安装，然后一并载入 R。
:::

## Importing Geospatial Data

### Importing polygon feature data in shapefile format

::: {.en}
`st_read()` is used to import `MP14_SUBZONE_WEB_PL` shapefile as a polygon feature data frame. The `dsn` argument specifies the data path and `layer` provides the shapefile name (no file extension needed).
:::

::: {.zh}
`st_read()` 把 `MP14_SUBZONE_WEB_PL` 这个 shapefile 读成多边形要素数据框。`dsn` 指定数据所在路径，`layer` 给出图层名，不需要带扩展名。
:::

```{r}
mpsz = st_read("data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson") %>%
  st_transform(crs = 3414)
```

::: {.en}
The GeoJSON file is in **wgs84** (EPSG: 4326) by default. `st_transform(crs = 3414)` is applied immediately upon import to reproject it to the **svy21** projected coordinate system (EPSG: 3414).
:::

::: {.zh}
这个 GeoJSON 文件默认是 **wgs84**（EPSG: 4326）。导入时紧接着调用 `st_transform(crs = 3414)`，把它投影到新加坡的 **svy21** 投影坐标系统（EPSG: 3414）。
:::

### Importing polyline feature data in shapefile format

```{r}
cyclingpath = st_read(dsn = "data/geospatial",
                      layer = "CyclingPathGazette")
```

::: {.en}
`cyclingpath` contains 4651 features and 19 fields, represented as multilinestring objects in **svy21** projected coordinates.
:::

::: {.zh}
`cyclingpath` 含 4651 个要素、19 个字段，几何类型是 multilinestring，坐标已在 **svy21** 投影下。
:::

### Importing GIS data in GeoJSON format

::: {.en}
For GeoJSON files, the full file path including the `.geojson` extension is provided directly to `st_read()`.
:::

::: {.zh}
读 GeoJSON 时，直接把含 `.geojson` 扩展名的完整路径交给 `st_read()` 即可。
:::

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.geojson")
```

::: {.en}
`preschool` is a point feature data frame with point features and location fields. It is in **wgs84** coordinate system. The projection transformation to svy21 will be performed later.
:::

::: {.zh}
`preschool` 是点要素数据框，带有点几何和位置字段，坐标系统是 **wgs84**。投影到 svy21 的转换留到后面再做。
:::

## Checking the Content of A Simple Feature Data Frame

### Working with `st_geometry()`

::: {.en}
`st_geometry()` retrieves the geometry list-column and prints basic information such as geometry type, geographic extent, and coordinate system.
:::

::: {.zh}
`st_geometry()` 取出几何列表列，并打印几何类型、地理范围、坐标系统这些基本信息。
:::

```{r}
st_geometry(mpsz)
```

### Working with `glimpse()`

::: {.en}
`glimpse()` from **dplyr** reveals the data type of each field in the data frame.
:::

::: {.zh}
**dplyr** 的 `glimpse()` 列出数据框每个字段的数据类型。
:::

```{r}
glimpse(mpsz)
```

::: {.en}
From the output, `FMEL-UPD_D` is a **date** field, and `X_ADDR`, `Y_ADDR`, `SHAPE_L`, and `SHAPE_AREA` are **double-precision** numeric fields.
:::

::: {.zh}
从输出可见，`FMEL-UPD_D` 是日期字段，`X_ADDR`、`Y_ADDR`、`SHAPE_L`、`SHAPE_AREA` 是双精度数值字段。
:::

### Working with `head()`

::: {.en}
`head()` from Base R displays the first few records of the feature object. The `n` argument controls how many rows to show.
:::

::: {.zh}
Base R 的 `head()` 显示要素对象的前若干条记录，显示多少行由 `n` 控制。
:::

```{r}
head(mpsz, n=5)
```

## Plotting the Geospatial Data

::: {.en}
`plot()` provides a quick way to visualise a sf object. By default, it produces a multi-plot of all attributes.
:::

::: {.zh}
`plot()` 是快速查看 sf 对象的办法，默认会把所有属性各画一张，拼成多图。
:::

```{r}
plot(mpsz)
```

::: {.en}
To plot only the geometry:
:::

::: {.zh}
只画几何：
:::

```{r}
plot(st_geometry(mpsz))
```

::: {.en}
To plot by a specific attribute:
:::

::: {.zh}
按某一个属性上色：
:::

```{r}
plot(mpsz["PLN_AREA_N"])
```

::: {.en}
Attempting to overlay the preschool points on the mpsz layer:
:::

::: {.zh}
尝试把幼儿园点叠到 mpsz 图层上：
:::

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

::: callout-note
The preschool points do not appear on the mpsz layer at this stage because `preschool` is still in **wgs84** while `mpsz` has been transformed to **svy21**. The coordinate systems do not match, so the points fall outside the visible extent. The transformation will be applied in the next section.
:::

## Working with Projection

### Checking the coordinate system of a simple feature data frame

::: {.en}
`st_crs()` is used to inspect the coordinate reference system of a sf object. Since `mpsz` was already transformed to svy21 during import, the EPSG code should be 3414:
:::

::: {.zh}
`st_crs()` 用来查看 sf 对象的坐标参考系统。`mpsz` 在导入时已经转到 svy21，所以 EPSG 代码应当是 3414：
:::

```{r}
st_crs(mpsz)
```

::: {.en}
The output confirms `mpsz` is in **svy21** (EPSG: 3414).
:::

::: {.zh}
输出确认 `mpsz` 处于 **svy21**（EPSG: 3414）。
:::

::: callout-note
In some workflows, particularly when importing shapefiles that were incorrectly labelled, the EPSG code may show as 9001 instead of 3414. In that case, `st_set_crs(mpsz, 3414)` can be used to assign the correct label. However, this only changes the label and does **not** mathematically reproject the coordinates. If the data is truly in WGS84 (as with GeoJSON files from data.gov.sg), `st_transform(crs = 3414)` must be used instead.
:::

### Transforming the projection of preschool from wgs84 to svy21

::: {.en}
`st_set_crs()` only reassigns the label without mathematically reprojecting the coordinates. To actually reproject data from one CRS to another, `st_transform()` must be used:
:::

::: {.zh}
`st_set_crs()` 只是改写标签，并不真正换算坐标。要把数据从一个坐标系真正投影到另一个，必须用 `st_transform()`：
:::

```{r}
preschool <- st_transform(preschool, crs = 3414)
```

::: {.en}
Verifying the transformation:
:::

::: {.zh}
核对转换结果：
:::

```{r}
#| echo: false
st_geometry(preschool)
```

::: {.en}
The preschool data is now in svy21. The bounding box values confirm this: the coordinates are well beyond the 0–360 range of decimal degrees used by geographic CRS.
:::

::: {.zh}
幼儿园数据现在处于 svy21。边界框的数值可以印证：坐标远远超出地理坐标系十进制度数的 0–360 范围。
:::

::: {.en}
Now overlaying the preschool layer on mpsz works correctly:
:::

::: {.zh}
此时再把幼儿园图层叠到 mpsz 上就正常了：
:::

```{r}
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

::: {.en}
The composite map now displays as expected.
:::

::: {.zh}
叠加后的地图显示符合预期。
:::

## Importing and Converting An Aspatial Data

### Importing the aspatial data

::: {.en}
`listings.csv` from Inside Airbnb is imported using `read_csv()` from **readr**:
:::

::: {.zh}
用 **readr** 的 `read_csv()` 导入 Inside Airbnb 的 `listings.csv`：
:::

```{r}
listings <- read_csv("data/aspatial/listings.csv")
```

::: {.en}
Checking the imported data:
:::

::: {.zh}
查看导入结果：
:::

```{r}
list(listings)
```

::: {.en}
The `listings` tibble has 4252 rows and 16 columns. The `latitude` and `longitude` fields are in decimal degree format, indicating the data is likely in **wgs84** Geographic Coordinate System.
:::

::: {.zh}
`listings` 这个 tibble 有 4252 行、16 列。`latitude` 和 `longitude` 两个字段是十进制度数，说明数据大概处于 **wgs84** 地理坐标系统。
:::

### Creating a simple feature data frame from an aspatial data frame

::: {.en}
`st_as_sf()` converts the tibble to a simple feature data frame. The `coords` argument takes the x-coordinate (longitude) first, then y-coordinate (latitude). The data is then reprojected from wgs84 (EPSG: 4326) to svy21 (EPSG: 3414):
:::

::: {.zh}
`st_as_sf()` 把 tibble 转成简单要素数据框。`coords` 的顺序是先 x（经度）后 y（纬度）。之后再从 wgs84（EPSG: 4326）投影到 svy21（EPSG: 3414）：
:::

```{r}
listings_sf <- st_as_sf(listings,
                        coords = c("longitude", "latitude"),
                        crs = 4326) %>%
  st_transform(crs = 3414)
```

::: {.en}
Examining the result:
:::

::: {.zh}
查看转换结果：
:::

```{r}
glimpse(listings_sf)
```

::: {.en}
A new `geometry` column has been added, while `longitude` and `latitude` columns have been dropped.
:::

::: {.zh}
新增了 `geometry` 列，原来的 `longitude` 与 `latitude` 两列被移除。
:::

::: {.en}
Plotting `listings_sf` on top of `mpsz`:
:::

::: {.zh}
把 `listings_sf` 叠到 `mpsz` 上：
:::

```{r}
#| echo: false
plot(st_geometry(mpsz))
plot(st_geometry(listings_sf), add = TRUE)
```

## Geoprocessing with sf package

### Use case 1: Land acquisition analysis

#### The scenario

::: {.en}
The authority plans to upgrade existing cycling paths and needs to acquire 5 metres of reserved land on both sides of each path. The task is to determine the extent and total area of land to be acquired.
:::

::: {.zh}
管理部门计划升级现有自行车道，需要在每条车道两侧各征收 5 米的保留地。任务是算出征地的范围和总面积。
:::

#### The solution

::: {.en}
First, `st_buffer()` is used to compute 5-metre buffers around cycling paths. The cycling path data is first transformed to svy21:
:::

::: {.zh}
先用 `st_buffer()` 为自行车道生成 5 米缓冲区。自行车道数据先转到 svy21：
:::

```{r}
#| echo: false
cyclingpath <- st_transform(cyclingpath, crs = 3414)
```

```{r}
buffer_cycling <- st_buffer(cyclingpath, dist = 5, nQuadSegs = 30)
```

::: {.en}
Calculating the area of each buffer using the tidyverse approach:
:::

::: {.zh}
用 tidyverse 的写法算出每个缓冲区的面积：
:::

```{r}
buffer_cycling <- buffer_cycling %>%
  mutate(AREA = st_area(geometry))
```

::: {.en}
Total land area to be acquired:
:::

::: {.zh}
需要征收的土地总面积：
:::

```{r}
sum(buffer_cycling$AREA)
```

::: {.en}
To visualise the buffer within a specific planning subzone (Tampines West), `filter()` is used to extract the subzone polygon, followed by `st_intersection()` to clip the buffers:
:::

::: {.zh}
若要看某一个规划分区（Tampines West）内的缓冲区，先用 `filter()` 取出该分区多边形，再用 `st_intersection()` 裁切缓冲区：
:::

```{r}
mpsz_selected <- mpsz %>%
  filter(SUBZONE_N == "TAMPINES WEST")
```

```{r}
buffer_cycling_selected <- st_intersection(buffer_cycling, mpsz_selected)
```

```{r}
#| echo: false
plot(st_geometry(buffer_cycling_selected))
```

### Use case 2: Number of pre-schools by planning subzone

#### The scenario

::: {.en}
The authority requires a count of pre-schools in each planning subzone to support forward planning.
:::

::: {.zh}
管理部门需要知道每个规划分区内幼儿园的数量，以支持后续规划。
:::

#### The solution

::: {.en}
`st_intersects()` identifies which pre-schools fall within each planning subzone, and `lengths()` counts them:
:::

::: {.zh}
`st_intersects()` 判断哪些幼儿园落在各个规划分区内，`lengths()` 把个数数出来：
:::

```{r}
mpsz$`PreSch Count` <- lengths(st_intersects(mpsz, preschool))
```

::: {.en}
Summary statistics of the new field:
:::

::: {.zh}
新字段的描述统计：
:::

```{r}
summary(mpsz$`PreSch Count`)
```

::: {.en}
The planning subzone with the most pre-schools:
:::

::: {.zh}
幼儿园最多的规划分区：
:::

```{r}
top_n(mpsz, 1, `PreSch Count`)
```

::: {.en}
Computing the area of each planning subzone:
:::

::: {.zh}
计算每个规划分区的面积：
:::

```{r}
mpsz$Area <- mpsz %>%
  st_area()
```

::: {.en}
Computing pre-school density (per km²):
:::

::: {.zh}
计算幼儿园密度（每平方公里）：
:::

```{r}
mpsz <- mpsz %>%
  mutate(`PreSch Density` = `PreSch Count` / Area * 1000000)
```

## Exploratory Data Analysis (EDA)

::: {.en}
Plotting the distribution of pre-school density using **ggplot2**:
:::

::: {.zh}
用 **ggplot2** 画出幼儿园密度的分布：
:::

```{r}
ggplot(data = mpsz,
       aes(x = as.numeric(`PreSch Density`))) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  labs(title = "Are pre-schools evenly distributed in Singapore?",
       subtitle = "There are many planning subzones with a single pre-school, on the other hand,\nthere are seven planning subzones with at least 30 or more pre-schools",
       x = "Pre-school density (per km sq)",
       y = "Frequency")
```

::: {.en}
Plotting the relationship between pre-school density and pre-school count:
:::

::: {.zh}
画出幼儿园密度与幼儿园数量之间的关系：
:::

```{r}
ggplot(data = mpsz,
       aes(y = `PreSch Count`,
           x = as.numeric(`PreSch Density`))) +
  geom_point(color = "black",
             fill = "light blue") +
  xlim(0, 40) +
  ylim(0, 40) +
  labs(title = "Relationship between Pre-school Density and Count",
       x = "Pre-school density (per km sq)",
       y = "Pre-school count")
```

::: {.en}
The scatterplot shows a general positive relationship: planning subzones with higher pre-school density also tend to have a higher absolute count of pre-schools.
:::

::: {.zh}
散点图显示两者大体正相关：幼儿园密度较高的规划分区，其幼儿园绝对数量也往往较多。
:::
